---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 2: Core cards Operations"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Learn to generate continuous, categorical, dichotomous, and missingness summaries using the core functions of the cards R package.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
```

---

## Chapter 2: Core cards Operations

### 2.1 Continuous, Categorical, and Dichotomous Functions

The `{cards}` package provides a set of specialized functions to calculate statistics based on the data type:

- **`ard_continuous()`**: Calculates continuous summaries (such as `mean`, `sd`, `median`, `min`, `max`, and quantiles).
- **`ard_categorical()`**: Computes frequency tables, outputting counts (`n`), denominators (`N`), and percentages (`p`) for every level of a categorical factor.
- **`ard_dichotomous()`**: Extracts categorical summaries but retains only a single positive outcome level (e.g. keeping only response `"Y"`).
- **`ard_missing()`**: Summarizes counts and percentages of missing (`NA`) vs non-missing values.
- **`ard_total_n()`**: Calculates the total subject count (denominator) per grouping.

All of these functions share the `by` argument to group calculations by variables such as `TRT01A`.

---

## Exercise 2.1 — Summarize Continuous Data

Summarize the continuous variable `AGE` in the `adsl` dataset, grouped by treatment arm (`TRT01A`).

```{r ch2-ex1, exercise=TRUE, exercise.lines=10}
# Generate the continuous ARD for AGE grouped by TRT01A
# Hint: use ard_continuous() with variables = AGE and by = TRT01A

ard_continuous(
  data = adsl,
  variables = ___,
  by = ___
)
```

```{r ch2-ex1-hint-1}
# Pass AGE to variables and TRT01A to by
ard_continuous(adsl, variables = AGE, by = TRT01A)
```

```{r ch2-ex1-solution}
ard_continuous(adsl, variables = AGE, by = TRT01A)
```

```{r ch2-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "mean" %in% .result$stat_name, "Correct! ard_continuous successfully generated the continuous statistics (mean, sd, etc.) for AGE by TRT01A."),
  fail_if(~ TRUE, "Ensure you are using ard_continuous with variables = AGE and by = TRT01A.")
)
```

---

## Exercise 2.2 — Summarize Categorical Data

Summarize the categorical variable `SEX` in `adsl`, grouped by treatment arm (`TRT01A`).

```{r ch2-ex2, exercise=TRUE, exercise.lines=10}
# Calculate categorical counts and percentages for SEX by TRT01A
# Hint: use ard_categorical()

ard_categorical(
  data = adsl,
  variables = ___,
  by = ___
)
```

```{r ch2-ex2-hint-1}
# Pass SEX to variables and TRT01A to by
ard_categorical(adsl, variables = SEX, by = TRT01A)
```

```{r ch2-ex2-solution}
ard_categorical(adsl, variables = SEX, by = TRT01A)
```

```{r ch2-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("F", "M") %in% .result$variable_grp_level), "Excellent! You generated categorical summaries. Notice that there are rows for both 'F' and 'M' levels."),
  fail_if(~ TRUE, "Verify that you are using ard_categorical() with variables = SEX and by = TRT01A.")
)
```

---

## Exercise 2.3 — Summarize Dichotomous Flag Variables

Clinical data contains many flag variables (like `SAFFL = "Y"`). We often only want to show the positive indicator in our table. Use `ard_dichotomous()` to summarize the `SAFFL` (Safety Population Flag) variable for the `"Y"` level, grouped by `TRT01A`.

```{r ch2-ex3, exercise=TRUE, exercise.lines=15}
# Summarize SAFFL where value is "Y"
# Hint: use ard_dichotomous() and specify value = list(SAFFL = "Y")

ard_dichotomous(
  data = adsl,
  by = TRT01A,
  variables = SAFFL,
  value = list(SAFFL = ___)
)
```

```{r ch2-ex3-hint-1}
# Supply "Y" to the SAFFL item in the value list
ard_dichotomous(adsl, variables = SAFFL, value = list(SAFFL = "Y"), by = TRT01A)
```

```{r ch2-ex3-solution}
ard_dichotomous(adsl, variables = SAFFL, value = list(SAFFL = "Y"), by = TRT01A)
```

```{r ch2-ex3-check}
grade_result(
  pass_if(~ is.data.frame(.result) && !("N" %in% .result$variable_grp_level), "Superb! ard_dichotomous kept only the rows for the 'Y' level, saving space and simplifying downstream table generation."),
  fail_if(~ TRUE, "Make sure you set value = list(SAFFL = 'Y') inside ard_dichotomous.")
)
```

---

## Chapter 2 Knowledge Check

```{r ch2-quiz, echo=FALSE}
quiz(
  caption = "Chapter 2 — Check Your Understanding",

  question(
    "Which function should you use to summarize continuous variables like weight or age?",
    answer("ard_categorical()"),
    answer("ard_continuous()", correct = TRUE),
    answer("ard_total_n()"),
    answer("ard_missing()"),
    allow_retry = TRUE
  ),

  question(
    "How does `ard_dichotomous()` differ from `ard_categorical()`?",
    answer("It operates on continuous data"),
    answer("It keeps only the statistic rows corresponding to the specified level (e.g. 'Y')", correct = TRUE),
    answer("It does not calculate percentages"),
    answer("It runs twice as fast"),
    allow_retry = TRUE
  ),

  question(
    "What argument do you use to group calculations by treatment arm in `{cards}` functions?",
    answer("group"),
    answer("split"),
    answer("by", correct = TRUE),
    answer("strata"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 2 Summary {.congratulations}

You now know how to:
- Generate summaries for continuous variables using **`ard_continuous()`**.
- Compute categorical frequency tables using **`ard_categorical()`**.
- Extract positive response flags using **`ard_dichotomous()`**.
- Perform grouped analysis across all functions via the uniform **`by`** argument.

**Next:** Chapter 3 — Stacking and Formatting (`bind_ard`, `ard_fmt_args`, etc.)
