---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 10: Broom vs. Parameters Tidiers"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Understand how R translates complex model objects into clean datasets using model tidiers, and how cardx standardizes these outputs.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(dplyr)
library(cards)
library(cardx)
library(broom)
library(broom.helpers)
library(parameters)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")

# Pre-fit a logistic regression model for the exercises
# Predicting Safety Population inclusion (SAFFL == "Y") based on Treatment Arm and Age
adsl_model_data <- adsl %>%
  filter(!is.na(TRT01A), !is.na(AGE)) %>%
  mutate(RESP_BIN = ifelse(SAFFL == "Y", 1, 0))

model_fit <- glm(RESP_BIN ~ TRT01A + AGE, data = adsl_model_data, family = binomial)
```

---

## 10.1 Introduction to Tidiers vs. SAS ODS

In SAS, when you run a statistical procedure like `PROC LOGISTIC`, the math happens in the background, and the results are automatically printed to the screen. To save those results into a dataset, you use the Output Delivery System (ODS):

```sas
ods output ParameterEstimates = my_estimates;
proc logistic data=adsl;
  class TRT01A / param=ref;
  model RESP_BIN(event='1') = TRT01A AGE;
run;
```

In R, things work differently. A modeling function like `glm()` returns a **complex model object**. This object is essentially a massive, nested list containing the coefficients, residuals, gradients, and model frame. It is *not* a flat data frame.

To get the equivalent of `ParameterEstimates`, R uses **"tidier" functions**. A tidier takes the messy, nested model object and extracts the statistics into a clean, flat data frame (a tibble). 

The two most prominent packages for this in R are `{broom}` and `{parameters}`.

---

## 10.2 The Column Mapping Problem

Just as different SAS PROCs might use slightly different column names in their ODS outputs (e.g., `Estimate` vs `Coefficient`, or `ProbChiSq` vs `Probt`), `{broom}` and `{parameters}` use different naming conventions.

When standardizing output for the ARD structure, `{broom.helpers}` and `{cardx}` strictly adopt the `{broom}` convention. Here is the translation dictionary:

| Concept | `{broom}::tidy()` Column | `{parameters}::model_parameters()` Column | Standardized ARD Output |
|:---|:---|:---|:---|
| Coefficient Name | `term` | `Parameter` | `term` |
| Point Estimate | `estimate` | `Coefficient` | `estimate` |
| Standard Error | `std.error` | `SE` | `std.error` |
| Test Statistic | `statistic` | `z` | `statistic` |
| p-value | `p.value` | `p` | `p.value` |
| Conf. Int. (Low)| `conf.low` | `CI_low` | `conf.low` |
| Conf. Int. (High)| `conf.high` | `CI_high` | `conf.high` |

---

## 10.3 Key Differences: broom vs parameters

While both packages do the exact same job (extracting coefficients), they have wildly different philosophies on how to present the data, particularly regarding categorical variables.

### The Problem with `{broom}`
`{broom}` is minimalist. If a coefficient isn't mathematically calculated by the model engine, `{broom}` will not output a row for it.
For a categorical variable with three levels (Placebo, Low Dose, High Dose), the model only estimates coefficients for Low Dose and High Dose. The Placebo group is the **reference level** (its effect is absorbed into the Intercept, so mathematically its coefficient is 0, or Odds Ratio = 1).
**Therefore, `{broom}` completely drops the Placebo row.**

For clinical programmers, this is a nightmare. Clinical TLFs (Tables, Listings, Figures) absolutely require the reference row to be displayed (e.g., displaying `1.00 (Reference)` in the Odds Ratio column).

### The Power of `{parameters}`
The `{parameters}` package is built for reporting. It is capable of intelligently injecting those missing reference rows back into the data frame, populating them with `0` (or `1` if exponentiated) and `NA` for p-values.

---

## Exercise 10.3 — Visualizing the Missing Reference Row

We have pre-fit a logistic regression model (`model_fit`) predicting response by `TRT01A` and `AGE`. `TRT01A` has three levels: Placebo, Xanomeline Low Dose, and Xanomeline High Dose. Placebo is the reference.

Let's use the standard `broom::tidy()` function to extract the coefficients and observe the missing reference row.

```{r ch10-ex3, exercise=TRUE, exercise.lines=10}
# Extract coefficients using broom::tidy
tidy_broom <- broom::tidy(model_fit)

# View the term and estimate columns
tidy_broom %>%
  select(term, estimate, p.value)
```

```{r ch10-ex3-solution}
tidy_broom <- broom::tidy(model_fit)

tidy_broom %>%
  select(term, estimate, p.value)
```

```{r ch10-ex3-check}
grade_result(
  pass_if(~ TRUE, "Notice that 'Placebo' is nowhere to be found! Only the Low Dose and High Dose terms appear."),
  fail_if(~ FALSE, "")
)
```

---

## 10.4 broom.helpers: The Rosetta Stone

We want the strict, predictable column names of `{broom}`, but we desperately need the reference rows and variable metadata provided by `{parameters}`. 

Enter the **`{broom.helpers}`** package. 

When you use `{cardx}` to build a regression ARD, it doesn't call `broom` directly. It calls `broom.helpers::tidy_plus_plus()`. This incredible wrapper function does several things simultaneously:
1. It extracts the model math.
2. It standardizes all column names to the `{broom}` convention.
3. **It magically rebuilds and injects the reference rows.**
4. It attaches the original variable labels from the dataset.

Think of `{broom.helpers}` as a powerful SAS macro that cleans up your `ODS OUTPUT`, merges it with `PROC CONTENTS` to get labels, and inserts dummy rows for reference classes automatically.

---

## Exercise 10.4 — Bringing Back the Reference Row

Let's run the exact same model through `broom.helpers::tidy_plus_plus()`.

```{r ch10-ex4, exercise=TRUE, exercise.lines=10}
# Extract coefficients using broom.helpers
tidy_helpers <- broom.helpers::tidy_plus_plus(model_fit)

# Select the variable, label, and reference_row indicator
tidy_helpers %>%
  select(variable, label, reference_row, estimate)
```

```{r ch10-ex4-solution}
tidy_helpers <- broom.helpers::tidy_plus_plus(model_fit)

tidy_helpers %>%
  select(variable, label, reference_row, estimate)
```

```{r ch10-ex4-check}
grade_result(
  pass_if(~ TRUE, "Brilliant! You can now see the Placebo row has been safely restored, labeled as the reference row, and given an estimate of 0 (which becomes OR=1 when exponentiated)."),
  fail_if(~ FALSE, "")
)
```

---

## 10.5 Customizing the Tidier in cardx

Because `{cardx::ard_regression()}` uses `tidy_plus_plus()` by default, your ARDs will automatically contain the reference levels needed for clinical tables!

If you ever find yourself working with an exotic model class that `{broom.helpers}` struggles with, you can bypass this automation and force `{cardx}` to use a specific tidying function via the `tidy_fun` argument.

```r
# Example of overriding the default tidier
ard_reg <- ard_regression(
  model_fit,
  tidy_fun = broom::tidy # Forces the use of strict broom
)
```

In 99% of clinical trial scenarios, however, the default `{broom.helpers}` integration works perfectly right out of the box.

---

## Exercise 10.5 — Creating the Regression ARD

Finally, let's wrap our model in `ard_regression()` to see how `{cardx}` takes the `broom.helpers` output and transforms it into the standard, long-format ARD structure you learned in Chapter 9.

```{r ch10-ex5, exercise=TRUE, exercise.lines=12}
# Create the ARD for the regression model
# Let's also ask it to exponentiate the coefficients into Odds Ratios
reg_ard <- ard_regression(model_fit, exponentiate = TRUE)

# Format and view the ARD
reg_ard %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt) %>%
  filter(variable == "TRT01A") # Focus only on the Treatment variable
```

```{r ch10-ex5-solution}
reg_ard <- ard_regression(model_fit, exponentiate = TRUE)

reg_ard %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt) %>%
  filter(variable == "TRT01A")
```

```{r ch10-ex5-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result), "Excellent! You can see the Placebo reference row is fully intact within the ARD structure, showing an Odds Ratio estimate of 1."),
  fail_if(~ TRUE, "Make sure to run the ard_regression code.")
)
```

---

## Chapter 10 Knowledge Check

```{r ch10-quiz, echo=FALSE}
quiz(
  caption = "Chapter 10 — Check Your Understanding",

  question(
    "What is the equivalent of SAS's `ODS OUTPUT ParameterEstimates = ds;` in the R modeling ecosystem?",
    answer("`glm()`"),
    answer("`summary()`"),
    answer("Model Tidiers like `broom::tidy()`", correct = TRUE),
    answer("`PROC PRINT`"),
    allow_retry = TRUE
  ),

  question(
    "Why is the `{broom}` package alone insufficient for clinical reporting of categorical variables?",
    answer("It crashes when trying to extract odds ratios."),
    answer("It strictly removes reference class levels (e.g. Placebo) from the output because their mathematical coefficient is 0.", correct = TRUE),
    answer("It uses `Parameter` instead of `term` for column names."),
    answer("It cannot calculate p-values."),
    allow_retry = TRUE
  ),

  question(
    "What crucial role does `{broom.helpers::tidy_plus_plus()}` play when `{cardx}` builds a regression ARD?",
    answer("It calculates the original model math."),
    answer("It generates standard TLF RTF files."),
    answer("It standardizes column names, adds variable labels, and intelligently reconstructs the missing reference rows.", correct = TRUE),
    answer("It converts datasets to SAS transport files."),
    allow_retry = TRUE
  )
)
```
