---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 6: Consuming ARD"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.9
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Learn how to build formatted tables from pre-calculated ARDs using gtsummary, apply global themes, modify aesthetics, and export clinical tables.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(gtsummary)
library(flextable)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")

# Add a missing value to test missing replacements
adsl$AGE[1] <- NA

# Pre-compute an ARD for downstream exercises to save time
ard_demog_pre <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A,
  .missing = TRUE
)
```

---

## Chapter 6: Consuming ARD

### 6.1 Downstream Table Rendering

Once you have calculated statistical results and stored them in the standardized ARD format, you can pass them to downstream reporting engines to render the final table. 

- **`{gtsummary}`**: Provides `tbl_ard_summary()`. This function accepts an ARD object and instantly creates a publication-ready clinical summary table. Because all statistics are already calculated, rendering takes milliseconds.

---

## Exercise 6.1 — Render a Demographics Table from an ARD

Before we run the exercise, let's exhaustively review the parameters of `tbl_ard_summary()`:
- `cards`: An ARD object of class `card` (typically created with `ard_stack()`).
- `by`: A single column to stratify the summary statistics by (Note: NO leading dot, unlike `ard_stack()`).
- `include`: Variables to include in the summary table (e.g., `c(AGE, SEX)`).
- `statistic`: A formula specifying the summary statistics for each variable.
- `type`: Specifies the summary type (`"continuous"`, `"continuous2"`, `"categorical"`, `"dichotomous"`).
- `label`: Overrides default labels in the summary table (e.g., `list(AGE = "Age, years")`).
- `missing`: Dictates if missing values are presented (`"no"`, `"ifany"`, `"always"`).
- `missing_text`: String indicating text shown on the missing row.
- `missing_stat`: Statistic to show on the missing row.
- `overall`: When `TRUE`, appends an overall column (requires `ard_stack(.overall = TRUE)`).

```{r ch6-ex1, exercise=TRUE, exercise.lines=20}
# 1. Generate the ARD data frame
ard_demog <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A,
  .missing = TRUE
)

# 2. Render the table using the pre-computed ARD
# Hint: call tbl_ard_summary() passing card = ard_demog, include = c(AGE, SEX), by = TRT01A, missing = "always"

demog_table <- ___(
  cards = ard_demog,
  include = c(AGE, SEX),
  by = TRT01A,
  missing = "always"
)

# Render table
demog_table
```

```{r ch6-ex1-hint-1}
# Call tbl_ard_summary(cards = ard_demog, include = c(AGE, SEX), by = TRT01A, missing = "always")
```

```{r ch6-ex1-solution}
tbl_ard_summary(
  cards = ard_demog,
  include = c(AGE, SEX),
  by = TRT01A,
  missing = "always"
)
```

```{r ch6-ex1-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Fantastic! `{gtsummary}` read the ARD and formatted the table. If you want to change titles or footnotes, you can do so in seconds without recalculating any statistics."),
  fail_if(~ TRUE, "Check that you are calling tbl_ard_summary() with cards = ard_demog, include = c(AGE, SEX), by = TRT01A, and missing = \"always\".")
)
```

---

## Exercise 6.2 — Apply a Global Theme

Before generating your table, you can set a global theme to adhere to specific journal or organizational standards. 

- **`theme_gtsummary_journal()`**: Takes a `journal` argument (e.g., `"jama"`, `"nejm"`, `"lancet"`).
- **`theme_gtsummary_compact()`**: Reduces font size and cell padding for large tables.
- **`reset_gtsummary_theme()`**: Clears all themes.

```{r ch6-ex2, exercise=TRUE, exercise.lines=15}
# Set the global theme to "jama" (Journal of the American Medical Association)
# Hint: Call theme_gtsummary_journal(journal = "jama")

___

# Render the table
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
)
```

```{r ch6-ex2-hint-1}
# Call theme_gtsummary_journal(journal = "jama") at the top of the block.
```

```{r ch6-ex2-solution}
theme_gtsummary_journal(journal = "jama")

tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
)
```

```{r ch6-ex2-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Great job! Setting a theme changes global aesthetics instantly without touching the underlying statistical data."),
  fail_if(~ TRUE, "Ensure you called theme_gtsummary_journal(journal = \"jama\") before rendering.")
)
```

---

## Selectors and Helper Functions

`{gtsummary}` and `{cards}` rely heavily on helper functions that "select" parts of the table or format data. If you don't know what these are, the code looks like magic!

### 1. Selectors
Selectors grab specific columns or rows of the table without you needing to explicitly type out their names:
- **`all_stat_cols()`**: Automatically selects all columns containing statistical results (e.g., the `stat_1`, `stat_2` columns generated by splitting by `TRT01A`).
- **`all_continuous()`** / **`all_categorical()`**: Automatically targets all rows/variables that are of the continuous or categorical summary type.

### 2. Formatters
Formatters are R functions that dictate how numbers are printed as strings:
- **`style_number`**: A highly versatile `{cards}` function. You can pass it arguments like `digits` to control decimal places (e.g., `style_number(digits = 2)`).
- **`style_percent`**: Specifically designed for formatting percentages, appending the `%` sign.

---

## Exercise 6.3 — Standard Aesthetics

The core engine behind all styling in `{gtsummary}` is **`modify_table_styling()`**. It has 15 parameters including `columns`, `rows`, `label`, `hide`, `footnote`, `align`, `missing_symbol`, `fmt_fun`, and `text_interpret`.

However, `{gtsummary}` provides convenient wrappers for standard typography and headers. Let's practice them:
- **`modify_header()`**: Modifies column headers.
- **`modify_spanning_header()`**: Adds headers that group multiple columns together. You can target all statistics using `all_stat_cols()`, or target specific columns using their internal names (e.g. `c(stat_3, stat_4)`).
- **`bold_labels()`** and **`italicize_levels()`**: Require no arguments, applying globally to row headers.

```{r ch6-ex3, exercise=TRUE, exercise.lines=18}
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
) %>%
  # Bold the row labels (AGE, SEX)
  ___() %>%
  # 1. Add a spanning header over ALL statistic columns
  # Hint: modify_spanning_header(all_stat_cols() ~ "**All Treatments**")
  ___ %>%
  # 2. Add a spanning header over ONLY the Xanomeline Active columns (stat_3 and stat_4)
  # Hint: modify_spanning_header(c(stat_3, stat_4) ~ "**Active Treatments**")
  ___
```

```{r ch6-ex3-hint-1}
# Chain the following functions:
# bold_labels() %>%
# modify_spanning_header(all_stat_cols() ~ "**All Treatments**") %>%
# modify_spanning_header(c(stat_3, stat_4) ~ "**Active Treatments**")
```

```{r ch6-ex3-solution}
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
) %>%
  bold_labels() %>%
  modify_spanning_header(all_stat_cols() ~ "**All Treatments**") %>%
  modify_spanning_header(c(stat_3, stat_4) ~ "**Active Treatments**")
```

```{r ch6-ex3-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Beautiful! Using modifiers is safe and fast because it strictly formats the presentation layer."),
  fail_if(~ TRUE, "Check that you called bold_labels() and modify_spanning_header().")
)
```

---

## Exercise 6.4 — Advanced Table Styling (Missing, Formatting, and Abbreviations)

Now let's use the advanced wrapper functions mapping to `modify_table_styling()`. 

- **`modify_abbreviation()`**: A specialized footnote function for defining acronyms in the footer.
- **`modify_missing_symbol()`**: Replaces `NA` blanks in the table body with a specific character string (`symbol` argument).
- **`modify_fmt_fun()`**: Updates the formatting function for specific statistics *after* the table is generated.
- **`text_interpret` parameter**: Instructs `{gtsummary}` to interpret strings as `"md"` (markdown) or `"html"`. Available in all `modify_*` wrappers!

Use these parameters below to heavily customize our table!

```{r ch6-ex4, exercise=TRUE, exercise.lines=25}
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A,
  missing = "always"
) %>%
  # 1. Update the header using markdown, setting text_interpret to "md"
  modify_header(
    label = "**Demographics**",
    text_interpret = "md"
  ) %>%
  # 2. Add an abbreviation to the footer
  # Hint: modify_abbreviation("TRT = Treatment")
  ___ %>%
  # 3. Replace the NA missing count blanks with the text "Not Assessed"
  # Hint: modify_missing_symbol(all_stat_cols(), symbol = "Not Assessed", rows = !is.na(variable))
  ___ %>%
  # 4. Change the formatting of the continuous statistic to 2 decimal places
  # Hint: modify_fmt_fun(update = all_continuous() ~ style_number)
  ___
```

```{r ch6-ex4-hint-1}
# Chain:
# modify_abbreviation("TRT = Treatment") %>%
# modify_missing_symbol(all_stat_cols(), symbol = "Not Assessed", rows = !is.na(variable)) %>%
# modify_fmt_fun(update = all_continuous() ~ style_number)
```

```{r ch6-ex4-solution}
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A,
  missing = "always"
) %>%
  modify_header(
    label = "**Demographics**",
    text_interpret = "md"
  ) %>%
  modify_abbreviation("TRT = Treatment") %>%
  modify_missing_symbol(all_stat_cols(), symbol = "Not Assessed", rows = !is.na(variable)) %>%
  modify_fmt_fun(update = all_continuous() ~ style_number)
```

```{r ch6-ex4-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Incredible! You have mastered the advanced styling API of `{gtsummary}`!"),
  fail_if(~ TRUE, "Ensure you chained modify_abbreviation(), modify_missing_symbol(), and modify_fmt_fun() correctly.")
)
```

---

## Exercise 6.5 — Exporting the Table

Clinical tables need to be exported to Microsoft Word, RTF, or PDF. `{gtsummary}` integrates seamlessly with the `{flextable}` package for Word outputs.

```{r ch6-ex5, exercise=TRUE, exercise.lines=12}
# 1. Create the base table
my_table <- tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
)

# 2. Convert to flextable
# Hint: Pipe my_table into as_flex_table()
flex_out <- ___

# Display the flextable (in practice, you'd pipe this into flextable::save_as_docx())
flex_out
```

```{r ch6-ex5-hint-1}
# flex_out <- my_table %>% as_flex_table()
```

```{r ch6-ex5-solution}
flex_out <- my_table %>% as_flex_table()
flex_out
```

```{r ch6-ex5-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Success! Converting to a flextable object is the final step before exporting to .docx."),
  fail_if(~ TRUE, "Ensure you pipe my_table into as_flex_table().")
)
```

---

## Chapter 6 Knowledge Check

```{r ch6-quiz, echo=FALSE}
quiz(
  caption = "Chapter 6 — Check Your Understanding",

  question(
    "Which `{gtsummary}` parameter explicitly replaces `NA` missing values with a string character?",
    answer("modify_header()"),
    answer("modify_missing_symbol()", correct = TRUE),
    answer("missing_text"),
    answer("theme_gtsummary_journal()"),
    allow_retry = TRUE
  ),

  question(
    "Which argument is used within `modify_header()` to ensure asterisks are rendered as bold text?",
    answer("label"),
    answer("text_interpret = \"md\"", correct = TRUE),
    answer("fmt_fun"),
    answer("bold_labels()"),
    allow_retry = TRUE
  ),
  
  question(
    "How do you export a `{gtsummary}` table to Microsoft Word?",
    answer("Pipe it into write.csv()"),
    answer("Pipe it into as_flex_table() and use save_as_docx()", correct = TRUE),
    answer("Use export_word()"),
    answer("Print it and copy-paste"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 6 Summary {.congratulations}

You now know:
- The advantages of the **Separation of Concerns** paradigm in clinical reporting.
- The exhaustive parameters of `tbl_ard_summary()` (`missing`, `label`, `statistic`).
- The low-level `modify_table_styling()` engine and advanced wrappers (`modify_fmt_fun`, `modify_abbreviation`, `modify_missing_symbol`).
- The usage of `text_interpret = "md"` to process markdown syntaxes.
- How to export tables using **`as_flex_table()`**.

**Next:** Chapter 7 — Advanced Workflows (Writing custom stats and mapping to CDISC ARS JSON)
