---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 6: Consuming ARD"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Learn how to build formatted tables from pre-calculated ARDs using gtsummary and other downstream reporting engines.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(gtsummary)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
```

---

## Chapter 6: Consuming ARD

### 6.1 Downstream Table Rendering

Once you have calculated statistical results and stored them in the standardized ARD format, you can pass them to downstream reporting engines to render the final table. 

- **`{gtsummary}`**: Provides `tbl_ard_summary()`. This function accepts an ARD object and instantly creates a publication-ready clinical summary table. Because all statistics are already calculated, rendering takes milliseconds.
- **`{rtables}`**: A cell-by-cell layout engine that uses ARD data frames as input to build complex, multi-level clinical tables.

---

## Exercise 6.1 — Render a Demographics Table from an ARD

Let's calculate an ARD containing Age and Sex statistics grouped by treatment, and then pass it to `tbl_ard_summary()` to display the final clinical table.

```{r ch6-ex1, exercise=TRUE, exercise.lines=20}
# 1. Generate the ARD data frame
ard_demog <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  by = TRT01A
)

# 2. Render the table using the pre-computed ARD
# Hint: call tbl_ard_summary() passing card = ard_demog, variables = c(AGE, SEX), and by = TRT01A

demog_table <- ___(
  cards = ard_demog,
  variables = c(AGE, SEX),
  by = TRT01A
)

# Render table
demog_table
```

```{r ch6-ex1-hint-1}
# Call tbl_ard_summary(cards = ard_demog, variables = c(AGE, SEX), by = TRT01A)
```

```{r ch6-ex1-solution}
tbl_ard_summary(
  cards = ard_demog,
  variables = c(AGE, SEX),
  by = TRT01A
)
```

```{r ch6-ex1-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Fantastic! `{gtsummary}` read the ARD and formatted the table. If you want to change titles or footnotes, you can do so in seconds without recalculating any statistics."),
  fail_if(~ TRUE, "Check that you are calling tbl_ard_summary() with cards = ard_demog, variables = c(AGE, SEX), and by = TRT01A.")
)
```

---

## Chapter 6 Knowledge Check

```{r ch6-quiz, echo=FALSE}
quiz(
  caption = "Chapter 6 — Check Your Understanding",

  question(
    "Which `{gtsummary}` function renders a formatted clinical table from a pre-calculated ARD data frame?",
    answer("tbl_summary()"),
    answer("tbl_ard_summary()", correct = TRUE),
    answer("tbl_render_ard()"),
    answer("tbl_regression()"),
    allow_retry = TRUE
  ),

  question(
    "What is a primary advantage of separating statistical calculations (ARD) from table rendering?",
    answer("The code runs slower but is more accurate"),
    answer("You can tweak the table's cosmetic formatting without having to recompute expensive statistical models", correct = TRUE),
    answer("It allows you to skip writing R code"),
    answer("It merges R and SAS codebases"),
    allow_retry = TRUE
  ),

  question(
    "Which pharmaverse package is a powerful layout engine used for complex, nested clinical tables (like AE SOC/PT tables)?",
    answer("ggplot2"),
    answer("rtables", correct = TRUE),
    answer("gtsummary"),
    answer("datasetjson"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 6 Summary {.congratulations}

You now know:
- The advantages of the **Separation of Concerns** paradigm in clinical reporting.
- How to pass pre-calculated ARD objects to **`tbl_ard_summary()`** in `{gtsummary}`.
- The role of **`{rtables}`** as a layout engine that consumes ARDs to build nested clinical tables.

**Next:** Chapter 7 — Advanced Workflows (Writing custom stats and mapping to CDISC ARS JSON)
