---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 6: Consuming ARD"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.3
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Learn how to build formatted tables from pre-calculated ARDs using gtsummary, apply global themes, modify aesthetics, and export clinical tables.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(gtsummary)
library(flextable)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")

# Pre-compute an ARD for downstream exercises to save time
ard_demog_pre <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A
)
```

---

## Chapter 6: Consuming ARD

### 6.1 Downstream Table Rendering

Once you have calculated statistical results and stored them in the standardized ARD format, you can pass them to downstream reporting engines to render the final table. 

- **`{gtsummary}`**: Provides `tbl_ard_summary()`. This function accepts an ARD object and instantly creates a publication-ready clinical summary table. Because all statistics are already calculated, rendering takes milliseconds.
- **`{rtables}`**: A cell-by-cell layout engine that uses ARD data frames as input to build complex, multi-level clinical tables.

---

## Exercise 6.1 — Render a Demographics Table from an ARD

Let's calculate an ARD containing Age and Sex statistics grouped by treatment, and then pass it to `tbl_ard_summary()` to display the final clinical table.

```{r ch6-ex1, exercise=TRUE, exercise.lines=20}
# 1. Generate the ARD data frame
ard_demog <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A
)

# 2. Render the table using the pre-computed ARD
# Hint: call tbl_ard_summary() passing card = ard_demog, include = c(AGE, SEX), and by = TRT01A

demog_table <- ___(
  cards = ard_demog,
  include = c(AGE, SEX),
  by = TRT01A
)

# Render table
demog_table
```

```{r ch6-ex1-hint-1}
# Call tbl_ard_summary(cards = ard_demog, include = c(AGE, SEX), by = TRT01A)
```

```{r ch6-ex1-solution}
tbl_ard_summary(
  cards = ard_demog,
  include = c(AGE, SEX),
  by = TRT01A
)
```

```{r ch6-ex1-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Fantastic! `{gtsummary}` read the ARD and formatted the table. If you want to change titles or footnotes, you can do so in seconds without recalculating any statistics."),
  fail_if(~ TRUE, "Check that you are calling tbl_ard_summary() with cards = ard_demog, include = c(AGE, SEX), and by = TRT01A.")
)
```

---

## Exercise 6.2 — Apply a Global Theme

`{gtsummary}` comes with built-in themes that automatically adjust padding, borders, and typography to meet specific journal guidelines (like JAMA or NEJM) or layout needs (like a compact view).

You must call the theme function **before** creating the table.

```{r ch6-ex2, exercise=TRUE, exercise.lines=15}
# Set the global theme to "jama" (Journal of the American Medical Association)
# Hint: Call theme_gtsummary_journal("jama")

___

# Render the table (we use ard_demog_pre which was pre-calculated for you)
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
)
```

```{r ch6-ex2-hint-1}
# Call theme_gtsummary_journal("jama") at the top of the block.
```

```{r ch6-ex2-solution}
theme_gtsummary_journal("jama")

tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
)
```

```{r ch6-ex2-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Great job! Setting a theme changes global aesthetics instantly without touching the underlying statistical data."),
  fail_if(~ TRUE, "Ensure you called theme_gtsummary_journal(\"jama\") before rendering.")
)
```

---

## Exercise 6.3 — Modify Table Aesthetics

You can pipe `tbl_ard_summary()` into various modifier functions to change table headers, add spanning titles, or bold variables.

Modify the table below by bolding the variable labels and adding a spanning header over the statistic columns.

```{r ch6-ex3, exercise=TRUE, exercise.lines=15}
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
) %>%
  # Bold the row labels (AGE, SEX)
  ___() %>%
  # Add a spanning header over the columns
  # Hint: use modify_spanning_header(all_stat_cols() ~ "**Treatment Group**")
  ___
```

```{r ch6-ex3-hint-1}
# Call bold_labels() %>% modify_spanning_header(all_stat_cols() ~ "**Treatment Group**")
```

```{r ch6-ex3-solution}
tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
) %>%
  bold_labels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Treatment Group**")
```

```{r ch6-ex3-check}
grade_result(
  pass_if(~ inherits(.result, "gtsummary"), "Beautiful! Using modifiers is safe and fast because it strictly formats the presentation layer."),
  fail_if(~ TRUE, "Check that you called bold_labels() and modify_spanning_header().")
)
```

---

## Exercise 6.4 — Exporting the Table

Clinical tables need to be exported to Microsoft Word, RTF, or PDF. `{gtsummary}` integrates seamlessly with the `{flextable}` package for Word outputs.

```{r ch6-ex4, exercise=TRUE, exercise.lines=12}
# 1. Create the base table
my_table <- tbl_ard_summary(
  cards = ard_demog_pre,
  include = c(AGE, SEX),
  by = TRT01A
)

# 2. Convert to flextable
# Hint: Pipe my_table into as_flex_table()
flex_out <- ___

# Display the flextable (in practice, you'd pipe this into flextable::save_as_docx())
flex_out
```

```{r ch6-ex4-hint-1}
# flex_out <- my_table %>% as_flex_table()
```

```{r ch6-ex4-solution}
flex_out <- my_table %>% as_flex_table()
flex_out
```

```{r ch6-ex4-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Success! Converting to a flextable object is the final step before exporting to .docx."),
  fail_if(~ TRUE, "Ensure you pipe my_table into as_flex_table().")
)
```

---

## Chapter 6 Knowledge Check

```{r ch6-quiz, echo=FALSE}
quiz(
  caption = "Chapter 6 — Check Your Understanding",

  question(
    "Which `{gtsummary}` function renders a formatted clinical table from a pre-calculated ARD data frame?",
    answer("tbl_summary()"),
    answer("tbl_ard_summary()", correct = TRUE),
    answer("tbl_render_ard()"),
    answer("tbl_regression()"),
    allow_retry = TRUE
  ),

  question(
    "What is a primary advantage of separating statistical calculations (ARD) from table rendering?",
    answer("The code runs slower but is more accurate"),
    answer("You can tweak the table's cosmetic formatting without having to recompute expensive statistical models", correct = TRUE),
    answer("It allows you to skip writing R code"),
    answer("It merges R and SAS codebases"),
    allow_retry = TRUE
  ),
  
  question(
    "How do you export a `{gtsummary}` table to Microsoft Word?",
    answer("Pipe it into write.csv()"),
    answer("Pipe it into as_flex_table() and use save_as_docx()", correct = TRUE),
    answer("Use export_word()"),
    answer("Print it and copy-paste"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 6 Summary {.congratulations}

You now know:
- The advantages of the **Separation of Concerns** paradigm in clinical reporting.
- How `{gtsummary}` maps ARD variables and stats to rows and columns natively.
- How to apply global themes using **`theme_gtsummary_*()`**.
- How to format tables dynamically using modifiers like **`modify_spanning_header()`**.
- How to export tables using **`as_flex_table()`**.

**Next:** Chapter 7 — Advanced Workflows (Writing custom stats and mapping to CDISC ARS JSON)
