---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 2: Core cards Operations"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Learn to generate continuous, categorical, dichotomous, and missingness summaries using the core functions of the cards R package.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
```

---

## Chapter 2: Core cards Operations

### 2.1 The `{cards}` Function Family — One Per Data Type

In Chapter 1 you learned what an ARD *is*. This chapter teaches you how to *produce* one. The `{cards}` package gives you a small, sharp set of functions — each one specialized for a single data-type pattern that appears constantly in clinical reporting.

```
                       INPUT: tidy data frame (ADaM)
                                   |
        +-----------------+--------+--------+----------------+
        |                 |                 |                |
        v                 v                 v                v
  +-------------+  +---------------+  +---------------+  +-------------+
  |ard_continuous|  |ard_categorical|  |ard_dichotomous|  |ard_missing  |
  +-------------+  +---------------+  +---------------+  +-------------+
  numeric vars     factor / char     yes/no flags         any var
  AGE, BMI, AVAL   SEX, RACE         SAFFL, TRTEMFL       missingness
  PROC MEANS       PROC FREQ         PROC FREQ (1 cell)   PROC MEANS NMISS
        |                 |                 |                |
        +-----------------+--------+--------+----------------+
                                   |
                                   v
                       OUTPUT: standardized ARD rows
                       (same schema you saw in Ch.1)
                                   |
                                   v
                          ard_total_n() adds
                          denominator rows used
                          for "N=…" column headers
```

The principle: pick the function that matches the *data type and the question*, not one giant function with many switches.

| Function              | Use for                                                | Closest SAS analog                              |
|-----------------------|--------------------------------------------------------|-------------------------------------------------|
| `ard_continuous()`    | Numeric variables: AGE, WEIGHT, AVAL, CHG, BMI         | `PROC MEANS` / `PROC SUMMARY`                   |
| `ard_categorical()`   | Multi-level categorical: SEX, RACE, AGEGR1, SEVERITY   | `PROC FREQ` (one-way, two-way)                  |
| `ard_dichotomous()`   | Yes/No flags shown as a single "Y" row: SAFFL, TRTEMFL | `PROC FREQ` then keeping one row in a data step |
| `ard_missing()`       | Missingness counts/percentages per variable            | `NMISS` in `PROC MEANS`                         |
| `ard_total_n()`       | Treatment-arm subject totals for table headers         | `_FREQ_` after `PROC SUMMARY` on a `BY` group   |

Every function shares the same three core arguments:

- `data` — the data frame (your filtered ADaM).
- `variables` — which column(s) to summarize.
- `by` — the stratification variable (treatment arm, visit, etc.).

And every function returns the same ARD schema (`group1`, `variable`, `stat_name`, `stat`, `stat_fmt`, ...). That uniformity is what lets you `bind_rows()` outputs from different `{cards}` functions into a single layout-ready dataset.

> **SAS analog mental model:** imagine a parallel universe where `PROC MEANS`, `PROC FREQ`, and a missingness PROC all emitted `ODS OUTPUT` datasets with identical column names. You could `SET` them together without a single `RENAME`. That is `{cards}`.

---

### 2.2 `ard_continuous()` — The Continuous Workhorse

#### 2.2.1 What it produces

For every (treatment arm × variable) combination, `ard_continuous()` emits one row per statistic. The default statistics are familiar: `N`, `mean`, `sd`, `median`, `p25`, `p75`, `min`, `max`, and a count of missing values.

```
       PROC MEANS DATA=adsl;                |    ard_continuous(
         CLASS TRT01A;                      |      data = adsl,
         VAR   AGE;                         |      variables = AGE,
       RUN;                                 |      by = TRT01A
                                            |    )
       --------------------------------     |    --------------------------------
       Wide output:                         |    Long output:
         TRT01A   N   Mean   Std   ...      |      group1_level | stat_name | stat_fmt
         Placebo  86  75.21  8.12  ...      |      Placebo      | n         | 86
         Xan Hi   84  74.38  7.89  ...      |      Placebo      | mean      | 75.2
                                            |      Placebo      | sd        | 8.12
                                            |      ...
```

This long shape is the single biggest mindset shift for a SAS programmer. `PROC MEANS` gives you a wide row per `BY` group with one column per statistic; `ard_continuous()` gives you a long row per *(BY group × statistic)*. The data is the same; the orientation is different. The long shape is what makes the output uniform across all `{cards}` functions and trivial to drive a layout from.

#### 2.2.2 Anatomy of an `ard_continuous()` call

```r
ard_continuous(
  data      = adsl,        # the input frame, like DATA= in a PROC
  variables = c(AGE, BMI), # which numeric columns to summarize
  by        = TRT01A,      # the BY / CLASS variable
  statistic = ~ list(      # OPTIONAL: customize which stats are computed
                "N", "mean", "sd", "median", "min", "max"
              )
)
```

- **`variables`** can be a single bare name (`AGE`) or several (`c(AGE, BMI, HEIGHT)`); every variable becomes its own block of rows in the ARD.
- **`by`** is the grouping variable. Multiple grouping variables become `c(TRT01A, AVISIT)`, producing nested `group1`, `group2`.
- **`statistic`** lets you override the default list. If you only want `n`, `mean`, and `sd`, say so — the ARD shrinks accordingly. This is the `{cards}` equivalent of writing `MEAN STD;` after a `PROC MEANS` statement.

> **SAS analog:** `VAR AGE BMI;` → `variables = c(AGE, BMI)`. `CLASS TRT01A;` → `by = TRT01A`. `PROC MEANS N MEAN STD;` → `statistic = ~ list("N", "mean", "sd")`.

#### 2.2.3 How the output maps to a table cell

```
ARD row:                          Table cell:
  variable      = AGE             +------------------------+
  group1_level  = Placebo         |  Age, Mean (SD)        |
  stat_name     = mean            |  Placebo: 75.2 (8.12)  |
  stat_fmt      = "75.2"          +------------------------+
  +
  variable      = AGE
  group1_level  = Placebo
  stat_name     = sd
  stat_fmt      = "8.12"
```

The layout engine downstream simply glues `mean_fmt` and `sd_fmt` together into the string `"75.2 (8.12)"` and drops it in the right cell. The arithmetic, the rounding, and the cell composition are three separate, individually verifiable steps.

---

## Exercise 2.1 — Summarize Continuous Data

Summarize the continuous variable `AGE` in the `adsl` dataset, grouped by treatment arm (`TRT01A`).

```{r ch2-ex1, exercise=TRUE, exercise.lines=10}
# Generate the continuous ARD for AGE grouped by TRT01A
# Hint: use ard_continuous() with variables = AGE and by = TRT01A

ard_continuous(
  data = adsl,
  variables = ___,
  by = ___
)
```

```{r ch2-ex1-hint-1}
# Pass AGE to variables and TRT01A to by
ard_continuous(adsl, variables = AGE, by = TRT01A)
```

```{r ch2-ex1-solution}
ard_continuous(adsl, variables = AGE, by = TRT01A)
```

```{r ch2-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "mean" %in% .result$stat_name, "Correct! ard_continuous successfully generated the continuous statistics (mean, sd, etc.) for AGE by TRT01A."),
  fail_if(~ TRUE, "Ensure you are using ard_continuous with variables = AGE and by = TRT01A.")
)
```

> **Read the output:** scroll through the `stat_name` column. You will see `n`, `mean`, `sd`, `median`, `p25`, `p75`, `min`, `max`, and `N_miss` — one row each per treatment arm. Three arms × nine statistics = roughly 27 rows for a single variable. That density is the price of the long format and the reason ARDs are so easy to filter, slice, and re-layout downstream.

---

### 2.3 `ard_categorical()` — Frequencies, Levels, and Denominators

#### 2.3.1 What it produces

For every (treatment arm × variable × level) combination, `ard_categorical()` emits two rows: the count `n` and the percentage `p`. It optionally produces the denominator `N` as well.

```
       PROC FREQ DATA=adsl;          |    ard_categorical(
         TABLES SEX*TRT01A;          |      data = adsl,
       RUN;                          |      variables = SEX,
                                     |      by = TRT01A
                                     |    )
       -----------------------       |    -----------------------
       Wide cross-tab:               |    Long ARD:
         SEX  Placebo  XanHi         |      group1_level | variable_grp_level | stat_name | stat_fmt
         F    53(61.6) 50(59.5)      |      Placebo      | F                  | n         | 53
         M    33(38.4) 34(40.5)      |      Placebo      | F                  | p         | 61.6%
                                     |      Placebo      | M                  | n         | 33
                                     |      Placebo      | M                  | p         | 38.4%
                                     |      Xan High     | F                  | n         | 50
                                     |      ...
```

A two-level categorical × three treatment arms produces *six cells*, and each cell becomes *two rows* (n and p) — twelve rows total. Add `N` and it becomes eighteen. The row count balloons quickly. That is normal. ARDs are meant to be filtered, not read whole.

#### 2.3.2 Denominator choices — the most important option

This is where many SAS programmers stumble on the first read, because `PROC FREQ` hides the choice inside a `PERCENT` keyword. In `{cards}`, the denominator is explicit.

| `denominator` value | Percentage is computed against...                                | SAS analog                            |
|---------------------|------------------------------------------------------------------|---------------------------------------|
| `"column"` (default)| Total non-missing observations within each `by` (treatment arm)  | `COLPERCENT` in `PROC FREQ`           |
| `"row"`             | Total observations within each level of the analysis variable    | `ROWPERCENT`                          |
| `"cell"`            | Total observations in the entire dataset                         | `PERCENT` (cell percentages)          |

Picking the wrong denominator silently produces wrong percentages. Always ask: *"What population does each percentage refer to?"* — that is the denominator.

> **Mnemonic:** column percent = "out of this treatment arm." Row percent = "out of all subjects in this category." Cell = "out of everyone."

#### 2.3.3 Level ordering

Level order in the ARD follows the level order of the variable in the input data. If `SEX` is a factor with levels `c("F", "M")`, the ARD rows appear in that order. If you need a custom order (Placebo first, active arms last), set the factor levels *before* calling `ard_categorical()`. Same rule as `PROC FREQ ORDER=DATA` vs. `ORDER=FORMATTED`.

---

## Exercise 2.2 — Summarize Categorical Data

Summarize the categorical variable `SEX` in `adsl`, grouped by treatment arm (`TRT01A`).

```{r ch2-ex2, exercise=TRUE, exercise.lines=10}
# Calculate categorical counts and percentages for SEX by TRT01A
# Hint: use ard_categorical()

ard_categorical(
  data = adsl,
  variables = ___,
  by = ___
)
```

```{r ch2-ex2-hint-1}
# Pass SEX to variables and TRT01A to by
ard_categorical(adsl, variables = SEX, by = TRT01A)
```

```{r ch2-ex2-solution}
ard_categorical(adsl, variables = SEX, by = TRT01A)
```

```{r ch2-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("F", "M") %in% .result$variable_grp_level), "Excellent! You generated categorical summaries. Notice that there are rows for both 'F' and 'M' levels."),
  fail_if(~ TRUE, "Verify that you are using ard_categorical() with variables = SEX and by = TRT01A.")
)
```

> **Check the math:** sum the `n` values for `variable_grp_level = "F"` and `"M"` within `group1_level = "Placebo"`. The total should equal the total Placebo subject count. That equality is what makes the default column denominator natural for demographics tables: each column tells you how the treatment arm breaks down by sex.

---

### 2.4 `ard_dichotomous()` — One Row Per Yes/No Flag

#### 2.4.1 Why it exists

Clinical data is full of binary flags: `SAFFL` ("Y"/"N"), `TRTEMFL` ("Y"/"N"), `FASFL`, `ITTFL`, `DTHFL`. In a typical table you only ever display the *positive* row — *"86 of 86 subjects (100%) were in the Safety Population"*, not also *"0 of 86 (0%) were not."*

`ard_categorical()` would happily produce both the `Y` and the `N` rows. `ard_dichotomous()` is a thin wrapper that produces *only the positive row*, saving downstream filtering and reducing the chance of an "N=0 (0%)" row leaking into your table.

```
                        ard_categorical(SAFFL)
                        +-------------------+
                        | Placebo | Y | n=86|   <-- kept
                        | Placebo | Y | p=1 |   <-- kept
                        | Placebo | N | n=0 |   <-- you didn't want this
                        | Placebo | N | p=0 |   <-- you didn't want this
                        +-------------------+

                        ard_dichotomous(SAFFL, value=list(SAFFL="Y"))
                        +-------------------+
                        | Placebo | Y | n=86|   <-- kept
                        | Placebo | Y | p=1 |   <-- kept
                        +-------------------+
```

#### 2.4.2 The `value` argument

`value = list(SAFFL = "Y")` says *"for the variable SAFFL, the positive level is 'Y'."* The list form lets you supply multiple variables in one call with possibly different positive levels:

```r
ard_dichotomous(
  data      = adsl,
  by        = TRT01A,
  variables = c(SAFFL, ITTFL, DTHFL),
  value     = list(SAFFL = "Y", ITTFL = "Y", DTHFL = "Y")
)
```

> **SAS analog:** `PROC FREQ` with `WHERE SAFFL = "Y";` then keeping only that level in the output dataset. `ard_dichotomous()` does it in one call without a `WHERE` mutation of the input data — important because the denominator should still reflect the full treatment arm (it does).

---

## Exercise 2.3 — Summarize Dichotomous Flag Variables

Clinical data contains many flag variables (like `SAFFL = "Y"`). We often only want to show the positive indicator in our table. Use `ard_dichotomous()` to summarize the `SAFFL` (Safety Population Flag) variable for the `"Y"` level, grouped by `TRT01A`.

```{r ch2-ex3, exercise=TRUE, exercise.lines=15}
# Summarize SAFFL where value is "Y"
# Hint: use ard_dichotomous() and specify value = list(SAFFL = "Y")

ard_dichotomous(
  data = adsl,
  by = TRT01A,
  variables = SAFFL,
  value = list(SAFFL = ___)
)
```

```{r ch2-ex3-hint-1}
# Supply "Y" to the SAFFL item in the value list
ard_dichotomous(adsl, variables = SAFFL, value = list(SAFFL = "Y"), by = TRT01A)
```

```{r ch2-ex3-solution}
ard_dichotomous(adsl, variables = SAFFL, value = list(SAFFL = "Y"), by = TRT01A)
```

```{r ch2-ex3-check}
grade_result(
  pass_if(~ is.data.frame(.result) && !("N" %in% .result$variable_grp_level), "Superb! ard_dichotomous kept only the rows for the 'Y' level, saving space and simplifying downstream table generation."),
  fail_if(~ TRUE, "Make sure you set value = list(SAFFL = 'Y') inside ard_dichotomous.")
)
```

> **Notice what is missing:** the output does not contain any row with `variable_grp_level = "N"`. That is the entire point of dichotomous. If you ever need both rows (rare for flag variables, common for disposition reasons), use `ard_categorical()` instead.

---

### 2.5 `ard_missing()` and `ard_total_n()` — The Helper Rows

These two short functions are easy to overlook and impossible to live without once you start assembling real table shells.

#### 2.5.1 `ard_missing()` — explicit missingness rows

Almost every regulatory table has a "Missing" line:

```
Age (years)
  n                     86
  Mean (SD)             75.2 (8.12)
  Median (Q1, Q3)       76.0 (70, 81)
  Min, Max              52, 89
  Missing               0
```

The default statistics of `ard_continuous()` already include `N_miss`, but for categorical variables or for cases where you want a dedicated, clearly-labeled missing summary, `ard_missing()` provides:

- `N_obs` — total observations.
- `N_miss` — count of `NA`.
- `p_miss` — percentage missing.

```r
ard_missing(
  data      = adsl,
  by        = TRT01A,
  variables = AGE
)
```

> **SAS analog:** `PROC MEANS NMISS;` or counting missing levels in `PROC FREQ MISSING;`. `ard_missing()` collapses both styles into one consistent ARD shape.

#### 2.5.2 `ard_total_n()` — the denominator row that drives column headers

Look at a demographics table header:

```
                  Placebo (N=86)   Xanomeline Low Dose (N=84)   Xanomeline High Dose (N=84)
```

Where do those `N=86`, `N=84`, `N=84` numbers come from? They are *not* the n of any one variable — they are the total subject count per treatment arm. That number lives in its own row of the ARD, produced by `ard_total_n()`:

```r
ard_total_n(data = adsl, by = TRT01A)
```

Output:

```
  group1_level         | stat_name | stat_fmt
  Placebo              | N         | 86
  Xanomeline Low Dose  | N         | 84
  Xanomeline High Dose | N         | 84
```

The layout engine reads these rows separately and uses them to populate the *column header* of the table — not the body. Without `ard_total_n()`, you have no clean way to display "Placebo (N=86)" because that "86" is not a property of any one analysis variable.

> **SAS analog:** the implicit `_FREQ_` variable that `PROC SUMMARY` produces for each `BY` group, but stored as an explicit, labeled row that survives all the way through to the layout.

#### 2.5.3 The handshake between helper rows and body rows

```
   +----------------------------+
   |   ard_total_n()  --> N=86  |---+
   +----------------------------+   |
                                    +---> drives column header
   +----------------------------+   |
   |   ard_continuous(AGE)      |   |
   |     mean, sd, median ...   |---+---> drives table body
   +----------------------------+   |
                                    |
   +----------------------------+   |
   |   ard_categorical(SEX)     |   |
   |     n, p per level         |---+
   +----------------------------+
```

All three ARDs share the same `group1` / `group1_level` columns (`TRT01A` / `Placebo`...), so a layout engine can join them on those keys and place each value exactly where it belongs.

---

## Exercise 2.4 — Build the Denominator Row

Use `ard_total_n()` to produce the subject counts per treatment arm in `adsl`. These will become the `(N=…)` numbers in the column headers of your demographics table.

```{r ch2-ex4, exercise=TRUE, exercise.lines=8}
# Generate total N per TRT01A
# Hint: ard_total_n() needs only data and by.

ard_total_n(
  data = adsl,
  by = ___
)
```

```{r ch2-ex4-hint-1}
ard_total_n(data = adsl, by = TRT01A)
```

```{r ch2-ex4-solution}
ard_total_n(data = adsl, by = TRT01A)
```

```{r ch2-ex4-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "N" %in% .result$stat_name, "Correct! These N rows will appear in the column headers of any treatment-arm table you build from this dataset."),
  fail_if(~ TRUE, "Pass TRT01A to the by argument of ard_total_n().")
)
```

---

### 2.6 The `by` Argument — Single, Multiple, and Zero Stratification

`by` is the unifying argument across the whole `{cards}` family. It has three useful shapes.

| Shape                              | Meaning                                    | Effect on the ARD                        |
|------------------------------------|--------------------------------------------|------------------------------------------|
| `by = TRT01A`                      | Stratify by one variable (treatment arm)   | Rows split by `group1_level`             |
| `by = c(TRT01A, AVISIT)`           | Two-level stratification (arm × visit)     | Adds `group2`, `group2_level` columns    |
| omitted (no `by`)                  | Overall summary, no stratification         | Single block of rows, no `group1` split  |

A common pattern is to call the same function twice — once with `by = TRT01A` for the per-arm columns and once without `by` for a "Total" column — then `bind_rows()` the two ARDs. That is how a "Treated / Placebo / Total" four-column table is typically assembled.

> **SAS analog:** a `BY` statement with one variable vs. multiple variables vs. omitted. The mental model is identical.

---

### 2.7 Summarizing Multiple Variables in One Call

You almost never want to call `ard_continuous()` separately for AGE, then BMI, then HEIGHT. Pass them all at once:

```r
ard_continuous(
  data      = adsl,
  variables = c(AGE, BMI, HEIGHTBL),
  by        = TRT01A
)
```

The resulting ARD has three blocks of rows — one for each variable — already in the standard schema and ready to stack with categorical, dichotomous, missing, and total-N ARDs into one big input for a layout engine. That single call is the `{cards}` answer to writing three `PROC MEANS` statements in a row.

```
  variable | group1_level | stat_name | stat_fmt
  AGE      | Placebo      | mean      | 75.2
  AGE      | Placebo      | sd        | 8.12
  AGE      | Xan High     | mean      | 74.4
  ...
  BMI      | Placebo      | mean      | 26.3
  BMI      | Placebo      | sd        | 4.20
  ...
  HEIGHTBL | Placebo      | mean      | 167.1
  ...
```

The same pattern works for `ard_categorical(variables = c(SEX, RACE, AGEGR1))`.

---

### 2.8 Choosing Which Statistics Get Computed

The `statistic` argument of `ard_continuous()` lets you trim or extend the default list. Two common patterns:

```r
# Only n, mean, sd
ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = ~ list("N", "mean", "sd")
)

# Default list plus the 95% quantile
ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = ~ list("N", "mean", "sd", "median",
                     "p25", "p75", "min", "max",
                     p95 = ~ stats::quantile(.x, 0.95, na.rm = TRUE))
)
```

A named entry in the list (here `p95 = ~ ...`) lets you plug in *any* R function. The result still lands in `stat` (raw) and `stat_fmt` (formatted) — the ARD shape never changes, only the rows present.

> **SAS analog:** the keyword list after a `PROC MEANS` statement (`PROC MEANS N MEAN STD P95;`). The `{cards}` `statistic` list is the same idea, with the bonus that any custom R function can be added without leaving the API.

---

## Exercise 2.5 — Multiple Variables in One Call

Summarize **two** continuous variables (`AGE` and `BMIBL`) in `adsl`, both grouped by `TRT01A`, in a single `ard_continuous()` call. Look at the output and observe that the ARD already contains *both* variables stacked into the same standard shape.

```{r ch2-ex5, exercise=TRUE, exercise.lines=10}
# Hint: use c(AGE, BMIBL) in the variables argument.

ard_continuous(
  data = adsl,
  variables = c(___, ___),
  by = TRT01A
)
```

```{r ch2-ex5-hint-1}
ard_continuous(data = adsl, variables = c(AGE, BMIBL), by = TRT01A)
```

```{r ch2-ex5-solution}
ard_continuous(data = adsl, variables = c(AGE, BMIBL), by = TRT01A)
```

```{r ch2-ex5-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("AGE", "BMIBL") %in% .result$variable), "Nice! Both variables now share the same ARD. This is the pattern you will use to feed multi-row sections of a demographics table."),
  fail_if(~ TRUE, "Pass both AGE and BMIBL to variables as c(AGE, BMIBL).")
)
```

> **Inspect the `variable` column:** it now contains both `"AGE"` and `"BMIBL"`. A layout engine groups rows by `variable` to build section headers ("Age (years)", "BMI (kg/m²)") with their statistic rows underneath — exactly the structure of a real demographics table.

---

### 2.9 The Big Picture — What You Have Built So Far

After this chapter you can produce, in a few function calls, every row a demographics or disposition table needs:

```
  ard_total_n(adsl, by = TRT01A)                     -> N=86, N=84, N=84    (header)
  ard_continuous(adsl, c(AGE,BMIBL), by=TRT01A)      -> Age, BMI rows       (body)
  ard_categorical(adsl, c(SEX,RACE), by=TRT01A)      -> Sex, Race rows      (body)
  ard_dichotomous(adsl, SAFFL,
                   value=list(SAFFL="Y"), by=TRT01A) -> Safety Pop row      (body)
  ard_missing(adsl, AGE, by=TRT01A)                  -> Missing row         (body)
                              |
                              |  bind_rows()
                              v
                  ONE STACKED ARD, ready for layout
```

What is missing? The mechanics of *stacking* those ARDs and *applying consistent formatting* across them. That is exactly Chapter 3.

---

## Chapter 2 Knowledge Check

```{r ch2-quiz, echo=FALSE}
quiz(
  caption = "Chapter 2 — Check Your Understanding",

  question(
    "Which function should you use to summarize continuous variables like weight or age?",
    answer("ard_categorical()"),
    answer("ard_continuous()", correct = TRUE),
    answer("ard_total_n()"),
    answer("ard_missing()"),
    allow_retry = TRUE
  ),

  question(
    "How does `ard_dichotomous()` differ from `ard_categorical()`?",
    answer("It operates on continuous data"),
    answer("It keeps only the statistic rows corresponding to the specified level (e.g. 'Y')", correct = TRUE),
    answer("It does not calculate percentages"),
    answer("It runs twice as fast"),
    allow_retry = TRUE
  ),

  question(
    "What argument do you use to group calculations by treatment arm in `{cards}` functions?",
    answer("group"),
    answer("split"),
    answer("by", correct = TRUE),
    answer("strata"),
    allow_retry = TRUE
  ),

  question(
    "In `ard_categorical()`, what does the default `denominator = \"column\"` mean?",
    answer("Percentages are computed against the total sample size in the entire dataset."),
    answer("Percentages are computed within each level of the analysis variable."),
    answer("Percentages are computed within each by group (e.g., within each treatment arm).", correct = TRUE),
    answer("Percentages are not computed at all."),
    allow_retry = TRUE
  ),

  question(
    "Which function produces the `(N=86)` numbers shown in the column header of a demographics table?",
    answer("ard_continuous()"),
    answer("ard_categorical()"),
    answer("ard_total_n()", correct = TRUE),
    answer("ard_missing()"),
    allow_retry = TRUE
  ),

  question(
    "A SAS programmer writes `PROC MEANS N MEAN STD;`. What is the equivalent in `ard_continuous()`?",
    answer("denominator = c(\"N\", \"mean\", \"sd\")"),
    answer("statistic = ~ list(\"N\", \"mean\", \"sd\")", correct = TRUE),
    answer("by = c(\"N\", \"mean\", \"sd\")"),
    answer("value = list(\"N\", \"mean\", \"sd\")"),
    allow_retry = TRUE
  ),

  question(
    "You pass `variables = c(AGE, BMIBL)` to `ard_continuous()`. What happens?",
    answer("The function errors because it accepts only one variable at a time."),
    answer("The function produces one block of rows per variable in the same standard ARD schema.", correct = TRUE),
    answer("Only AGE is summarized; BMIBL is silently dropped."),
    answer("AGE and BMIBL are averaged together before summarization."),
    allow_retry = TRUE
  )
)
```

---

## Chapter 2 Summary {.congratulations}

You now know how to:

- Pick the right `{cards}` function for each data type and clinical question.
- Generate continuous summaries with **`ard_continuous()`** — including customizing the list of statistics via the `statistic` argument.
- Compute categorical frequencies with **`ard_categorical()`** — and pick the correct denominator (`"column"`, `"row"`, `"cell"`).
- Extract a single positive flag level with **`ard_dichotomous()`** using the `value` argument.
- Build dedicated missingness rows with **`ard_missing()`**.
- Produce header denominator rows with **`ard_total_n()`** — the numbers that feed `(N=…)` in column headers.
- Group calculations across one or more stratification variables via the **`by`** argument.
- Summarize multiple analysis variables in a single call by passing a vector to `variables`.
- See how every `{cards}` call returns the same standard ARD schema, ready to be stacked with `bind_rows()` into a single dataset that drives a complete table layout.

**Next:** Chapter 3 — Stacking and Formatting (`bind_ard`, `ard_fmt_args`, etc.) — turning the pile of ARDs you just learned to produce into one polished, display-ready dataset.
