---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 7: Exporting and MS Word Styling with flextable + officer"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.2
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build production-grade MS Word (.docx) and RTF deliverables from gtsummary
  tables using flextable for cell-level styling and officer for document layout,
  with every step mapped to its SAS ODS / PROC REPORT counterpart.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(dplyr)
library(cards)
library(cardx)
library(gtsummary)
library(flextable)
library(officer)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

data("adsl", package = "pharmaverseadam")

# Pre-compute an ARD and table for exercises to save time
ard_demog <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A,
  .missing = TRUE
)

base_table <- tbl_ard_summary(
  cards = ard_demog,
  include = c(AGE, SEX),
  by = TRT01A
) %>%
  modify_spanning_header(all_stat_cols() ~ "**Treatment Group**")
```

---

## Chapter 7: Exporting to MS Word

### 7.1 Why MS Word Bridging Matters — The Last Mile of a Clinical Deliverable

In Chapters 1–6 you built a calculation pipeline that is correct, reproducible, and machine-readable. The numbers exist as ARD rows; a layout engine like `{gtsummary}` turned them into a structured table object. So far so good.

But the final deliverable that lands in a Clinical Study Report (CSR), a regulatory submission, or a Data Monitoring Committee (DMC) package is almost never an HTML preview or an in-RStudio render. It is a Microsoft Word `.docx` file (or an `.rtf` that Word can open), with:

- a sponsor-branded header,
- a protocol identifier,
- a sequential table number,
- a footnote block describing the source dataset and run timestamp,
- exact margins, exact paper size, exact orientation (often **landscape** because clinical tables are wide),
- and frequently a Table of Contents entry generated by Word itself.

`{gtsummary}` does not know how to write any of that. It produces a portable in-memory table object. The job of getting that object into a styled, paginated, regulator-grade Word document is split between two packages:

- **`{flextable}`** — controls everything *inside the rectangle of the table*: borders, fonts, colors, column widths, cell merges, conditional formatting, header banding.
- **`{officer}`** — controls everything *outside the table*: page size, margins, orientation, document headers/footers, paragraphs, page breaks, sections.

Together they replace the SAS ODS RTF / ODS WORD pipeline. This chapter walks through that pipeline call-by-call, from the moment a `{gtsummary}` object becomes a `flextable` to the moment `save_as_docx()` writes a regulator-grade file.

```
  THE OLD SAS WORKFLOW                THE R WORKFLOW
  +----------------------+            +----------------------+
  |  PROC REPORT         |            |  gtsummary table     |
  |  + ODS RTF FILE=...  |            |        |             |
  |  + TITLE / FOOTNOTE  |            |  as_flex_table()     |
  |  + ESCAPECHAR        |            |        v             |
  |  + STYLE=            |            |  flextable           |
  +----------+-----------+            |  (cell styling)      |
             |                        |        v             |
             v                        |  officer prop_section|
        .rtf file                     |  (page layout)       |
                                      |        v             |
                                      |  save_as_docx()      |
                                      |        v             |
                                      |  .docx file          |
                                      +----------------------+
```

Every SAS programmer who has wrestled with an ODS RTF call has the same instinct: *"There must be a way to set the font, the margin, the orientation, and the header all at once."* In R, that "way" is the `{flextable}` + `{officer}` pair — and once you learn the vocabulary, you will find it strictly more flexible than `ODS RTF STYLE=`.

---

### 7.2 SAS Programmer's Quick Map — ODS / PROC REPORT vs flextable / officer

Almost every output-styling lever you know in SAS has a direct R counterpart. The two-package split (`{flextable}` for in-table, `{officer}` for the document) maps cleanly onto the SAS split between `PROC REPORT` (table layout) and `ODS RTF` (document container).

| SAS construct                                       | R counterpart                                      | Lives in        | Note                                                                                   |
|-----------------------------------------------------|----------------------------------------------------|-----------------|----------------------------------------------------------------------------------------|
| `ODS RTF FILE = "out.rtf";`                         | `save_as_rtf(my_flex, path = "out.rtf")`           | `{flextable}`   | Write the styled table to an RTF container.                                            |
| `ODS RTF FILE = "out.rtf" STARTPAGE=NO;`            | Multiple `body_add_flextable()` calls in `read_docx()` | `{officer}` | Stack several tables in one document without forced page breaks.                       |
| ODS WORD destination                                | `save_as_docx(my_flex, path = "out.docx")`         | `{flextable}`   | Native `.docx` writer (no Word installation required).                                 |
| `TITLE "Table 14.1.1 ...";`                         | `add_header_lines("Table 14.1.1 ...")`             | `{flextable}`   | Title banded into the table, travels with it.                                          |
| `FOOTNOTE "Generated 2026-06-08";`                  | `add_footer_lines("Generated 2026-06-08")`         | `{flextable}`   | Footnote banded into the table, travels with it.                                       |
| `PROC REPORT COLUMN(...)` order                     | `gtsummary` column order (preserved by `as_flex_table()`) | `{gtsummary}` | The conversion preserves the column order you set upstream.                            |
| `DEFINE varname / DISPLAY WIDTH=15;`                 | `width(my_flex, j = "varname", width = 1.5)`       | `{flextable}`   | Width in inches in R, characters in SAS.                                               |
| `DEFINE varname / JUST=CENTER;`                     | `align(my_flex, j = "varname", align = "center")`  | `{flextable}`   | Per-column alignment.                                                                  |
| `STYLE(HEADER) = { FONT_WEIGHT = bold };`           | `bold(my_flex, part = "header")`                   | `{flextable}`   | Bold all header cells.                                                                 |
| `STYLE(REPORT) = { CELLPADDING=4pt };`              | `padding(my_flex, padding = 4)`                    | `{flextable}`   | Inner cell padding in points.                                                          |
| `STYLE = [BACKGROUND=lightblue];`                    | `bg(my_flex, bg = "#D9E1F2")`                      | `{flextable}`   | Cell background. Use hex codes for sponsor branding.                                   |
| `STYLE = [FOREGROUND=red FONT_WEIGHT=bold];`        | `color(...) %>% bold(...)`                         | `{flextable}`   | Combine with `%>%` for any number of style facets.                                     |
| PROC REPORT `COMPUTE` block changing a cell color   | `bg(my_flex, i = ~ p.value < 0.05, j = "p.value", bg = "yellow")` | `{flextable}` | Conditional row formula via `i = ~ expression`.                                        |
| `OPTIONS PAPERSIZE=LETTER ORIENTATION=LANDSCAPE;`    | `prop_section(page_size = page_size(orient = "landscape"))` | `{officer}` | Page orientation lives at the document section level, not the table.                   |
| `OPTIONS LEFTMARGIN=1in RIGHTMARGIN=1in ...;`        | `page_mar(top = 1, bottom = 1, left = 1, right = 1)` | `{officer}`   | Inches by default.                                                                     |
| `ODS RTF HEADER="Sponsor name";`                     | `headers_default(text = "Sponsor name")` block on a Word template | `{officer}` | Document-level (every page) header, distinct from in-table header.                     |
| `STARTPAGE=NEVER;` then `STARTPAGE=NOW;`            | `body_add_break()`                                 | `{officer}`    | Explicit page break between sections.                                                  |
| `PROC REPORT MERGE`                                  | `merge_v()`, `merge_h()`, `merge_at()`             | `{flextable}`   | Merge adjacent identical cells vertically/horizontally.                                |
| `STYLE = [LINESTYLE=double LINECOLOR=black]`         | `border()`, `border_outer()`, `hline()`, `vline()` | `{flextable}`   | Border control is granular: pick a part, a side, an `i`/`j` selector.                  |
| RTF templates (`.rtf` shells)                        | `read_docx(path = "template.docx")`                | `{officer}`     | Open a reference `.docx`, append tables, save under a new name.                        |

**Mental model:** `{flextable}` is your in-process `PROC REPORT`; `{officer}` is your in-process `ODS RTF`/`ODS WORD`. The two are designed to chain — every `body_add_flextable()` call places a styled `flextable` into an `officer` document.

---

### 7.3 The Three-Package Architecture — gtsummary → flextable → officer

You will use three packages, each owning one well-bounded job. Forgetting the boundaries leads to "I tried to set the margin on the table but nothing happened" confusion — because the margin does not belong to the table; it belongs to the document section.

```
   +-----------+      as_flex_table()      +-------------+
   | gtsummary | ------------------------> |  flextable  |
   |  (cells)  |                           |  (cells +   |
   +-----------+                           |   styling)  |
                                           +------+------+
                                                  |
                                  body_add_flextable() / save_as_docx()
                                                  |
                                                  v
                                           +-------------+
                                           |   officer   |
                                           |  (page,     |
                                           |   margins,  |
                                           |   sections) |
                                           +------+------+
                                                  |
                                                  v
                                           +-------------+
                                           |  .docx /    |
                                           |  .rtf file  |
                                           +-------------+
```

| Package        | Owns                                       | Does NOT own                       |
|----------------|--------------------------------------------|------------------------------------|
| `{gtsummary}`  | which statistic goes in which row/column   | fonts, colors, page size           |
| `{flextable}`  | borders, fonts, colors, widths, alignment, header lines, footer lines, conditional cell formatting | page size, orientation, margins, document headers |
| `{officer}`    | document structure: sections, page size, orientation, margins, document headers/footers, page breaks, paragraphs around tables | the contents of any single table cell |

> **SAS analog:** `PROC TABULATE` / `PROC REPORT` own the row-and-column layout (`{gtsummary}` / `{flextable}`). `OPTIONS PAPERSIZE`, `ODS RTF FILE=`, and the RTF document's header/footer band own the page (`{officer}`). In SAS, a single `ODS` invocation hides the boundary; in R, the boundary is explicit.

---

### 7.4 Anatomy of a flextable Object — the Three Parts

A `flextable` is divided into three named parts, exactly like an `rtables` shell or like the `BODY` / `HEADER` / `FOOTER` regions of a `PROC REPORT`:

```
   +------------------------------------------------+
   |  header   <-- title rows, column headers,      |
   |               spanning headers                 |
   +------------------------------------------------+
   |  body     <-- the data cells (one row per      |
   |               ARD-driven layout row)           |
   +------------------------------------------------+
   |  footer   <-- footnote lines, abbreviations    |
   +------------------------------------------------+
```

Almost every styling function in `{flextable}` accepts a `part = "header" | "body" | "footer" | "all"` argument. This is the equivalent of `STYLE(HEADER)` / `STYLE(REPORT)` / `STYLE(LINES)` in PROC REPORT. When you do not specify `part`, the default is `"body"` — the data region.

| `{flextable}` part | PROC REPORT region            | Contents                                                     |
|--------------------|-------------------------------|--------------------------------------------------------------|
| `"header"`         | `STYLE(HEADER)` / spanning headers | Column titles, spanning bands, sponsor/protocol title lines |
| `"body"`           | `STYLE(REPORT)` data area     | One row per layout row (the "cells")                         |
| `"footer"`         | `STYLE(LINES)` footnote block | Abbreviations, source program, timestamp                     |
| `"all"`            | The whole table               | Apply the style to every part at once                        |

Knowing which `part` you mean is half the battle. The other half is the row/column selector — typically `i = ` (rows) and `j = ` (columns), which we cover in section 7.8.

---

### 7.5 Converting `{gtsummary}` to `{flextable}` — `as_flex_table()`

The single function that bridges `{gtsummary}` to the MS Word ecosystem is `as_flex_table()`. It is the moment your table stops being a pure layout object and becomes a styled cell grid.

```r
flex_tbl <- base_table %>% as_flex_table()
```

What `as_flex_table()` does, step by step:

1. **Walks the `gtsummary` table body**, mapping each row of the layout to a row of `flextable` cells.
2. **Preserves column headers and spanning headers** (the `modify_spanning_header()` / `modify_header()` choices you made upstream are carried over).
3. **Preserves cell-level formatting strings** — the `stat_fmt` values rendered by `apply_fmt_fun()` arrive as strings, unchanged.
4. **Preserves variable labels** — the row labels you set via `modify_table_body()` or via the `label =` argument upstream.
5. **Resets fonts and themes to the `{flextable}` default** (Calibri 11pt). Any color/font you applied via `{gt}` style functions does *not* carry; you re-apply it on the `flextable` side.

> **SAS analog:** think of `as_flex_table()` as the moment an `ODS OUTPUT` dataset becomes a `PROC REPORT` invocation. The numbers exist; now you are choosing how they render.

Once the table is a `flextable`, you can chain any `{flextable}` styling function:

```r
flex_tbl %>%
  bold(part = "header") %>%
  fontsize(size = 9, part = "all") %>%
  autofit()
```

This is exactly the shape of every styling pipeline you will write in this chapter: one `as_flex_table()`, then a chain of `%>%`-piped style verbs, then the `officer` layer.

---

### 7.6 Title Region — `add_header_lines()` (the `TITLE` Statement)

Clinical tables require a multi-line title block: table number, table description, population label. In SAS you would write:

```sas
TITLE1 "Protocol ABC-123";
TITLE2 "Table 14.1.1 Demographics — Safety Population";
```

In `{flextable}` the equivalent is `add_header_lines()`, which prepends a row at the top of the *header* part:

```r
flex_tbl %>%
  add_header_lines("Protocol ABC-123") %>%       # becomes top line
  add_header_lines("Table 14.1.1 Demographics")  # pushed above the protocol line
```

Each call prepends *above* the existing header — calls stack in reverse order. To get the protocol line on top, add it last.

The added lines automatically span the entire width (all columns are merged), which is what you want for a title. They live in `part = "header"`, so any subsequent `bold(part = "header")` will style them too.

> **SAS analog with one subtlety:** in SAS, `TITLE` statements live at the *document* level — they reappear on every page automatically. In `{flextable}`, `add_header_lines()` lives at the *table* level — if your table breaks across pages, the title bands with the table only on the first page. To re-print on every page, use the `officer` document header (section 7.18) instead.

---

### 7.7 Footnote Region — `add_footer_lines()` (the `FOOTNOTE` Statement)

The mirror image. `add_footer_lines()` appends rows at the bottom of the *footer* part:

```r
flex_tbl %>%
  add_footer_lines("Abbreviations: SD, standard deviation; N, number of subjects.") %>%
  add_footer_lines("Source: adsl.sas7bdat. Run: 2026-06-08 14:22 UTC.")
```

The first call goes immediately under the table body; each subsequent call lands *below* the previous footer line. (Opposite stacking order from `add_header_lines()`.)

Typical footnote conventions used in CSRs:

| Footer line content                       | Why it is there                                                  |
|-------------------------------------------|------------------------------------------------------------------|
| `"Note: ..."` / `"Source: ..."`           | The ICH/regulatory expectation that every TFL declares its source. |
| `"Abbreviations: ..."`                    | Every term used in headers/labels is spelled out at least once.   |
| `"[date] [time] UTC"`                     | Reproducibility — pins the file to a calculation moment.          |
| `"Program: tables/t14_1_1_demog.R"`       | Audit trail — names the program that generated this output.       |
| `"sessionInfo() snapshot ..."`            | (Less common) freezes the R/package versions used.                |

> **SAS analog:** `FOOTNOTE1 ... FOOTNOTE10`. Same intent, same idiom; the only difference is that R stacks them by call order while SAS numbers them explicitly.

---

### 7.8 Cell Formatting — Bold, Italic, Color, Background, Font Size

`{flextable}` exposes one style verb per facet. Every one accepts:

- `x` — the flextable (piped in).
- `i` — a row selector (integer indices *or* a formula `~ <expr>` evaluated on the body data).
- `j` — a column selector (integer or column name).
- `part` — `"header"`, `"body"`, `"footer"`, `"all"`.
- the style argument itself (`bold = TRUE`, `color = "red"`, `bg = "#DDDDDD"`, `size = 9`, ...).

| Verb        | Sets                                  | SAS `STYLE=` equivalent             |
|-------------|---------------------------------------|-------------------------------------|
| `bold()`    | font weight                           | `FONT_WEIGHT = bold`                |
| `italic()`  | font style                            | `FONT_STYLE = italic`               |
| `color()`   | text foreground color                 | `FOREGROUND = ...`                  |
| `bg()`      | cell background color                 | `BACKGROUND = ...`                  |
| `fontsize()`| point size                            | `FONT_SIZE = 9pt`                   |
| `font()`    | font family                           | `FONT_FACE = "Arial"`               |
| `padding()` | inner cell padding (points)           | `CELLPADDING = ...`                 |
| `line_spacing()` | line spacing inside cell         | `LINESPACE = ...`                   |

The selector model is the lever PROC REPORT does *not* have natively without writing a `COMPUTE` block. Example: bold every body row whose `p.value` column is less than 0.05.

```r
flex_tbl %>%
  bold(i = ~ p.value < 0.05, j = "p.value", part = "body")
```

That single line is the `{flextable}` answer to a SAS `COMPUTE` block:

```sas
COMPUTE p_value;
  IF p_value < 0.05 THEN CALL DEFINE(_COL_, "STYLE", "STYLE=[FONT_WEIGHT=bold]");
ENDCOMP;
```

Same conditional logic, expressed once as a one-liner.

> **Color values:** any R-recognized color works — `"red"`, `"steelblue"`, or hex codes like `"#1F77B4"`. Sponsor brand colors are almost always provided as hex; paste them in directly.

---

### 7.9 Column Widths — `width()`, `autofit()`, `dim_pretty()`

Clinical tables are dense and wide. Controlling column width is half of getting the table to fit on the page.

`{flextable}` has three width strategies:

1. **`autofit()`** — measures the content of each cell at the current font size and resizes all columns to fit. Quick, often good enough for previews.

   ```r
   flex_tbl %>% autofit()
   ```

   > **SAS analog:** `PROC REPORT NOWD;` with no `WIDTH=` clauses. Lets the engine choose.

2. **`dim_pretty()`** — like `autofit()` but more conservative: returns measured widths that look balanced rather than minimal. Often used as a starting point you then trim by hand.

3. **`width(j = , width = )`** — pin a column to an exact width in inches. The deterministic option.

   ```r
   flex_tbl %>%
     width(j = "label", width = 2.0) %>%   # row-label column
     width(j = 2:5,     width = 1.1)        # four treatment-arm columns
   ```

   > **SAS analog:** `DEFINE label / DISPLAY WIDTH=15;` — except units are inches, not characters. Convert: 1 inch ≈ 10 characters in Calibri 10pt.

A common production pipeline:

```r
flex_tbl %>%
  autofit() %>%
  width(j = "label", width = 2.5) %>%   # pin the label column wider
  width(j = ~ . %in% c("p.value"), width = 0.8)   # squeeze p-value column
```

> **Pixels, inches, and SAS:** when you read existing SAS layout specs, widths in characters/percent convert to `{flextable}` inches via `n_chars * 0.10` or `pct/100 * page_width`. Always pin the *label* column first, then split the remainder among data columns.

---

### 7.10 Alignment — `align()`

PROC REPORT controls alignment with `JUST=LEFT|CENTER|RIGHT` in each `DEFINE`. `{flextable}` uses one verb with a selector:

```r
flex_tbl %>%
  align(j = 1,            align = "left",   part = "all") %>%   # row labels left
  align(j = -1,           align = "center", part = "all") %>%   # everything else centered
  align(j = "p.value",    align = "right",  part = "body")      # numeric column right
```

Negative `j` excludes those columns. The convention in CSR tables is:

| Column kind        | Typical alignment | `align(...)` value   |
|--------------------|-------------------|----------------------|
| Row labels         | Left              | `"left"`             |
| Categorical counts | Center            | `"center"`           |
| Numeric (means, % )| Center or right (sponsor SOP)  | `"center"` / `"right"` |
| p-values           | Right             | `"right"`            |

> **SAS analog:** `DEFINE p_value / DISPLAY JUST=RIGHT;`. One `align()` call per logical column-group is usually enough.

For convenience, `align_text_col(align = "left")` and `align_nottext_col(align = "center")` apply alignment to the text vs. numeric columns automatically — a one-call replacement for typing out every `DEFINE`.

---

### 7.11 Borders and Lines — `border()`, `border_outer()`, `hline()`, `vline()`

Clinical tables follow the *booktabs* convention by default: a thick top rule, a thin rule under the column headers, a thick bottom rule, no vertical lines. `{flextable}` lets you carve that exactly.

| Verb             | Effect                                                       | SAS `STYLE=` analog                                |
|------------------|--------------------------------------------------------------|----------------------------------------------------|
| `border_remove()`| Strip every border                                           | `RULES=NONE FRAME=VOID`                            |
| `border_outer()` | Apply a border on the outside of the chosen part             | `FRAME=BOX` for that part                          |
| `border_inner()` | Apply a border to every internal edge                        | `RULES=ALL`                                        |
| `border_inner_h()` | Internal horizontal lines only                             | `RULES=ROWS`                                       |
| `border_inner_v()` | Internal vertical lines only                               | `RULES=COLS`                                       |
| `hline()`        | A specific horizontal line above/below selected rows         | `STYLE(LINES) = ...` on a specific row             |
| `vline()`        | A specific vertical line between columns                     | `STYLE(LINES) = ...` between columns               |

Border specifications are built with `fp_border()`:

```r
big_line   <- officer::fp_border(color = "black", width = 1.5)
small_line <- officer::fp_border(color = "black", width = 0.5)

flex_tbl %>%
  border_remove() %>%
  hline_top(part    = "header", border = big_line) %>%   # top rule
  hline_bottom(part = "header", border = small_line) %>% # rule under col headers
  hline_bottom(part = "body",   border = big_line)       # bottom rule
```

That is the booktabs look, by hand. Most production teams wrap this into a sponsor theme function (see 7.14) so they never type it twice.

> **SAS analog:** the painstaking nested-`STYLE()` calls inside a custom ODS template. `{flextable}` collapses each rule into a single chained verb.

---

### 7.12 Merging Cells — `merge_v()`, `merge_h()`, `merge_at()`

Demographics tables often need the "Female" / "Male" labels visually merged into a single "Sex" header cell, or repeated row labels collapsed. SAS solves this with `PROC REPORT MERGE` and `ORDER`/`GROUP` definitions; `{flextable}` exposes three verbs:

| Verb         | Merges                                                       | Use case                                                         |
|--------------|--------------------------------------------------------------|------------------------------------------------------------------|
| `merge_v(j)` | Adjacent identical values *vertically* within column(s) `j`  | Collapse repeated section labels in the row-label column         |
| `merge_h(i)` | Adjacent identical values *horizontally* within row(s) `i`   | Collapse a header row to a single spanning cell                  |
| `merge_at(i, j)` | A specific rectangular block of cells                    | Build a custom spanning header (e.g., "Pooled Active") manually  |

```r
flex_tbl %>%
  merge_v(j = "label")     # any consecutive identical labels collapse into one cell
```

> **SAS analog:** `DEFINE label / ORDER;` in `PROC REPORT` collapses consecutive identical values automatically. `merge_v()` is the direct counterpart.

For the spanning headers themselves, the modern `{gtsummary}` workflow handles those upstream via `modify_spanning_header()` — and `as_flex_table()` carries the spans through, so you rarely need to call `merge_at()` by hand.

---

### 7.13 Conditional Cell Styling — the `COMPUTE` Block Equivalent

PROC REPORT's `COMPUTE` block is the SAS workhorse for "color this cell red if X" logic. `{flextable}` collapses that into selector formulas: any style verb accepts `i = ~ <expression>` where the expression is evaluated on the body data with column names as variables.

Three idiomatic patterns you will use over and over:

**A. Highlight a column where a condition holds:**

```r
flex_tbl %>%
  bg(i = ~ p.value < 0.05, j = "p.value", bg = "#FFF2CC", part = "body")
```

**B. Bold an entire row that matches a label condition:**

```r
flex_tbl %>%
  bold(i = ~ label == "Mean (SD)", part = "body")
```

**C. Color a numeric column by value sign (e.g., difference vs. baseline):**

```r
flex_tbl %>%
  color(i = ~ as.numeric(diff) < 0, j = "diff", color = "red") %>%
  color(i = ~ as.numeric(diff) > 0, j = "diff", color = "darkgreen")
```

The mental conversion from SAS is:

```
   SAS COMPUTE block ... CALL DEFINE(_COL_, "STYLE", "STYLE=[BACKGROUND=lightyellow]");
                       |
                       v
   {flextable}:  bg(i = ~ <same condition>, j = "<col>", bg = "lightyellow")
```

> **Why this matters for QC:** because the conditional is a one-liner, you can re-read the table-generation script and immediately *see* which rules apply. With `COMPUTE` blocks scattered through a SAS macro, the same check requires hunting through the macro body.

---

### 7.14 Themes — One-call Look-and-Feel

Once you have written the same color/border/spacing chain three times, freeze it into a theme function:

```r
theme_sponsor <- function(x) {
  x %>%
    border_remove() %>%
    hline_top(part    = "header", border = officer::fp_border(width = 1.5)) %>%
    hline_bottom(part = "header", border = officer::fp_border(width = 0.5)) %>%
    hline_bottom(part = "body",   border = officer::fp_border(width = 1.5)) %>%
    bold(part         = "header") %>%
    fontsize(size = 9, part = "all") %>%
    padding(padding = 3, part = "all") %>%
    align_text_col(align    = "left") %>%
    align_nottext_col(align = "center")
}

flex_tbl %>% theme_sponsor()
```

`{flextable}` also ships a few pre-built themes you can use as starting points:

| Theme                 | Look                                              | SAS template analog               |
|-----------------------|---------------------------------------------------|-----------------------------------|
| `theme_vanilla()`     | Black/white booktabs, no shading                  | RTF default                       |
| `theme_booktabs()`    | Classic booktabs (matches LaTeX `\toprule`)       | Custom ODS template               |
| `theme_box()`         | Every cell bordered                               | `FRAME=BOX RULES=ALL`             |
| `theme_zebra()`       | Alternating row banding                          | `ROWODD=`/`ROWEVEN=` style banding |
| `theme_alafoli()`     | Minimal, no borders                              | `RULES=NONE FRAME=VOID`           |

Apply them like any style verb:

```r
flex_tbl %>% theme_booktabs()
```

> **SAS analog:** the custom `PROC TEMPLATE STYLE=...` you wrote once and then included via `STYLE=...` in every PROC REPORT. A theme function is exactly that, in 5 lines of R.

---

### 7.15 The `{officer}` Engine — Document Skeletons

So far, every verb has been about the rectangle of the table. `{officer}` is what wraps that rectangle in a `.docx` file with a page, a margin, a header band, and (optionally) other paragraphs.

Two usage patterns:

**Pattern A — single-table shortcut:** `save_as_docx()` accepts one or more flextables plus a `prop_section`. It builds the document for you. Use this when the deliverable is one table per file.

**Pattern B — explicit document construction:** start with `read_docx()` (empty or from a template), build up paragraphs and tables with `body_add_*()`, then `print(target = "out.docx")`. Use this for multi-table reports, TOC generation, or sponsor-templated outputs.

```
   Pattern A:  flex --(prop_section)--> save_as_docx() --> file
   Pattern B:  read_docx() ─► body_add_par() ─► body_add_flextable() ─► body_add_break()
                          └─► body_add_flextable() ─► body_set_default_section()
                                                                |
                                                                v
                                                          print(target = ...)
```

> **SAS analog:** Pattern A is `ODS RTF FILE = "out.rtf"; PROC REPORT ...; ODS RTF CLOSE;` — one-call wonder. Pattern B is the multi-PROC, multi-`ODS RTF TEXT` block you write when one file holds many tables and section breaks.

---

### 7.16 Section Properties — `prop_section()`, Page Size, Orientation

A Word document is partitioned into *sections*. Each section has its own page size, orientation, and margins. A `prop_section` object is the in-R representation of those properties.

```r
my_section <- prop_section(
  page_size    = page_size(orient = "landscape", width = 11, height = 8.5),
  page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1),
  type         = "continuous"
)
```

| Argument        | Controls                                                    | SAS `OPTIONS` analog               |
|-----------------|-------------------------------------------------------------|------------------------------------|
| `page_size`     | width, height, orientation                                  | `OPTIONS PAPERSIZE= ORIENTATION=`  |
| `page_margins`  | top/bottom/left/right margins in inches                     | `OPTIONS TOPMARGIN= ...`           |
| `type`          | `"nextPage"`, `"continuous"`, `"oddPage"`, `"evenPage"`     | `STARTPAGE=` behavior              |
| `header_default`/ `footer_default` | document header/footer attached to the section | `ODS RTF HEADER=` / `FOOTER=`      |
| `page_size = page_size(...)`'s `orient`  | `"landscape"` or `"portrait"`            | `ORIENTATION=`                     |

`prop_section` is what you pass into `save_as_docx(pr_section = my_section)`. Without one, `{flextable}` writes a portrait Letter page with default margins — almost never the right choice for a CSR table.

> **Why pin both width and height when you specify `orient = "landscape"`:** because Word's interpretation of "landscape" depends on the locale. Pinning the two dimensions explicitly removes that ambiguity.

> **SAS analog:** `OPTIONS PAPERSIZE=LETTER ORIENTATION=LANDSCAPE LEFTMARGIN=1IN RIGHTMARGIN=1IN TOPMARGIN=1IN BOTTOMMARGIN=1IN;` collapses to one `prop_section()` call in R.

---

### 7.17 Margins — `page_mar()`

Clinical SOPs almost always specify margins to the inch (often 1.0 in on all four sides, sometimes 0.75 in for dense tables). `page_mar()` accepts six arguments — the four sides, plus `header` and `footer` which control how much vertical space the *document header band* (section 7.18) gets.

```r
page_mar(
  top    = 1,
  bottom = 1,
  left   = 1,
  right  = 1,
  header = 0.5,   # distance from page edge to header band
  footer = 0.5,   # distance from page edge to footer band
  gutter = 0
)
```

Units default to inches. To use centimeters, divide accordingly (`2.54 cm = 1 in`).

> **Pitfall the SAS world has too:** if your top margin is 1.0 in and your header band is 0.5 in, your content area starts 1.0 in down — not 1.5 in. The header band sits *within* the margin. Set `header = 0.5` and `top = 1.0` to get the header midway in the upper inch of the page.

---

### 7.18 Document-level Headers and Footers — `headers_default()`

`add_header_lines()` (section 7.6) bands a title into the *table*. But CSR conventions usually require a **document header** that appears on *every page* — sponsor name on the left, page number on the right, etc. That belongs to the `officer` section, not the table.

Set it via a Word **template** with the desired headers already in place:

```r
template <- read_docx(path = "sponsor_template.docx")  # template has the header band

template %>%
  body_add_par("Table 14.1.1 Demographics", style = "heading 1") %>%
  body_add_flextable(flex_tbl) %>%
  print(target = "table_14_1_1.docx")
```

The header (sponsor name, page number, study ID) lives in the template; you do not re-specify it in R. This mirrors the SAS workflow of pointing `ODS RTF` at a pre-built `.rtf` shell.

For a fully programmatic header without a template, build a `block_list`:

```r
hdr <- block_list(
  fpar(ftext("Sponsor: ACME Pharma  |  Protocol ABC-123"),
       fp_p = fp_par(text.align = "left"))
)
sec <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1, header = 0.4),
  header_default = hdr
)
```

> **SAS analog:** the `ODS RTF HEADER="..."` directive (SAS) and the templated `.rtf` shell with a `\header{...}` block (sponsor SOPs) are the two ways SAS programmers do this. R has both options too — programmatic and templated.

---

### 7.19 Exporting — `save_as_docx()`, `save_as_rtf()`, `save_as_pptx()`, `save_as_image()`

The terminal call of the pipeline. `{flextable}` ships writers for the four most common formats; choose by where the deliverable will land.

| Function           | Output                       | Use case                                                |
|--------------------|------------------------------|---------------------------------------------------------|
| `save_as_docx()`   | `.docx` (Word native)        | Modern CSR deliverables, sponsor templates              |
| `save_as_rtf()`    | `.rtf` (Rich Text Format)    | Legacy CSR pipelines, eCTD submissions                  |
| `save_as_pptx()`   | `.pptx` (PowerPoint)         | DMC slides, ad-board presentations                      |
| `save_as_image()`  | `.png` / `.jpg` / `.svg`     | Embedding into HTML reports, Shiny dashboards           |
| `save_as_html()`   | `.html` standalone           | Quick previews to share by email                        |

The two CSR-relevant ones — `save_as_docx()` and `save_as_rtf()` — take the same three arguments:

```r
save_as_docx(my_flex, path = "out.docx", pr_section = my_section)
save_as_rtf(my_flex,  path = "out.rtf")
```

Multiple flextables in one document:

```r
save_as_docx(
  "Demographics" = flex_demog,
  "Disposition"  = flex_disp,
  "Exposure"     = flex_exp,
  path = "tables.docx",
  pr_section = my_section
)
```

Names on the left become the *captions* above each table. Page breaks are inserted automatically between tables. This is the most economical way to ship a small TFL bundle without touching `officer` explicitly.

> **SAS analog:** `ODS RTF FILE = "tables.rtf";` enclosing several `PROC REPORT` blocks. The R version is one function call, named arguments instead of nested PROCs.

---

### 7.20 Multiple Tables per Document & Page Breaks

When the deliverable is a TFL bundle (Table 14.1, Table 14.2, Table 14.3, ...), use the explicit `officer` document pattern:

```r
doc <- read_docx() %>%
  body_add_par("Table 14.1.1 Demographics — Safety Population", style = "heading 1") %>%
  body_add_flextable(flex_demog) %>%
  body_add_break() %>%
  body_add_par("Table 14.1.2 Disposition — Safety Population", style = "heading 1") %>%
  body_add_flextable(flex_disp) %>%
  body_add_break() %>%
  body_add_par("Table 14.2.1 Exposure — Safety Population", style = "heading 1") %>%
  body_add_flextable(flex_exp)

# Pin the section properties for the whole document
doc <- doc %>%
  body_set_default_section(prop_section(
    page_size    = page_size(orient = "landscape"),
    page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
  ))

print(doc, target = "tfl_bundle.docx")
```

Key verbs:

| Verb                          | Effect                                                                      |
|-------------------------------|-----------------------------------------------------------------------------|
| `read_docx()`                 | Start an empty document (or load a `.docx` template).                       |
| `body_add_par(text, style)`   | Add a paragraph. `style` is a Word paragraph style name ("heading 1", "Normal"). |
| `body_add_flextable(flex)`    | Drop a styled flextable into the document.                                  |
| `body_add_break()`            | Force a page break.                                                         |
| `body_add_toc()`              | Insert a Table of Contents placeholder (Word generates the entries).        |
| `body_set_default_section()`  | Pin the section properties (orientation, margins) for the whole doc.        |
| `print(target = "file.docx")` | Serialize the in-memory `officer` document to disk.                         |

> **SAS analog:** `ODS RTF FILE="tfl_bundle.rtf"; PROC REPORT ...; ODS RTF STARTPAGE=NOW; PROC REPORT ...; ODS RTF CLOSE;` and the `ODS PROCLABEL=...` for TOC entries. Pattern B in R is the same idea expressed as `body_add_*()` calls.

---

### 7.21 Using a Word Template (`.dotx` / Reference `.docx`)

Sponsors typically provide a reference `.docx` with the corporate header, footer, fonts, and styles already configured. Point `{officer}` at it as the starting document; your tables inherit everything.

```r
doc <- read_docx(path = "sponsor_reference.docx") %>%
  body_add_par("Table 14.1.1 Demographics", style = "Table Caption") %>%
  body_add_flextable(flex_demog)

print(doc, target = "table_14_1_1.docx")
```

The output `.docx` has the sponsor header, footer, font defaults, and TOC styles from the reference — and your data tables inside. This is the most common pattern in regulated environments because it lets the QA-approved template carry the brand without re-implementing it in R.

> **SAS analog:** pointing `ODS RTF` at a `BODYTITLE` shell or using `STYLE=` to reference a pre-built ODS template. Same idea — separate the brand from the data.

---

### 7.22 The `PROC DOCUMENT` Pattern — Combining Standalone Outputs with a TOC

In SAS, the canonical pipeline for an end-of-study TFL package is:

1. Each table program runs alone, sends its `PROC REPORT` output to an ODS document store: `ODS DOCUMENT NAME=t14_1_1(WRITE); PROC REPORT ...; ODS DOCUMENT CLOSE;`.
2. A separate "assembly" program calls `PROC DOCUMENT` again with the `REPLAY` statement to read each saved item store and stitch them into one combined RTF/PDF.
3. `ODS PROCLABEL` strings and `STARTPAGE=NOW` directives generate the Table of Contents and section breaks.

The R analog uses the same separation of concerns: **render-once, assemble-later**. Two production routes exist, depending on whether the deliverable is `.docx` or `.rtf`:

| SAS step                                            | `{officer}` route (`.docx`)                                     | `{r2rtf}` route (`.rtf`)                                            |
|-----------------------------------------------------|-----------------------------------------------------------------|---------------------------------------------------------------------|
| `ODS DOCUMENT NAME=t14_1_1(WRITE); PROC REPORT ...` | Save each `flextable` as `t14_1_1.rds` via `saveRDS(flex_demog, "t14_1_1.rds")` | Render each table to an individual `.rtf` file with `r2rtf::rtf_encode()` |
| `PROC DOCUMENT NAME=t14_1_1; REPLAY;`               | `readRDS("t14_1_1.rds")` and `body_add_flextable()` it          | Read the individual `.rtf` files and concatenate with `r2rtf::rtf_assemble()` |
| `ODS PROCLABEL="Table 14.1.1 Demographics";`        | `body_add_par("Table 14.1.1 Demographics", style = "heading 1")` followed by `body_add_toc()` | Use the `as_table` argument and `rtf_assemble()`'s combined TOC support |
| `STARTPAGE=NOW;`                                    | `body_add_break()`                                              | Each `.rtf` already paginated; `rtf_assemble()` preserves it       |

**The `{officer}` assembly script** for a fully-TOC'd CSR appendix:

```r
library(officer)
library(flextable)

# Step 1 — each table program has already saved its flextable to disk:
#   saveRDS(flex_demog, "out/t14_1_1_demog.rds")
#   saveRDS(flex_disp,  "out/t14_1_2_disp.rds")
#   saveRDS(flex_ae,    "out/t14_3_1_ae.rds")

# Step 2 — the assembly program walks the directory and stitches them together.
manifest <- tibble::tribble(
  ~caption,                                      ~path,
  "Table 14.1.1 Demographics — Safety Population", "out/t14_1_1_demog.rds",
  "Table 14.1.2 Disposition — Safety Population",  "out/t14_1_2_disp.rds",
  "Table 14.3.1 Adverse Events — Safety Population","out/t14_3_1_ae.rds"
)

doc <- read_docx(path = "sponsor_reference.docx") %>%
  body_add_par("Table of Contents", style = "heading 1") %>%
  body_add_toc(level = 1) %>%
  body_add_break()

for (i in seq_len(nrow(manifest))) {
  flex <- readRDS(manifest$path[i])
  doc  <- doc %>%
    body_add_par(manifest$caption[i], style = "heading 1") %>%   # TOC entry
    body_add_flextable(flex)
  if (i < nrow(manifest)) doc <- doc %>% body_add_break()
}

doc <- doc %>%
  body_set_default_section(prop_section(
    page_size    = page_size(orient = "landscape"),
    page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
  ))

print(doc, target = "csr_appendix.docx")
```

Key verbs new to this section:

| Verb                          | Effect                                                                                  | SAS analog                                  |
|-------------------------------|-----------------------------------------------------------------------------------------|---------------------------------------------|
| `saveRDS(flex, "x.rds")`      | Serialize the styled flextable to a single-file binary store.                            | `ODS DOCUMENT NAME=x(WRITE);`               |
| `readRDS("x.rds")`            | Reload the stored flextable, intact with all styling.                                    | `PROC DOCUMENT NAME=x; REPLAY;`             |
| `body_add_toc(level = 1)`     | Drop a Word TOC field. When Word opens the file, the field auto-populates from heading styles. | `ODS PROCLABEL` + ODS-generated TOC         |
| `body_add_par(..., style = "heading 1")` | Add a heading paragraph that Word indexes into the TOC.                       | `ODS PROCLABEL = "...";`                    |
| `body_set_default_section()`  | Pin the section properties (orientation, margins) for the whole doc.                     | `OPTIONS PAPERSIZE= ORIENTATION= ...;`      |

> **Critical Word behavior:** `body_add_toc()` inserts a *placeholder field*, not the rendered text of the TOC. When the user first opens the file in Word, they will be prompted with "This document contains fields that may refer to other files. Do you want to update the fields?" — clicking *Yes* populates the TOC from the `heading 1` paragraphs you added. This is the same behavior as opening a Word document generated from any TOC-aware template.

> **Why heading styles matter:** the TOC is built from paragraphs whose Word *style name* matches `heading 1`, `heading 2`, etc. Always use `body_add_par(..., style = "heading 1")` for table captions you want in the TOC. Captions added via `add_header_lines()` live inside the flextable and are not indexed by Word's TOC engine.

---

### 7.23 `{r2rtf}` — The Pharma-Native RTF Assembler

`{r2rtf}` is the pharmaverse package built specifically for the regulated-submission RTF pipeline. Many sponsors prefer it over the `flextable` → `save_as_rtf()` route because:

- It writes RTF natively (no intermediate `.docx`).
- It models tables as **data frames + attribute lists**, which is closer to the ADaM-to-TFL mindset than `flextable`'s cell-grid.
- It ships with explicit support for **assembling multiple standalone `.rtf` files into one combined deliverable** — the direct `PROC DOCUMENT REPLAY` equivalent.

The minimum `{r2rtf}` pipeline for a single table:

```r
library(r2rtf)

# Step 1: prepare the data frame that will become the table
df <- ard_demog %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_label, group1_level, stat_fmt) %>%
  tidyr::pivot_wider(names_from = group1_level, values_from = stat_fmt)

# Step 2: chain r2rtf verbs (analogous to PROC REPORT DEFINE statements)
df %>%
  rtf_page(orientation = "landscape", margin = c(1, 1, 1, 1, 0.5, 0.5)) %>%
  rtf_title("Table 14.1.1 Demographics", "Safety Population") %>%
  rtf_colheader(" | Placebo | Xanomeline Low | Xanomeline High",
                col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2),
           text_justification = c("l", "c", "c", "c")) %>%
  rtf_footnote("Source: adsl.sas7bdat. Run: 2026-06-08.") %>%
  rtf_encode() %>%
  write_rtf("out/t14_1_1.rtf")
```

| `{r2rtf}` verb            | Job                                                  | SAS analog                          |
|---------------------------|------------------------------------------------------|-------------------------------------|
| `rtf_page()`              | Page size, orientation, margins                      | `OPTIONS PAPERSIZE= ORIENTATION=`   |
| `rtf_title()`             | Title lines above the table                          | `TITLE1 ... TITLE10`                |
| `rtf_colheader()`         | Column header rows with relative widths              | `DEFINE varname / DISPLAY WIDTH=`   |
| `rtf_body()`              | Body rendering options (alignment, widths)           | `DEFINE / DISPLAY JUST= WIDTH=`     |
| `rtf_footnote()`          | Footnote rows below the table                        | `FOOTNOTE1 ... FOOTNOTE10`          |
| `rtf_subline()`           | A line above the title block (often population name) | A second-tier `TITLE` line          |
| `rtf_source()`            | A source-program line at the very bottom             | The "Source:" `FOOTNOTE`            |
| `rtf_encode()`            | Build the RTF text in memory                         | `ODS RTF` text generation           |
| `write_rtf()`             | Write the RTF text to a file                         | `ODS RTF CLOSE;`                    |

**The `PROC DOCUMENT REPLAY` equivalent — `rtf_assemble()`** (and `assemble_rtf()` in older releases) takes several already-written `.rtf` files and concatenates them into a single, paginated, TOC-aware combined deliverable:

```r
# Each table program produced its own .rtf:
#   out/t14_1_1.rtf, out/t14_1_2.rtf, out/t14_3_1.rtf

assemble_rtf(
  input_files = c("out/t14_1_1.rtf",
                  "out/t14_1_2.rtf",
                  "out/t14_3_1.rtf"),
  output_file = "out/csr_appendix.rtf"
)
```

This is the most direct `PROC DOCUMENT NAME=...; REPLAY;` translation in the R ecosystem. The combined `.rtf` opens in Word with all tables paginated correctly, titles preserved, and footnotes intact — exactly the SAS workflow's behavior.

| Job                                                      | SAS                                          | R (`{r2rtf}`)                         |
|----------------------------------------------------------|----------------------------------------------|---------------------------------------|
| Write a single table to its own RTF store                | `ODS DOCUMENT NAME=t14_1_1(WRITE);`          | `rtf_encode() %>% write_rtf("x.rtf")` |
| Reopen the store later for assembly                      | `PROC DOCUMENT NAME=t14_1_1;`                | Just point at the `.rtf` path         |
| Combine several stored outputs into one delivery file    | `REPLAY;` across multiple `NAME=` arguments  | `assemble_rtf(input_files = c(...))`  |
| Page breaks between sections                             | `STARTPAGE=NOW;`                             | Automatic between assembled files     |
| Auto-generated Table of Contents                         | `ODS PROCLABEL` + ODS-built TOC              | `as_table = TRUE` in `assemble_rtf()` or post-process TOC in Word |

> **When to choose `{r2rtf}` over `{flextable}` + `{officer}`:** `{r2rtf}` is the pharmaverse-blessed choice for regulated-submission RTF and is widely used at large sponsors precisely because of `assemble_rtf()`. Choose `{flextable}` + `{officer}` when the deliverable is native `.docx`, or when you need rich conditional cell styling. Many shops use both — `{flextable}` for one-off `.docx`, `{r2rtf}` for the formal CSR `.rtf` appendix.

> **`{r2rtf}` is not loaded in this tutorial's setup chunk** because it requires a separate install. The code above is shown for reference; install with `install.packages("r2rtf")` to run it in a regular R session.

---

### 7.24 `{officedown}` — R Markdown Driving `{officer}`

For sponsors who want the **full report** (narrative + tables + figures) generated from a single source file, the gap between R Markdown's HTML/PDF output and `{officer}`'s `.docx` engine is bridged by **`{officedown}`**. It is essentially "R Markdown that compiles to `{officer}` documents."

```yaml
---
output:
  officedown::rdocx_document:
    reference_docx: sponsor_reference.docx
    tables:
      style: Table
      layout: autofit
    plots:
      style: Normal
      caption:
        style: Image Caption
        pre: 'Figure '
        sep: ': '
---
```

Then in the body of the `.Rmd`:

````markdown
# Demographics

```r
flex_demog   # a flextable object — officedown drops it in with styling intact
```

\newpage

# Disposition

```r
flex_disp
```
````

`{officedown}` features that matter for the PROC DOCUMENT pattern:

| Feature                    | Effect                                                                            | SAS analog                          |
|----------------------------|-----------------------------------------------------------------------------------|-------------------------------------|
| `reference_docx:` YAML key | Use a sponsor template for fonts/headers/styles.                                  | RTF template file                   |
| `\newpage` in body         | Force a page break between sections.                                              | `STARTPAGE=NOW;`                    |
| `# Heading` Markdown       | Heading style → indexed by Word's TOC engine.                                     | `ODS PROCLABEL`                     |
| `block_toc()`              | Insert a Table of Contents placeholder field.                                     | ODS-generated TOC                   |
| `block_section()`          | Set section properties (landscape, margins) mid-document.                         | `OPTIONS PAPERSIZE= ORIENTATION=`   |
| `block_caption()`          | Cross-referencable table/figure captions.                                         | `ODS PROCLABEL` + numbered captions |

A typical use: one `.Rmd` per study report, knitted at the end of the analysis pipeline, that walks through every TFL in order and produces the combined `.docx` with a TOC — the closest R equivalent to a SAS macro that wraps `PROC DOCUMENT REPLAY` around every output.

> **When to reach for `{officedown}`:** when the deliverable is more than just a table bundle — for example, an analysis narrative with embedded tables and figures, a DMC report, or an integrated summary of safety document where prose paragraphs alternate with statistical tables. For a pure table-only CSR appendix, `{officer}` (section 7.22) or `{r2rtf}` (section 7.23) is leaner.

> **`{officedown}` is not loaded in this tutorial's setup chunk.** Install with `install.packages("officedown")` and `install.packages("rmarkdown")` to use it.

---

### 7.25 The Three Routes Compared — Pick by Deliverable

| Need                                                                 | Best tool                              |
|----------------------------------------------------------------------|----------------------------------------|
| One `.docx` with a few highly styled tables                          | `{flextable}` + `{officer}`            |
| A TFL appendix in `.rtf` matching pharmaverse conventions            | `{r2rtf}` + `assemble_rtf()`           |
| A combined `.docx` from many standalone tables, with auto-TOC        | `{officer}` `read_docx()` + `body_add_toc()` + per-table `readRDS()` |
| A full narrative report (prose + tables + figures) from one source   | `{officedown}` driving `{officer}`     |

The first three routes share a common discipline: **render-once-assemble-later**. Save each table to disk (`.rds` for `{flextable}`, `.rtf` for `{r2rtf}`) immediately after its calculation program runs. The assembly program then reads the saved artifacts and stitches them into the combined deliverable — exactly the SAS `ODS DOCUMENT NAME=...(WRITE);` then `PROC DOCUMENT NAME=...; REPLAY;` pattern, translated into R idiom.

---

### 7.26 The Full Pipeline — One Diagram, One Call Chain

Putting all of the above together, the canonical clinical-table export looks like this:

```
  ARD (from {cards})
        |
        v
  gtsummary table  (rows/cols + spanning headers)
        |
        |  as_flex_table()
        v
  flextable                +---------------------------+
        |                  |  styling verbs:           |
        |                  |  bold/italic/color/bg     |
        |                  |  width/align              |
        |                  |  borders / merges         |
        |  %>% theme_*     |  add_header_lines(...)    |
        |                  |  add_footer_lines(...)    |
        v                  +---------------------------+
  styled flextable
        |
        |  prop_section(page_size, page_margins)   <-- officer
        v
  save_as_docx(... pr_section = my_section)
        |
        v
  .docx file ready for the CSR appendix
```

And the corresponding R code, end to end:

```r
flex_final <- base_table %>%
  as_flex_table() %>%
  add_header_lines("Protocol ABC-123") %>%
  add_header_lines("Table 14.1.1 Demographics — Safety Population") %>%
  add_footer_lines("Source: adsl.sas7bdat. Run: 2026-06-08.") %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  fontsize(size = 9, part = "all") %>%
  width(j = "label", width = 2.5) %>%
  align(j = 1, align = "left", part = "all") %>%
  align(j = -1, align = "center", part = "all")

my_section <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
)

save_as_docx(
  flex_final,
  path       = "table_14_1_1.docx",
  pr_section = my_section
)
```

That single call chain is the R equivalent of a 60-line `PROC REPORT` block wrapped in `ODS RTF FILE=...;` with multiple `STYLE=` and `OPTIONS` statements. The remaining exercises walk you through assembling it one verb at a time.

---

## Exercise 7.1 — Convert and Add Title/Footnote Bands

Convert the existing `base_table` into a `flextable` and inject a sponsor protocol header and a timestamp footer. This is the `TITLE` / `FOOTNOTE` equivalent embedded in the table itself.

```{r ch7-ex1, exercise=TRUE, exercise.lines=18}
# Step 1: Convert the gtsummary object to flextable
flex_tbl <- base_table %>%
  ___()

# Step 2: Add header and footer lines
# Hint: add_header_lines("Protocol: ABC-123")
# Hint: add_footer_lines("Generated 2026-06-08 14:22 UTC")

flex_tbl_styled <- flex_tbl %>%
  ___ %>%
  ___

flex_tbl_styled
```

```{r ch7-ex1-hint-1}
# flex_tbl <- base_table %>% as_flex_table()
# flex_tbl_styled <- flex_tbl %>%
#   add_header_lines("Protocol: ABC-123") %>%
#   add_footer_lines("Generated 2026-06-08 14:22 UTC")
```

```{r ch7-ex1-solution}
flex_tbl <- base_table %>%
  as_flex_table()

flex_tbl_styled <- flex_tbl %>%
  add_header_lines("Protocol: ABC-123") %>%
  add_footer_lines("Generated 2026-06-08 14:22 UTC")

flex_tbl_styled
```

```{r ch7-ex1-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Correct! The protocol line bands into the header (above column titles) and the timestamp bands into the footer (below the data) — exactly what a SAS TITLE/FOOTNOTE pair does."),
  fail_if(~ TRUE, "Use as_flex_table(), then add_header_lines() and add_footer_lines() in sequence.")
)
```

> **SAS reading:** the rendered table is the equivalent of a `PROC REPORT` block wrapped in `TITLE1 "Protocol: ABC-123"; FOOTNOTE1 "Generated 2026-06-08 14:22 UTC";` — but the title and footnote travel *with* the table object, not at the session level.

---

## Exercise 7.2 — Style the Header Row and Resize Fonts

Now apply per-part styling: bold every cell in the header part and shrink the body font to 9pt. This is the `STYLE(HEADER) = { FONT_WEIGHT = bold } STYLE(REPORT) = { FONT_SIZE = 9pt }` of SAS in two function calls.

```{r ch7-ex2, exercise=TRUE, exercise.lines=18}
flex_tbl <- base_table %>% as_flex_table()

# Bold the header part and set font size to 9 across the whole table.
# Hint: bold(part = "header") and fontsize(size = 9, part = "all")

flex_tbl_styled <- flex_tbl %>%
  ___ %>%
  ___

flex_tbl_styled
```

```{r ch7-ex2-hint-1}
# flex_tbl_styled <- flex_tbl %>%
#   bold(part = "header") %>%
#   fontsize(size = 9, part = "all")
```

```{r ch7-ex2-solution}
flex_tbl <- base_table %>% as_flex_table()

flex_tbl_styled <- flex_tbl %>%
  bold(part = "header") %>%
  fontsize(size = 9, part = "all")

flex_tbl_styled
```

```{r ch7-ex2-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Good. The two verbs chain the same way every other style verb does — pick the part, pick the property, move on."),
  fail_if(~ TRUE, "Combine bold(part = 'header') and fontsize(size = 9, part = 'all').")
)
```

> **SAS analog reminder:** in PROC REPORT you would write `STYLE(HEADER) = { FONT_WEIGHT = bold } STYLE(REPORT) = { FONT_SIZE = 9pt }`. One inline ODS phrase per facet — the `{flextable}` verbs are the same idea, one verb per facet, chained with `%>%`.

---

## Exercise 7.3 — Column Width and Alignment

Pin the row-label column to 2.5 inches, left-align it, and center-align everything else. This is the per-column `DEFINE` block of `PROC REPORT` in two verbs.

```{r ch7-ex3, exercise=TRUE, exercise.lines=18}
flex_tbl <- base_table %>% as_flex_table()

# Step 1: Pin the row-label column (j = 1) to width 2.5
# Step 2: Left-align column 1, center-align columns 2 through the end
# Hint: width(j = 1, width = 2.5)
# Hint: align(j = 1, align = "left", part = "all")
# Hint: align(j = -1, align = "center", part = "all")

flex_tbl_styled <- flex_tbl %>%
  ___ %>%
  ___ %>%
  ___

flex_tbl_styled
```

```{r ch7-ex3-hint-1}
# flex_tbl_styled <- flex_tbl %>%
#   width(j = 1, width = 2.5) %>%
#   align(j = 1,  align = "left",   part = "all") %>%
#   align(j = -1, align = "center", part = "all")
```

```{r ch7-ex3-solution}
flex_tbl <- base_table %>% as_flex_table()

flex_tbl_styled <- flex_tbl %>%
  width(j = 1, width = 2.5) %>%
  align(j = 1,  align = "left",   part = "all") %>%
  align(j = -1, align = "center", part = "all")

flex_tbl_styled
```

```{r ch7-ex3-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Excellent. The label column is wide and left-justified; data columns are centered — the convention used in every CSR demographics table."),
  fail_if(~ TRUE, "Chain width(j = 1, width = 2.5) and the two align() calls.")
)
```

> **SAS analog:** `DEFINE label / DISPLAY WIDTH=25 JUST=LEFT; DEFINE _ALL_OTHER_ / DISPLAY JUST=CENTER;` — but the negative-`j` selector lets you do "all other columns" in a single call instead of one `DEFINE` per column.

---

## Exercise 7.4 — Apply a Booktabs Theme

Apply `theme_booktabs()` to get the classic clinical-table look: thick top rule, thin rule under the column headers, thick bottom rule, no vertical lines.

```{r ch7-ex4, exercise=TRUE, exercise.lines=15}
flex_tbl <- base_table %>% as_flex_table()

# Apply the booktabs theme. Hint: theme_booktabs()

flex_tbl_styled <- flex_tbl %>%
  ___

flex_tbl_styled
```

```{r ch7-ex4-hint-1}
# flex_tbl_styled <- flex_tbl %>% theme_booktabs()
```

```{r ch7-ex4-solution}
flex_tbl <- base_table %>% as_flex_table()
flex_tbl_styled <- flex_tbl %>% theme_booktabs()
flex_tbl_styled
```

```{r ch7-ex4-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Perfect. theme_booktabs() applies the canonical thick-thin-thick rule pattern in a single call — what you would otherwise build by hand with hline_top() / hline_bottom()."),
  fail_if(~ TRUE, "Call theme_booktabs() on the flextable.")
)
```

> **SAS analog:** a custom `PROC TEMPLATE STYLE=...` definition baked once into your sitewide style and then included via `STYLE = sponsor_style;`. The theme function is the same idea — encapsulated look-and-feel.

---

## Exercise 7.5 — Define a Landscape Section with Custom Margins

Define a `prop_section` object that sets landscape orientation and 1-inch margins on all four sides. This is the `OPTIONS PAPERSIZE= ORIENTATION= LEFTMARGIN= ...` block of SAS, expressed as one R function call.

```{r ch7-ex5, exercise=TRUE, exercise.lines=16}
# Define a landscape section property with custom margins
# Hint: prop_section(page_size = page_size(orient = "landscape"),
#                    page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1))

my_section <- ___

my_section
```

```{r ch7-ex5-hint-1}
# my_section <- prop_section(
#   page_size = page_size(orient = "landscape"),
#   page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
# )
```

```{r ch7-ex5-solution}
my_section <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
)

my_section
```

```{r ch7-ex5-check}
grade_result(
  pass_if(~ inherits(.result, "prop_section"), "Correct. You now have a section spec — landscape, 1-inch margins — ready to hand to save_as_docx() via pr_section."),
  fail_if(~ TRUE, "Build prop_section(page_size = page_size(orient = 'landscape'), page_margins = page_mar(1,1,1,1)).")
)
```

> **SAS analog:** one line replaces `OPTIONS PAPERSIZE=LETTER ORIENTATION=LANDSCAPE LEFTMARGIN=1IN RIGHTMARGIN=1IN TOPMARGIN=1IN BOTTOMMARGIN=1IN;`.

---

## Exercise 7.6 — Export the Styled Table to a `.docx` File

Wire the styling and the section together and run the writer. Because this is an interactive web tutorial, the file goes to `tempdir()`; we verify it was created.

```{r ch7-ex6, exercise=TRUE, exercise.lines=20}
my_flex <- as_flex_table(base_table) %>%
  theme_booktabs() %>%
  bold(part = "header")

my_section <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
)

# Step 1: Define the export path
export_path <- file.path(tempdir(), "demographics_table.docx")

# Step 2: Export with save_as_docx()
# Hint: save_as_docx(my_flex, path = export_path, pr_section = my_section)
___

# Step 3: Verify the file exists
file.exists(export_path)
```

```{r ch7-ex6-hint-1}
# save_as_docx(my_flex, path = export_path, pr_section = my_section)
```

```{r ch7-ex6-solution}
my_flex <- as_flex_table(base_table) %>%
  theme_booktabs() %>%
  bold(part = "header")

my_section <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
)

export_path <- file.path(tempdir(), "demographics_table.docx")

save_as_docx(
  my_flex,
  path       = export_path,
  pr_section = my_section
)

file.exists(export_path)
```

```{r ch7-ex6-check}
grade_result(
  pass_if(~ identical(.result, TRUE), "Excellent! You just exported a fully styled, landscape-oriented clinical table to MS Word. Open it in Word and you will see the booktabs rules, bold headers, and 1-inch margins exactly as specified."),
  fail_if(~ TRUE, "Call save_as_docx(my_flex, path = export_path, pr_section = my_section).")
)
```

> **SAS reading:** the file you just wrote is the R equivalent of `ODS RTF FILE = "demographics_table.rtf"; PROC REPORT DATA=demog ...; ODS RTF CLOSE;` — except the format is native `.docx` instead of `.rtf`, and the orientation/margin spec is one R object instead of three `OPTIONS` statements.

---

## Exercise 7.7 — Multi-table Bundle with Page Breaks

Build a single `.docx` containing two named tables (Demographics, Disposition) — the smallest realistic TFL bundle. `save_as_docx()` accepts named flextables; each name becomes the caption, and page breaks are inserted automatically.

```{r ch7-ex7, exercise=TRUE, exercise.lines=22}
# Two flextables — for this exercise we reuse base_table twice as a stand-in.
flex_a <- base_table %>% as_flex_table() %>% theme_booktabs()
flex_b <- base_table %>% as_flex_table() %>% theme_booktabs()

my_section <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
)

export_path <- file.path(tempdir(), "tfl_bundle.docx")

# Export both tables with named captions and a landscape section.
# Hint: save_as_docx("Demographics" = flex_a, "Disposition" = flex_b,
#                   path = export_path, pr_section = my_section)
___

file.exists(export_path)
```

```{r ch7-ex7-hint-1}
# save_as_docx(
#   "Demographics" = flex_a,
#   "Disposition"  = flex_b,
#   path           = export_path,
#   pr_section     = my_section
# )
```

```{r ch7-ex7-solution}
flex_a <- base_table %>% as_flex_table() %>% theme_booktabs()
flex_b <- base_table %>% as_flex_table() %>% theme_booktabs()

my_section <- prop_section(
  page_size    = page_size(orient = "landscape"),
  page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
)

export_path <- file.path(tempdir(), "tfl_bundle.docx")

save_as_docx(
  "Demographics" = flex_a,
  "Disposition"  = flex_b,
  path           = export_path,
  pr_section     = my_section
)

file.exists(export_path)
```

```{r ch7-ex7-check}
grade_result(
  pass_if(~ identical(.result, TRUE), "Done — one .docx, two captioned tables, a forced page break between them, and one shared section spec. This is the smallest realistic TFL bundle pattern."),
  fail_if(~ TRUE, "Call save_as_docx() with two named arguments and pr_section = my_section.")
)
```

> **SAS reading:** this is `ODS RTF FILE = "tfl_bundle.rtf"; PROC REPORT ...; ODS RTF STARTPAGE=NOW; PROC REPORT ...; ODS RTF CLOSE;` — three SAS statements, one R function call.

---

## Exercise 7.8 — Conditional Cell Styling (COMPUTE Block Equivalent)

Use a row-selector formula to highlight specific rows. In this exercise, *italicize* every body row whose label is `"Mean (SD)"`. This is the conditional `CALL DEFINE(_COL_, "STYLE", ...)` pattern from a SAS `COMPUTE` block.

```{r ch7-ex8, exercise=TRUE, exercise.lines=18}
flex_tbl <- base_table %>% as_flex_table()

# Italicize body rows where the label column equals "Mean (SD)".
# Hint: italic(i = ~ label == "Mean (SD)", part = "body")

flex_tbl_styled <- flex_tbl %>%
  ___

flex_tbl_styled
```

```{r ch7-ex8-hint-1}
# flex_tbl_styled <- flex_tbl %>%
#   italic(i = ~ label == "Mean (SD)", part = "body")
```

```{r ch7-ex8-solution}
flex_tbl <- base_table %>% as_flex_table()

flex_tbl_styled <- flex_tbl %>%
  italic(i = ~ label == "Mean (SD)", part = "body")

flex_tbl_styled
```

```{r ch7-ex8-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Nice — that one-liner is the R equivalent of an entire SAS COMPUTE block. The body data is treated as a data frame; the formula evaluates per row; matching rows get the italic style."),
  fail_if(~ TRUE, "Use italic(i = ~ label == 'Mean (SD)', part = 'body').")
)
```

> **SAS analog:** the equivalent SAS is `COMPUTE label; IF label = "Mean (SD)" THEN CALL DEFINE(_ROW_, "STYLE", "STYLE=[FONT_STYLE=italic]"); ENDCOMP;` — five lines vs. one.

---

## Exercise 7.9 — Build a Combined `.docx` with a Table of Contents (PROC DOCUMENT REPLAY equivalent)

Build a single `.docx` that combines two tables, each preceded by a `heading 1` caption, with a TOC at the top. This is the R equivalent of `PROC DOCUMENT NAME=...; REPLAY;` with `ODS PROCLABEL=...;` directives.

Steps:

1. Create two styled flextables (we reuse `base_table` twice as a stand-in for two distinct outputs).
2. Open an empty `officer` document with `read_docx()`.
3. Add a "Table of Contents" heading and a TOC field with `body_add_toc()`.
4. For each table, add a `heading 1` caption (so it shows up in the TOC) and then the flextable.
5. Pin landscape orientation + 1-inch margins via `body_set_default_section()`.
6. Print the document.

```{r ch7-ex9, exercise=TRUE, exercise.lines=30}
flex_a <- base_table %>% as_flex_table() %>% theme_booktabs()
flex_b <- base_table %>% as_flex_table() %>% theme_booktabs()

export_path <- file.path(tempdir(), "csr_appendix.docx")

# Build the document step by step.
# Hint: read_docx() %>%
#         body_add_par("Table of Contents", style = "heading 1") %>%
#         body_add_toc(level = 1) %>%
#         body_add_break() %>%
#         body_add_par("Table 14.1.1 Demographics", style = "heading 1") %>%
#         body_add_flextable(flex_a) %>%
#         body_add_break() %>%
#         body_add_par("Table 14.1.2 Disposition", style = "heading 1") %>%
#         body_add_flextable(flex_b)
# Then body_set_default_section() and print().

doc <- ___

doc <- doc %>%
  body_set_default_section(prop_section(
    page_size    = page_size(orient = "landscape"),
    page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
  ))

print(doc, target = export_path)

file.exists(export_path)
```

```{r ch7-ex9-hint-1}
# doc <- read_docx() %>%
#   body_add_par("Table of Contents", style = "heading 1") %>%
#   body_add_toc(level = 1) %>%
#   body_add_break() %>%
#   body_add_par("Table 14.1.1 Demographics", style = "heading 1") %>%
#   body_add_flextable(flex_a) %>%
#   body_add_break() %>%
#   body_add_par("Table 14.1.2 Disposition", style = "heading 1") %>%
#   body_add_flextable(flex_b)
```

```{r ch7-ex9-solution}
flex_a <- base_table %>% as_flex_table() %>% theme_booktabs()
flex_b <- base_table %>% as_flex_table() %>% theme_booktabs()

export_path <- file.path(tempdir(), "csr_appendix.docx")

doc <- read_docx() %>%
  body_add_par("Table of Contents", style = "heading 1") %>%
  body_add_toc(level = 1) %>%
  body_add_break() %>%
  body_add_par("Table 14.1.1 Demographics", style = "heading 1") %>%
  body_add_flextable(flex_a) %>%
  body_add_break() %>%
  body_add_par("Table 14.1.2 Disposition", style = "heading 1") %>%
  body_add_flextable(flex_b)

doc <- doc %>%
  body_set_default_section(prop_section(
    page_size    = page_size(orient = "landscape"),
    page_margins = page_mar(top = 1, bottom = 1, left = 1, right = 1)
  ))

print(doc, target = export_path)

file.exists(export_path)
```

```{r ch7-ex9-check}
grade_result(
  pass_if(~ identical(.result, TRUE), "Done. The .docx now contains a TOC placeholder, two heading-1 entries that Word will index, and two paginated tables. When the user opens the file in Word, the TOC populates automatically — the direct R equivalent of PROC DOCUMENT REPLAY with ODS PROCLABEL."),
  fail_if(~ TRUE, "Use read_docx() then chain body_add_par(..., style = 'heading 1'), body_add_toc(), body_add_break(), and body_add_flextable() calls.")
)
```

> **What Word does when the user opens the file:** the TOC placeholder is an unpopulated field. Word prompts "Update fields?" — clicking *Yes* walks the document, finds every `heading 1` paragraph, and fills in the TOC with page numbers. This is exactly the behavior you get from SAS's ODS-generated TOC under `PROC DOCUMENT REPLAY`.

> **Render-once, assemble-later in production:** in a real CSR pipeline you would not build `flex_a` and `flex_b` inline. Instead, each table program would end with `saveRDS(flex, "out/t14_1_1.rds")`, and this assembly script would `readRDS()` each one. The discipline matches SAS's `ODS DOCUMENT NAME=...(WRITE);` / `PROC DOCUMENT REPLAY;` separation.

---

## Chapter 7 Knowledge Check

```{r ch7-quiz, echo=FALSE}
quiz(
  caption = "Chapter 7 — Check Your Understanding",

  question(
    "Which package is responsible for defining *document-level* properties such as page size, orientation, and margins?",
    answer("gtsummary"),
    answer("cards"),
    answer("flextable"),
    answer("officer", correct = TRUE),
    allow_retry = TRUE
  ),

  question(
    "Which `{flextable}` function bridges a `{gtsummary}` object into the MS Word ecosystem?",
    answer("as_gt()"),
    answer("as_flex_table()", correct = TRUE),
    answer("save_as_docx()"),
    answer("ard_stack()"),
    allow_retry = TRUE
  ),

  question(
    "What does `add_header_lines(\"Protocol: ABC-123\")` do?",
    answer("Adds a paragraph above the table in the document, but not in the table itself."),
    answer("Prepends a new row at the top of the table's header part, spanning all columns.", correct = TRUE),
    answer("Sets the page header for every page of the document."),
    answer("Adds a footnote."),
    allow_retry = TRUE
  ),

  question(
    "Which `{flextable}` verb is the closest equivalent of a SAS `PROC REPORT` `COMPUTE` block that conditionally styles a cell?",
    answer("`add_header_lines()`"),
    answer("`theme_booktabs()`"),
    answer("Any style verb with `i = ~ <expression>` as a row selector (e.g., `bg(i = ~ p.value < 0.05, j = 'p.value', bg = 'yellow')`).", correct = TRUE),
    answer("`save_as_docx()`"),
    allow_retry = TRUE
  ),

  question(
    "Which argument of `save_as_docx()` carries the landscape/portrait + margin specification?",
    answer("`layout`"),
    answer("`pr_section`", correct = TRUE),
    answer("`section`"),
    answer("`docx_options`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{flextable}` `part` would you target with `bold(part = ...)` to make the column titles bold?",
    answer("`\"body\"`"),
    answer("`\"footer\"`"),
    answer("`\"header\"`", correct = TRUE),
    answer("`\"all\"`"),
    allow_retry = TRUE
  ),

  question(
    "Which SAS construct corresponds most directly to `prop_section(page_size = page_size(orient = 'landscape'), page_margins = page_mar(1,1,1,1))`?",
    answer("`PROC REPORT WIDTH=`"),
    answer("`OPTIONS PAPERSIZE=LETTER ORIENTATION=LANDSCAPE LEFTMARGIN=1IN ...;`", correct = TRUE),
    answer("`TITLE` and `FOOTNOTE` statements"),
    answer("`ODS RTF STARTPAGE=NO;`"),
    allow_retry = TRUE
  ),

  question(
    "Which verb forces a page break between two tables in an `{officer}` document?",
    answer("`body_add_par()`"),
    answer("`body_add_break()`", correct = TRUE),
    answer("`body_add_flextable()`"),
    answer("`body_set_default_section()`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{flextable}` function corresponds to SAS `DEFINE varname / DISPLAY WIDTH=15;`?",
    answer("`width(j = 'varname', width = 1.5)`", correct = TRUE),
    answer("`align(j = 'varname', align = 'left')`"),
    answer("`fontsize(size = 15)`"),
    answer("`autofit()`"),
    allow_retry = TRUE
  ),

  question(
    "When you call `save_as_docx(\"Demographics\" = flex_a, \"Disposition\" = flex_b, path = ..., pr_section = ...)`, what is the effect of the *names* on the left of the `=`?",
    answer("They are ignored."),
    answer("They become the file names of two separate output files."),
    answer("They become captions printed above each table in the same .docx, with an automatic page break between tables.", correct = TRUE),
    answer("They become column headers."),
    allow_retry = TRUE
  ),

  question(
    "Which `{flextable}` theme produces the classic clinical-table look (thick top rule, thin rule under header, thick bottom rule, no vertical lines)?",
    answer("`theme_zebra()`"),
    answer("`theme_box()`"),
    answer("`theme_booktabs()`", correct = TRUE),
    answer("`theme_alafoli()`"),
    allow_retry = TRUE
  ),

  question(
    "Which R verbs together implement the SAS `ODS DOCUMENT NAME=...(WRITE);` then `PROC DOCUMENT NAME=...; REPLAY;` render-then-assemble pattern for `.docx`?",
    answer("`as_flex_table()` and `theme_booktabs()`"),
    answer("`saveRDS()` of the flextable in the table program, then `readRDS()` and `body_add_flextable()` in the assembly program.", correct = TRUE),
    answer("`bind_ard()` and `ard_stack()`"),
    answer("`PROC REPORT` and `ODS RTF`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{officer}` verb inserts the placeholder field that Word will auto-populate as a Table of Contents from `heading 1` paragraphs?",
    answer("`body_add_par()`"),
    answer("`body_add_break()`"),
    answer("`body_add_toc()`", correct = TRUE),
    answer("`body_set_default_section()`"),
    allow_retry = TRUE
  ),

  question(
    "Which pharmaverse package provides `assemble_rtf()` — the most direct R analog of `PROC DOCUMENT REPLAY` for combining multiple standalone `.rtf` files into one deliverable?",
    answer("`{officer}`"),
    answer("`{flextable}`"),
    answer("`{r2rtf}`", correct = TRUE),
    answer("`{officedown}`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{r2rtf}` verb corresponds to SAS `TITLE1 \"Table 14.1.1 ...\";`?",
    answer("`rtf_page()`"),
    answer("`rtf_title()`", correct = TRUE),
    answer("`rtf_footnote()`"),
    answer("`rtf_body()`"),
    allow_retry = TRUE
  ),

  question(
    "Which package would you reach for to compile a full study report (prose paragraphs + embedded tables + figures) from a single `.Rmd` source into a `.docx`, using a sponsor reference template?",
    answer("`{officedown}` (driving `{officer}`)", correct = TRUE),
    answer("`{flextable}` alone"),
    answer("`{cards}`"),
    answer("`{r2rtf}`"),
    allow_retry = TRUE
  ),

  question(
    "In a `body_add_par(text, style = ?)` call, which `style` value causes the paragraph to be indexed by Word's automatic TOC?",
    answer("`\"Normal\"`"),
    answer("`\"Table Caption\"`"),
    answer("`\"heading 1\"` (and `\"heading 2\"`, etc.)", correct = TRUE),
    answer("`\"Body Text\"`"),
    allow_retry = TRUE
  ),

  question(
    "You need a document header (sponsor name + page number) that appears on *every page*. Where does that belong?",
    answer("In `add_header_lines()` on the flextable."),
    answer("In the `officer` section's `header_default`, ideally inherited from a sponsor template `.docx`.", correct = TRUE),
    answer("In `TITLE` lines passed to `as_flex_table()`."),
    answer("In the `pr_section` object's `page_margins` argument."),
    allow_retry = TRUE
  )
)
```

---

## Chapter 7 Summary {.congratulations}

You now know how to:

- Split a clinical Word deliverable into its two natural concerns:
  - **Inside the table** — owned by `{flextable}` (cells, borders, fonts, colors, widths, alignment, conditional styling).
  - **Outside the table** — owned by `{officer}` (page size, orientation, margins, document headers/footers, page breaks).
- Bridge `{gtsummary}` into the Word ecosystem with **`as_flex_table()`**.
- Add CSR-style title and footnote bands directly onto the table with **`add_header_lines()`** and **`add_footer_lines()`** (the in-table `TITLE`/`FOOTNOTE` equivalent).
- Style cells with the per-facet verbs **`bold()`**, **`italic()`**, **`color()`**, **`bg()`**, **`fontsize()`**, **`font()`**, **`padding()`** — each accepting `i`/`j`/`part` selectors that replace SAS `STYLE(...)=` clauses.
- Replicate PROC REPORT `COMPUTE` blocks with one-line row-selector formulas like **`bg(i = ~ p.value < 0.05, j = 'p.value', bg = 'yellow')`**.
- Pin column widths with **`width()`** and align columns with **`align()`** (the `DEFINE varname / WIDTH= JUST=` equivalents).
- Apply pre-built look-and-feel via **`theme_booktabs()`**, **`theme_vanilla()`**, **`theme_zebra()`**, or your own sponsor theme function.
- Define document layout with **`prop_section()`**, **`page_size()`**, **`page_mar()`** — the R version of `OPTIONS PAPERSIZE= ORIENTATION= LEFTMARGIN=...;`.
- Export single tables with **`save_as_docx()`** / **`save_as_rtf()`** and multi-table TFL bundles with named arguments or the explicit `read_docx() %>% body_add_*()` pattern.
- Use a sponsor reference `.docx` template via **`read_docx(path = "template.docx")`** so corporate headers/footers/styles are inherited rather than re-implemented.
- Replicate the SAS `PROC DOCUMENT` / `REPLAY` render-once-assemble-later pattern in two regulator-grade routes:
  - **`.docx` route:** `saveRDS(flex, ...)` per table program, then in the assembly script `read_docx() %>% body_add_toc() %>% body_add_par(..., style = "heading 1") %>% body_add_flextable(readRDS(...))` to produce a TOC-aware combined deliverable.
  - **`.rtf` route:** `{r2rtf}` — render each table to its own `.rtf` with `rtf_page() %>% rtf_title() %>% rtf_colheader() %>% rtf_body() %>% rtf_footnote() %>% rtf_encode() %>% write_rtf()`, then concatenate with **`assemble_rtf()`** — the most direct R analog of `PROC DOCUMENT REPLAY`.
- Recognize **`{officedown}`** as the right tool when the deliverable is a full narrative report (prose + tables + figures) compiled from a single `.Rmd` to `.docx` with sponsor template inheritance.
- Map every SAS construct (`ODS RTF`, `ODS DOCUMENT`, `PROC DOCUMENT REPLAY`, `TITLE`, `FOOTNOTE`, `ODS PROCLABEL`, `PROC REPORT DEFINE`, `STYLE(...)`, `COMPUTE` blocks, `OPTIONS PAPERSIZE/ORIENTATION/MARGIN`, `STARTPAGE`) to its `{flextable}` + `{officer}` + `{r2rtf}` + `{officedown}` counterpart so the mental switch from SAS to R is one of *vocabulary*, not of concepts.

**Next:** Chapter 8 — Advanced Workflows (Writing custom statistics and mapping to CDISC ARS JSON).
