---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 10: Broom vs. Parameters Tidiers"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.2
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Compare the two dominant model-tidier ecosystems in R — broom and parameters —
  understand their column-naming conflict, see why reference rows go missing,
  and learn how broom.helpers and cardx::ard_regression bridge the gap for
  clinical reporting.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(dplyr)
library(cards)
library(cardx)
library(broom)
library(broom.helpers)
library(parameters)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")

# Pre-fit a logistic regression model used across exercises
# Predicting Safety Population inclusion (SAFFL == "Y") from treatment arm and age.
# Placebo is the reference level of TRT01A.
adsl_model_data <- adsl %>%
  filter(!is.na(TRT01A), !is.na(AGE)) %>%
  mutate(
    RESP_BIN = ifelse(SAFFL == "Y", 1L, 0L),
    TRT01A   = factor(TRT01A,
                      levels = c("Placebo",
                                 "Xanomeline Low Dose",
                                 "Xanomeline High Dose"))
  )

model_fit <- glm(
  RESP_BIN ~ TRT01A + AGE,
  data   = adsl_model_data,
  family = binomial
)
```

---

## Chapter 10: Broom vs. Parameters Tidiers

### 10.1 Why This Chapter Exists — The Tidier Crossroads

Chapter 5 introduced the four-layer pipeline that lets `{cards}` turn a fitted model into an ARD: `broom::tidy()` extracts the math, `broom.helpers::tidy_plus_plus()` enriches it, `cardx::ard_regression()` casts it into the ARD schema, and a downstream tool renders the table.

That chapter took `broom.helpers` as a given and showed you the result. This chapter zooms in on the most consequential choice inside that pipeline: **which tidier do you actually run, and why does it matter?**

There are two competing tidier ecosystems in R, and they were built by different people with different goals:

```
            +-----------------------------+        +-----------------------------+
            |   {broom}                   |        |   {parameters} (easystats)  |
            |                             |        |                             |
            |  - oldest                   |        |  - newer                    |
            |  - minimalist               |        |  - reporting-oriented       |
            |  - tidymodels heritage      |        |  - rich metadata            |
            |  - drops reference rows     |        |  - keeps reference rows     |
            |  - uses term, estimate, ... |        |  - uses Parameter, Coefficient, ... |
            +--------------+--------------+        +---------------+-------------+
                           |                                       |
                           |       SAME fitted model object        |
                           +-------------------+-------------------+
                                               |
                                               v
                                  TWO DIFFERENT DATA FRAMES
```

A clinical reporting workflow can survive on either one, but only if you understand the trade-offs. Picking the wrong default leads to two real-world consequences:

- **A missing Placebo row** in your odds-ratio table, because `broom` silently dropped it.
- **A regulator's QC script that cannot find your `estimate` column**, because you accidentally shipped `parameters`-style column names.

The fix is to layer the two ecosystems together through `{broom.helpers}` — a "Rosetta Stone" package that converts between the two vocabularies *and* reinjects the missing reference rows. This chapter walks through every piece of that translation.

> **Note on overlap with Chapter 5:** Chapter 5 shows you the *pipeline*. This chapter is the *comparison drill*. Each section here pairs the two tidiers on identical input and explains every difference column by column.

---

### 10.2 SAS Programmer's Quick Map

If your reflex is to write `ODS OUTPUT ParameterEstimates = est;` after a `PROC LOGISTIC` call, here is the dictionary that maps every part of that workflow into R.

| SAS construct                                                  | R / `{cards}` equivalent                                          | Note                                                               |
|----------------------------------------------------------------|-------------------------------------------------------------------|--------------------------------------------------------------------|
| `PROC LOGISTIC; MODEL y = x;`                                  | `glm(y ~ x, family = binomial, data = ...)`                       | The model fit object holds everything.                              |
| `ODS OUTPUT ParameterEstimates = est;`                         | `broom::tidy(fit)`                                                | One row per estimated coefficient.                                  |
| `ODS OUTPUT FitStatistics = fit;`                              | `broom::glance(fit)`                                              | One row per model — AIC, BIC, deviance.                             |
| `OUTPUT OUT=pred PRED=p RESID=r;`                              | `broom::augment(fit)`                                             | Adds per-observation predicted/residual columns.                    |
| `ODS OUTPUT OddsRatios = or;`                                  | `broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE)`          | Tidy with `exponentiate = TRUE`.                                     |
| Reference class row in PROC LOGISTIC `Class Level Information` | `tidy_plus_plus()` reference-row reinjection                      | SAS prints the reference automatically; `broom::tidy()` does not.   |
| `LABEL TRT01A = "Treatment Arm";`                              | `broom.helpers` reads `label()` attributes                        | Variable labels survive into the tidied frame.                       |
| `PROC FORMAT VALUE` applied for the displayed cell             | `cardx::ard_regression()` + `fmt_fun`                             | Format lives next to raw value (as everywhere in `{cards}`).         |
| Multiple ODS datasets stitched together by hand                | `cardx::ard_regression()` returns one ARD covering everything     | The four ODS outputs become rows of one ARD.                         |
| `OPTIONS NOFMTERR;`                                            | n/a                                                               | No format errors in R — numeric values are not coupled to formats.   |
| Manually adding "Reference" row via `DATA STEP`                | `tidy_plus_plus(add_reference_rows = TRUE)` (default)             | Automated.                                                           |

**Key mental shift:** in SAS, every PROC writes a *fixed* set of ODS datasets with *fixed* column names. In R, *which* function you call to tidy determines both the column names and which rows appear. The "ODS OUTPUT" step is no longer hidden inside the procedure; it is an explicit, swappable function call.

---

### 10.3 The Two Tidier Ecosystems, Side by Side

`{broom}` and `{parameters}` are both wrappers around the same fitted model object. Both ultimately call methods like `coef(fit)`, `vcov(fit)`, and `confint(fit)`. What they differ on is *how* they package the result.

```
                          +------------------+
                          |  fitted model    |
                          |   (glm, lm, ...) |
                          +--------+---------+
                                   |
              +--------------------+---------------------+
              |                                          |
              v                                          v
   +----------------------+                  +-----------------------------+
   | broom::tidy(fit)     |                  | parameters::model_parameters|
   +----------------------+                  +-----------------------------+
   columns:                                  columns:
   - term                                    - Parameter
   - estimate                                - Coefficient
   - std.error                               - SE
   - statistic                               - z (or t)
   - p.value                                 - p
   - conf.low, conf.high                     - CI_low, CI_high
                                             - df_error, Effects, Component
   rows: ONLY estimated coefficients         rows: estimated + reference + random structure
   philosophy: minimalist, programmatic      philosophy: reporting-ready, human-friendly
```

A few practical consequences fall out of the diagram:

- Code written against `broom` will *break* if you feed it the output of `parameters`, because the column names differ. The opposite is also true.
- `parameters` includes a richer set of model-structure rows (random effects, interaction components) that downstream layout tools sometimes do not want.
- `broom` is the dominant choice in the tidymodels universe; `parameters` is the dominant choice in the easystats universe. Both ecosystems are large.

`{cards}` resolves the conflict by *standardizing on the `broom` vocabulary* and using `{broom.helpers}` to fill the gaps that `broom` alone leaves behind (reference rows, variable labels). The reasoning is pragmatic: `broom` shipped first, has wider adoption, and matches the column names that `cardx` internals expect.

> **SAS analog:** picking between `broom` and `parameters` is roughly like picking between two SAS macros that both wrap `PROC LOGISTIC` but produce different output dataset structures. The math is the same; the column dictionary differs.

---

### 10.4 The Column-Naming Dictionary

Memorize this once and you can translate any tidier example from a blog post on the fly. Every cell of the table refers to the *same number* — the difference is the column name.

| Concept                       | `broom::tidy()` column | `parameters::model_parameters()` column | ARD `stat_name` after `ard_regression()` |
|-------------------------------|------------------------|-----------------------------------------|-------------------------------------------|
| Coefficient name              | `term`                 | `Parameter`                             | `term`                                    |
| Point estimate                | `estimate`             | `Coefficient`                           | `estimate`                                |
| Standard error                | `std.error`            | `SE`                                    | `std.error`                               |
| Test statistic (z / t / chi)  | `statistic`            | `z` (logistic) / `t` (linear)           | `statistic`                               |
| p-value                       | `p.value`              | `p`                                     | `p.value`                                 |
| Confidence interval, lower    | `conf.low`             | `CI_low`                                | `conf.low`                                |
| Confidence interval, upper    | `conf.high`            | `CI_high`                               | `conf.high`                               |
| Degrees of freedom            | `df`                   | `df_error`                              | `df`                                      |
| Effect family (fixed/random)  | n/a                    | `Effects`                               | n/a                                       |
| Model component               | n/a                    | `Component`                             | n/a                                       |

The last two rows show the metadata that `parameters` adds but `broom` omits. Random-effect and component columns matter for mixed models (`lmer`, `glmer`) — those models often need the extra metadata, which is why a programmer working with multilevel models may prefer `parameters` upstream of `broom.helpers`.

> **SAS analog:** the `Parameter`, `Estimate`, `StdErr`, `WaldChiSq`, `ProbChiSq` columns inside `PROC LOGISTIC`'s `ParameterEstimates` ODS dataset map to `broom`'s `term`, `estimate`, `std.error`, `statistic`, `p.value`. The mapping is direct.

---

### 10.5 `broom::tidy()` — What You Get and Why a Row Goes Missing

Run `broom::tidy()` on a logistic regression with a three-level categorical predictor (`Placebo`, `Xanomeline Low Dose`, `Xanomeline High Dose`) and you will get **four rows**, not five:

```
   term                            estimate    std.error   statistic   p.value
   ------------------------------  ----------  ----------  ----------  --------
   (Intercept)                       2.31         0.78       2.96       0.003
   TRT01AXanomeline Low Dose         0.04         0.35       0.11       0.913
   TRT01AXanomeline High Dose       -0.18         0.34      -0.53       0.598
   AGE                              -0.01         0.01      -1.20       0.230
```

Notice three things:

- There is no row for `Placebo`. Mathematically, its coefficient is *defined to be zero* (the reference level is folded into the intercept), so `broom::tidy()` does not emit it. From `broom`'s point of view, no number was estimated; no row appears.
- The non-reference levels appear concatenated to the variable name: `TRT01AXanomeline Low Dose`. The factor-level boundary is not parsed out.
- The intercept is included as just another row. SAS would label it `Intercept`; R labels it `(Intercept)` with parentheses.

For a clinical Odds Ratio table you typically *need*:

```
                                   Odds Ratio   95% CI         p-value
   Placebo                            1.00      (Reference)    —
   Xanomeline Low Dose                1.04      (0.52, 2.08)   0.913
   Xanomeline High Dose               0.84      (0.43, 1.62)   0.598
```

The Placebo row is required, the `(Reference)` text is required, and the factor-level cleanup is required. None of that comes out of `broom::tidy()` alone.

> **SAS analog:** `PROC LOGISTIC`'s default output *does* show the reference class in the `Class Level Information` table and reinforces it in the `OddsRatios` table. SAS does this for you; `broom::tidy()` does not.

---

### 10.6 `parameters::model_parameters()` — Reference Rows for Free

The same model fed to `parameters::model_parameters()` produces:

```
   Parameter                       Coefficient   SE      z         p       CI_low   CI_high
   ------------------------------  -----------   ------  --------  ------  -------  -------
   (Intercept)                        2.31        0.78    2.96     0.003     0.78     3.84
   TRT01A [Xanomeline Low Dose]       0.04        0.35    0.11     0.913    -0.65     0.73
   TRT01A [Xanomeline High Dose]     -0.18        0.34   -0.53     0.598    -0.85     0.49
   AGE                               -0.01        0.01   -1.20     0.230    -0.03     0.01
```

Differences from `broom`:

- Column names follow PascalCase / mixed-case conventions, not snake_case.
- The factor variable name and its level are *separated by spaces and square brackets*: `TRT01A [Xanomeline Low Dose]` instead of `TRT01AXanomeline Low Dose`. This is a small thing that turns out to matter a lot when downstream code needs to parse the term back into variable + level.
- `parameters` also defaults to including confidence intervals (`broom` does not, unless you set `conf.int = TRUE`).

Still missing in the default call: the reference row for `Placebo`. By default `parameters::model_parameters()` also drops it. To get it, you need `parameters::model_parameters(..., keep_reference = TRUE)`.

> **The honest take:** neither tidier alone gives a clinical-table-ready output. Both need an extra step to reinject the reference row. `broom.helpers::tidy_plus_plus()` is the package that bakes that extra step in by default — and that is why `{cardx}` chose it as the foundation of `ard_regression()`.

---

### 10.7 Side-by-Side on the Same Model

To make the differences concrete, here is the same fit pushed through both tidiers and the columns lined up.

```
   broom::tidy()                          parameters::model_parameters()

   term                                   Parameter
   estimate                               Coefficient
   std.error                              SE
   statistic                              z
   p.value                                p
   conf.low      (if conf.int = TRUE)     CI_low                     (by default)
   conf.high     (if conf.int = TRUE)     CI_high                    (by default)
   ---                                    Effects                    (fixed/random tag)
   ---                                    Component                  (which equation in a mixed model)
   ---                                    df_error                   (residual df)
```

There is no row that `broom` *adds* over `parameters`; the relationship is strictly that `parameters` is a superset of metadata. The only thing `broom` does that `parameters` does not is keep column names short and snake_case — which matters when downstream code needs to write `df$estimate` instead of `df$Coefficient`.

---

## Exercise 10.1 — Run `broom::tidy()` and See the Missing Reference Row

Tidy the pre-fit `model_fit` with `broom::tidy()`. Confirm by inspection that there is *no* `Placebo` row in the output. The fix is the next section's job; right now you just need to see the problem.

```{r ch10-ex1, exercise=TRUE, exercise.lines=8}
# Tidy the logistic model with broom
tidy_broom <- broom::___(model_fit)

# Show the term, estimate, and p.value columns
tidy_broom %>%
  select(term, estimate, p.value)
```

```{r ch10-ex1-hint-1}
# The function is broom::tidy
```

```{r ch10-ex1-solution}
tidy_broom <- broom::tidy(model_fit)
tidy_broom %>%
  select(term, estimate, p.value)
```

```{r ch10-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("term", "estimate", "p.value") %in% names(.result)) && !any(grepl("Placebo", .result$term)), "Correct. broom::tidy() returns one row per estimated coefficient and silently omits the reference level (Placebo). This is the gap that {broom.helpers} fills."),
  fail_if(~ TRUE, "Call broom::tidy(model_fit), then select term, estimate, and p.value.")
)
```

> **Read the output:** none of the `term` strings contains the word `Placebo`. The reference row is not "filtered out" — it never existed in the first place, because `broom` only outputs rows for which the model actually estimated a number.

---

## Exercise 10.2 — Run `parameters::model_parameters()` and Spot the Column-Name Differences

Tidy the same model with `parameters::model_parameters()`. Compare the column names to what `broom::tidy()` produced.

```{r ch10-ex2, exercise=TRUE, exercise.lines=10}
# Tidy the logistic model with parameters
tidy_param <- parameters::model_parameters(model_fit)

# Look at the column names returned
names(tidy_param)
```

```{r ch10-ex2-hint-1}
# Call names(tidy_param) on the parameters output.
```

```{r ch10-ex2-solution}
tidy_param <- parameters::model_parameters(model_fit)
names(tidy_param)
```

```{r ch10-ex2-check}
grade_result(
  pass_if(~ is.character(.result) && "Parameter" %in% .result && "Coefficient" %in% .result, "Correct. parameters::model_parameters() uses `Parameter` and `Coefficient` (PascalCase) where broom uses `term` and `estimate` (snake_case). Same numbers, different vocabulary."),
  fail_if(~ TRUE, "Pass the parameters output to names() to print the column names.")
)
```

> **Why this matters:** if you write `tidy_df$estimate` you get the broom column; on the parameters output that expression returns `NULL` and your downstream code silently produces wrong values. The column-name discipline is a real source of bugs.

---

### 10.8 The Reference Row Problem in Detail

A clinical Odds Ratio table needs to *show* the reference level with an OR of 1.00 (or estimate 0 on the log scale). The reasons are part regulatory expectation and part reader friendliness — the human eye needs to see "Placebo: 1.00" to understand what the other rows are being compared against.

Two facts conspire to make this hard:

1. **R's model engine never computes a number for the reference level.** It is the *baseline*; its effect is the intercept. Mathematically, the coefficient is exactly zero.
2. **`broom::tidy()` only emits rows for computed numbers.** Honoring fact (1), `broom` leaves the reference level out.

The result is a tidied data frame that is *mathematically faithful* but *clinically incomplete*. Three remedies exist:

- **Use `parameters::model_parameters(keep_reference = TRUE)`.** Works, but locks you into the `parameters` column-name dictionary.
- **Manually patch the row.** Possible but brittle: you have to compute the level name, set `estimate = 0` (or `1` if exponentiated), set the CI bounds appropriately, decide what `p.value` should be (typically `NA`), and remember to do this for every categorical variable in the model.
- **Use `broom.helpers::tidy_plus_plus()`.** Adds the reference row automatically, in the `broom` column dictionary, with `estimate = 0` and a sentinel `reference_row = TRUE` flag.

Option three is what `{cardx}` ships with. It is also the only one of the three that scales painlessly to models with several categorical predictors.

```
   GAP IN broom::tidy() OUTPUT      WHAT tidy_plus_plus() ADDS

   +-------------------------+      +---------------------------+
   | term                    |      | TRT01A: Placebo           |
   | TRT01AXanomeline Low... |  --> | TRT01A: Xanomeline Low... |
   | TRT01AXanomeline High.. |      | TRT01A: Xanomeline High.. |
   | (Intercept)             |      |                           |
   | AGE                     |      | reference_row column = T  |
   +-------------------------+      | label column = "Placebo"  |
                                    | variable column = "TRT01A"|
                                    +---------------------------+
```

---

### 10.9 `broom.helpers::tidy_plus_plus()` — The Rosetta Stone

`tidy_plus_plus()` is the single most important function in the clinical R reporting ecosystem. It is a thin-looking call that does five non-trivial jobs:

```r
broom.helpers::tidy_plus_plus(
  x                 = model_fit,
  exponentiate      = FALSE,       # set TRUE for ORs/HRs
  conf.int          = TRUE,
  conf.level        = 0.95,
  add_reference_rows = TRUE,       # the reinjection magic
  add_estimate_to_reference_rows = TRUE,
  tidy_fun          = NULL         # which tidier to wrap (default: broom::tidy)
)
```

The output uses `broom`'s column names *plus* a few additional columns specific to clinical tables:

| Added column        | Meaning                                                      |
|---------------------|--------------------------------------------------------------|
| `variable`          | The raw variable name as it appears in the model formula (e.g., `TRT01A`). |
| `var_class`         | The data class — `factor`, `numeric`, etc.                   |
| `var_type`          | `categorical` or `continuous`.                               |
| `label`             | The variable label (from a `label()` attribute, if present). |
| `var_label`         | Same idea, top-level variable label.                          |
| `contrasts`         | The contrast type used for categorical encoding.              |
| `reference_row`     | `TRUE` for the synthetically-added reference rows.            |
| `term`              | The raw term string from `broom::tidy()` for non-reference rows; constructed for reference rows. |

That `reference_row` flag is the hook that downstream layout tools use to print `1.00 (Reference)` instead of running the formatter on the raw estimate.

> **SAS analog:** `tidy_plus_plus()` is like a SAS macro that takes `ODS OUTPUT ParameterEstimates`, joins it with `PROC CONTENTS` to recover variable labels, joins it again with the class-level table to recover the reference row, and outputs a single unified dataset. SAS programmers usually write this macro themselves once per organisation; `broom.helpers` ships it as a package.

---

## Exercise 10.3 — Reinject the Reference Row with `tidy_plus_plus()`

Run the same `model_fit` through `broom.helpers::tidy_plus_plus()` and confirm that a row for `Placebo` appears with `reference_row == TRUE`.

```{r ch10-ex3, exercise=TRUE, exercise.lines=10}
# Tidy with broom.helpers — the reference row should reappear
tidy_helpers <- broom.helpers::tidy_plus_plus(model_fit)

# Show variable, label, reference_row, and estimate
tidy_helpers %>%
  select(variable, label, reference_row, estimate) %>%
  filter(variable == "___")
```

```{r ch10-ex3-hint-1}
# Filter on variable == "TRT01A" to focus on the treatment arm rows.
```

```{r ch10-ex3-solution}
tidy_helpers <- broom.helpers::tidy_plus_plus(model_fit)
tidy_helpers %>%
  select(variable, label, reference_row, estimate) %>%
  filter(variable == "TRT01A")
```

```{r ch10-ex3-check}
grade_result(
  pass_if(~ is.data.frame(.result) && any(.result$reference_row, na.rm = TRUE) && any(grepl("Placebo", .result$label)), "Correct. tidy_plus_plus() rebuilt the Placebo row, tagged it with reference_row = TRUE, and gave it estimate = 0 (which exponentiates to OR = 1)."),
  fail_if(~ TRUE, "Filter on variable == 'TRT01A' so you can see all three treatment levels including the reinjected Placebo reference row.")
)
```

> **What the flag means downstream:** `{cardx::ard_regression()}` reads `reference_row == TRUE` and decides not to format the estimate as a number; instead it places the literal string "Reference" in the row's `stat_fmt`. That is why your final OR table reads `1.00 (Reference)` instead of `0.00 (CI: NA, NA)`.

---

### 10.10 Inside `tidy_plus_plus()` — The Five Things It Adds

Pull `tidy_plus_plus()` apart and you find five clinical-reporting features stacked on top of `broom::tidy()`:

| Feature                              | What it does                                                                          |
|--------------------------------------|---------------------------------------------------------------------------------------|
| 1. Reference-row injection           | Adds a row per dropped reference level with `estimate = 0` and `reference_row = TRUE`.|
| 2. Variable-label recovery           | Reads `label()` attributes from the original data and stores them in `label`/`var_label`. |
| 3. Term parsing                       | Splits `TRT01AXanomeline Low Dose` into `variable = TRT01A`, `label = Xanomeline Low Dose`. |
| 4. Contrast metadata                  | Records which contrast (treatment, sum, helmert) was used to encode the factor.        |
| 5. Optional exponentiation            | Applies `exp()` to estimate and CI bounds in one consistent step (no manual `exp()`). |

Each of these maps to a manual step a SAS programmer might write in a data step after `ODS OUTPUT`. The package collapses them into one call.

#### 10.10.1 Reading variable labels

If your input data carries variable labels (set with `attr(data$AGE, "label") <- "Age, years"` or via the `{labelled}` package), `tidy_plus_plus()` picks those up automatically:

```r
adsl_labeled <- adsl
attr(adsl_labeled$AGE, "label")    <- "Age, years"
attr(adsl_labeled$TRT01A, "label") <- "Treatment Arm"

fit <- glm(RESP_BIN ~ TRT01A + AGE, data = adsl_labeled, family = binomial)
broom.helpers::tidy_plus_plus(fit) %>%
  select(variable, var_label, label)
```

The `var_label` column contains the top-level variable label; `label` contains the level-specific label (the factor level name, for categorical variables).

> **SAS analog:** the equivalent of writing `LABEL TRT01A = "Treatment Arm";` in a data step and then running `PROC LOGISTIC`. SAS prints the label in the output PDF; `tidy_plus_plus()` makes it available as a column you can map to a row label in your final table.

---

### 10.11 Exponentiation — The Single Most Common Reporting Mistake

A logistic regression coefficient is on the **log-odds** scale. A Cox regression coefficient is on the **log-hazard** scale. Neither of those is what a clinical reader expects to see; the reader wants the **Odds Ratio** or **Hazard Ratio** — the *exponentiated* value.

Three ways to exponentiate exist, and only one of them is reliable:

| Approach                                          | Outcome                                                                                  |
|---------------------------------------------------|-------------------------------------------------------------------------------------------|
| `exp(coef(fit))` after the fact                   | Exponentiates the point estimate but **not** the CI — silent inconsistency.               |
| `broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE)` | Exponentiates estimate and CI together. Reliable.                                  |
| `cardx::ard_regression(fit, exponentiate = TRUE)` | Same as above, packaged into the ARD shape. Recommended default.                          |

The bug is so common that it has a name in QC circles: the *exp-the-estimate-only* bug. The point estimate ends up at, say, OR = 2.50, and the CI ends up reported as `(0.92, 1.10)` because the bounds were exponentiated separately or not at all. The mismatch is visually obvious to a careful reviewer but easy to miss in an auto-generated table.

```
   WRONG                                       RIGHT
   exp(estimate)        = 2.50                 estimate     = 0.92      (log-odds)
   CI lower             = 0.92                 conf.low     = 0.45
   CI upper             = 1.10                 conf.high    = 1.65
   "OR = 2.50, CI: (0.92, 1.10)" <-- wrong     exponentiate = TRUE
                                               OR = 2.50, CI: (1.57, 5.20)
```

The fix is always the same: pass `exponentiate = TRUE` to the tidier (or to `ard_regression`), never to the post-hoc data step.

> **SAS analog:** the difference between calling `EXP()` only on `Estimate` in a data step vs. asking `PROC LOGISTIC` for the `OddsRatios` table directly. SAS programmers learn this lesson the same way.

---

### 10.12 Confidence Intervals — Wald, Profile, and Robust

The `conf.int` flag is binary — give me a CI or don't — but the *method* used to compute that CI varies by model class:

| Model class             | Default CI method in `broom::tidy()`         | Method in `parameters::model_parameters()`   |
|-------------------------|-----------------------------------------------|----------------------------------------------|
| `lm`                    | Wald (normal approximation)                  | Wald                                          |
| `glm`                    | Profile likelihood (`MASS::confint.glm`)     | Wald (faster, slightly less accurate)         |
| `coxph`                 | Wald                                         | Wald                                          |
| `lmer` / `glmer`        | Wald                                         | Wald or profile (configurable)                |
| `geeglm`                | Wald with robust SE                          | Wald with robust SE                           |

The defaults are not always identical, and for a logistic regression on a small sample the profile-likelihood CI from `broom` can differ noticeably from the Wald CI you would see from `parameters`. For a regulatory submission you should *state* which method you used and stick to it.

To force a specific CI method through `cardx::ard_regression()`, pass a `tidy_fun` that has the method baked in:

```r
my_tidier <- function(x, ...) broom::tidy(x, conf.int = TRUE, conf.level = 0.95, ...)

ard_reg <- cardx::ard_regression(model_fit, tidy_fun = my_tidier)
```

The `tidy_fun` argument is the seam that lets you take full control of the upstream tidier without losing the downstream `cardx` machinery.

> **SAS analog:** in `PROC LOGISTIC`, `CLPARM=PL` requests profile-likelihood CIs and `CLPARM=WALD` requests Wald CIs. The R-side equivalent is choosing the tidier and its arguments.

---

### 10.13 Customizing the Tidier in `cardx::ard_regression()`

`cardx::ard_regression()` defaults to `broom.helpers::tidy_plus_plus()`. You can override it with the `tidy_fun` argument when the default does not fit:

```r
# Default - clinical-style with reference rows
ard_default <- cardx::ard_regression(model_fit)

# Strict broom - no reference rows, no labels
ard_basic   <- cardx::ard_regression(model_fit, tidy_fun = broom::tidy)

# parameters-based - reference rows on, but column names get translated
ard_param   <- cardx::ard_regression(
  model_fit,
  tidy_fun = function(x, ...) parameters::model_parameters(x, keep_reference = TRUE, ...)
)
```

Three quick rules of thumb:

- **Default** (`tidy_plus_plus`) is right for clinical OR / HR tables, demographics-adjacent regression rows, and any output that follows the standard "Reference: 1.00" convention.
- **`broom::tidy`** is right for forest plots, exploratory model diagnostics, or any case where you do *not* want reference rows cluttering the data.
- **`parameters::model_parameters`** is right when you need mixed-model metadata (random effects, components) that `broom` does not include — but expect to translate column names downstream.

In `{cardx}` v0.x, there is also `ard_regression_basic()` which is shorthand for "use `broom::tidy` and skip enrichment." Its existence is a hint that the strict-`broom` path is a real, supported choice — not a fallback.

> **SAS analog:** swapping `tidy_fun` is equivalent to choosing which ODS dataset(s) you collect from a PROC. `tidy_plus_plus` collects everything and joins it; `broom::tidy` collects just `ParameterEstimates`; `parameters` collects a parallel-universe version of the same content.

---

## Exercise 10.4 — Build a Regression ARD with `cardx::ard_regression()`

Wrap `model_fit` in `cardx::ard_regression()` with `exponentiate = TRUE` so you get Odds Ratios. Filter the output to the `TRT01A` variable so you can see the Placebo reference row inline with the other levels.

```{r ch10-ex4, exercise=TRUE, exercise.lines=12}
# Build the regression ARD on the OR scale
reg_ard <- cardx::ard_regression(model_fit, exponentiate = ___)

# View formatted Odds Ratios for TRT01A only
reg_ard %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt) %>%
  filter(variable == "TRT01A", stat_name == "estimate")
```

```{r ch10-ex4-hint-1}
# exponentiate = TRUE turns log-odds into odds ratios.
```

```{r ch10-ex4-solution}
reg_ard <- cardx::ard_regression(model_fit, exponentiate = TRUE)
reg_ard %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt) %>%
  filter(variable == "TRT01A", stat_name == "estimate")
```

```{r ch10-ex4-check}
grade_result(
  pass_if(~ is.data.frame(.result) && nrow(.result) >= 3 && any(grepl("Placebo", .result$variable_level)), "Excellent. Three rows for TRT01A — Placebo (reference, OR = 1.00), Low Dose, and High Dose — all in the standard ARD schema."),
  fail_if(~ TRUE, "Pass exponentiate = TRUE and filter on variable == 'TRT01A' and stat_name == 'estimate'.")
)
```

> **Look at the Placebo row's `stat_fmt`:** it reads as a plain `1.00` (or whatever your formatter renders the literal 1.0 as). The CI columns for that row are NA. This is the ARD-shape encoding of "OR = 1 (Reference)" — and the downstream `{gtsummary}` table will substitute the string `"(Reference)"` automatically when it sees the `reference_row` flag.

---

### 10.14 Model Classes Both Tidiers Cover (and the Gaps)

Both `broom` and `parameters` are method-dispatched: each adds a `tidy.modelclass()` method for every class they support. Coverage is broad but not identical.

| Model class      | `broom::tidy()` | `parameters::model_parameters()` | Notes                                                                |
|------------------|-----------------|-----------------------------------|----------------------------------------------------------------------|
| `lm`             | yes             | yes                               | Both default to Wald CIs.                                            |
| `glm`            | yes             | yes                               | `broom` default CI is profile likelihood; `parameters` is Wald.       |
| `coxph` (survival) | yes           | yes                               | Both produce log-hazard estimates; `exponentiate = TRUE` recommended. |
| `lmer` (lme4)    | yes             | yes                               | `parameters` returns fixed and random structure rows; `broom` only fixed effects (`broom.mixed::tidy()` covers more). |
| `glmer`          | yes             | yes                               | Same as `lmer`.                                                       |
| `geeglm`         | yes             | yes                               | Robust SE by construction; both tidiers respect it.                   |
| `lmrob`, `glmrob` | partial        | yes                               | `broom` may miss some robust-fit columns.                             |
| `gam` (mgcv)     | yes             | yes                               | Smooth terms tidy into recognisable rows.                             |
| `polr` (ordinal) | yes             | yes                               | Both support, but column meanings differ slightly.                    |
| Bayesian models (`brms`, `rstanarm`) | via `broom.mixed` | yes              | `parameters` is generally more polished for Bayesian draws.           |

The practical heuristic: if your model is a standard frequentist regression (`lm`, `glm`, `coxph`), pick the default `tidy_plus_plus` path. If it is a mixed model or a Bayesian model, evaluate both tidiers and pick the one whose output the rest of your pipeline expects.

> **SAS analog:** the difference between `PROC LOGISTIC`, `PROC GENMOD`, `PROC MIXED`, `PROC PHREG`, `PROC NLMIXED`, `PROC GLIMMIX`, and `PROC MCMC` — each has its own ODS dataset shape. In R, `broom` and `parameters` *normalize* across model classes. Which one you pick determines which normalization you get.

---

### 10.15 Common Pitfalls When Choosing Between Tidiers

A short troubleshooting catalog drawn from real reporting projects:

| Symptom                                                          | Likely cause                                                    | Fix                                                                       |
|------------------------------------------------------------------|------------------------------------------------------------------|---------------------------------------------------------------------------|
| `$estimate` returns NULL                                         | Tidier used was `parameters`, column is `Coefficient`.           | Either rename the column or switch the tidier to `broom`.                  |
| Reference row missing in the final OR table                      | Used `broom::tidy` or `ard_regression_basic()`.                  | Use `ard_regression()` (default) or `tidy_plus_plus(add_reference_rows = TRUE)`. |
| Reference row appears, but with a numeric estimate instead of "Reference" | `reference_row` flag not honored by the layout tool.    | Pass through `apply_fmt_fun()` *after* `ard_regression()`; do not handcraft formatting.|
| OR look like ~ 0.4 to 0.6 instead of ~ 1.0                       | Forgot `exponentiate = TRUE`.                                    | Add it to `ard_regression()` or to `tidy_fun`.                            |
| CI bounds are nonsensical relative to the OR                     | Exponentiated estimate manually but not the CI.                  | Always exponentiate through the tidier, never by hand.                     |
| Column names mix `term`/`Parameter`                              | Two tidiers feeding the same downstream code.                    | Standardize on one (recommended: `broom` via `tidy_plus_plus`).            |
| Profile-likelihood CI is too slow                                | `glm` with many predictors and `broom::tidy` defaults.           | Pass `tidy_fun = function(x, ...) broom::tidy(x, conf.int = TRUE, ...)` with Wald.|
| Variable labels missing from the final table                     | Input data has no `label()` attributes.                          | Set them with `attr(data$var, "label") <- "..."` before fitting the model. |
| Factor level appears glued to variable name (`TRT01AXan...`)     | Used `broom::tidy` directly without `tidy_plus_plus`.            | Switch to `ard_regression()` so the term is parsed.                        |

> **SAS analog:** these are the same pitfalls you encountered the first time you tried to build a clinical Odds Ratio table from raw `ODS OUTPUT` datasets without writing a wrapper macro. The fixes here are package-level, not per-script.

---

### 10.16 The Big Picture — One Model, Three Possible Outputs

Putting the whole chapter on one sticky note:

```
                              +-------------------+
                              |   fitted model    |
                              +---------+---------+
                                        |
            +---------------------------+----------------------------+
            |                           |                            |
            v                           v                            v
   +--------------------+   +----------------------+   +--------------------------+
   | broom::tidy(fit)   |   | parameters::         |   | broom.helpers::          |
   |                    |   |   model_parameters() |   |   tidy_plus_plus()       |
   +--------------------+   +----------------------+   +--------------------------+
   - snake_case names    - PascalCase names           - broom column names
   - no reference row    - reference row optional     - reference row INJECTED
   - minimalist          - mixed-effects metadata     - variable labels recovered
                                                      - term parsed into var + level
                                                      - exponentiate handled
                                                      |
                                                      v
                                       +-------------------------------+
                                       |   cardx::ard_regression()     |
                                       |   --> ARD schema you know     |
                                       |   --> ready for gtsummary,    |
                                       |       flextable, ARS JSON     |
                                       +-------------------------------+
```

Pick a tidier, and the downstream is largely determined. `broom.helpers` + `cardx` is the recipe that "just works" for ~99% of clinical reports. The other 1% is when you need to drop into `parameters` for mixed-model metadata — and even then you re-emerge through `tidy_fun` back into the same `ard_regression()` call.

---

## Exercise 10.5 — Override the Default Tidier in `ard_regression()`

Build *two* ARDs from the same `model_fit`: one using the default tidier (`tidy_plus_plus`) and one using the strict `broom::tidy`. Compare the row counts. The default version should have **more rows** because it includes the reference level.

```{r ch10-ex5, exercise=TRUE, exercise.lines=15}
# ARD using the default (broom.helpers::tidy_plus_plus)
ard_default <- cardx::ard_regression(model_fit, exponentiate = TRUE)

# ARD using strict broom::tidy (no reference rows)
ard_basic <- cardx::ard_regression(
  model_fit,
  exponentiate = TRUE,
  tidy_fun = ___
)

# Compare row counts for the estimate stat only
list(
  default_rows = nrow(ard_default %>% filter(stat_name == "estimate")),
  basic_rows   = nrow(ard_basic   %>% filter(stat_name == "estimate"))
)
```

```{r ch10-ex5-hint-1}
# Pass broom::tidy by name (no parentheses) into tidy_fun.
```

```{r ch10-ex5-solution}
ard_default <- cardx::ard_regression(model_fit, exponentiate = TRUE)
ard_basic <- cardx::ard_regression(
  model_fit,
  exponentiate = TRUE,
  tidy_fun = broom::tidy
)
list(
  default_rows = nrow(ard_default %>% filter(stat_name == "estimate")),
  basic_rows   = nrow(ard_basic   %>% filter(stat_name == "estimate"))
)
```

```{r ch10-ex5-check}
grade_result(
  pass_if(~ is.list(.result) && .result$default_rows > .result$basic_rows, "Correct. The default ARD has more `estimate` rows because tidy_plus_plus injected the Placebo reference row; the strict-broom path did not."),
  fail_if(~ TRUE, "Set tidy_fun = broom::tidy on the basic ARD, then compare row counts.")
)
```

> **The trade-off in one line:** the default `tidy_plus_plus` path gives you a clinical-ready table at the cost of a row whose "estimate" is a synthetic 1.00. The strict-broom path gives you only what the model actually computed, at the cost of a missing row that a human reviewer expects to see.

---

## Exercise 10.6 — Use `parameters::model_parameters()` Inside `ard_regression()`

Wrap `parameters::model_parameters()` in a small lambda so it returns broom-compatible column names with `keep_reference = TRUE`, then pass it as the `tidy_fun` of `ard_regression()`. Confirm the resulting ARD still contains a Placebo row.

```{r ch10-ex6, exercise=TRUE, exercise.lines=20}
# Wrap parameters in a function that keeps reference rows
param_tidier <- function(x, ...) {
  parameters::model_parameters(x, keep_reference = ___, ...) %>%
    parameters::standardize_names(style = "broom")
}

# Build the ARD
ard_param <- cardx::ard_regression(
  model_fit,
  exponentiate = TRUE,
  tidy_fun = param_tidier
)

# Confirm Placebo is in the resulting ARD
ard_param %>%
  filter(variable == "TRT01A", stat_name == "estimate") %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt)
```

```{r ch10-ex6-hint-1}
# keep_reference = TRUE keeps the Placebo row in parameters' output.
```

```{r ch10-ex6-solution}
param_tidier <- function(x, ...) {
  parameters::model_parameters(x, keep_reference = TRUE, ...) %>%
    parameters::standardize_names(style = "broom")
}
ard_param <- cardx::ard_regression(
  model_fit,
  exponentiate = TRUE,
  tidy_fun = param_tidier
)
ard_param %>%
  filter(variable == "TRT01A", stat_name == "estimate") %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt)
```

```{r ch10-ex6-check}
grade_result(
  pass_if(~ is.data.frame(.result) && any(grepl("Placebo", .result$variable_level)), "Brilliant. You replaced the tidier engine but kept the rest of the pipeline. The Placebo row survives because the lambda enabled keep_reference; the column-name translation happened via standardize_names(style = 'broom')."),
  fail_if(~ TRUE, "Set keep_reference = TRUE inside the lambda so parameters keeps the Placebo row.")
)
```

> **What you just learned:** `tidy_fun` is the *only* seam you ever need to reach for in `ard_regression()`. Everything else — formatting, ARD shape, downstream layout — is unaffected by which engine you pick.

---

## Chapter 10 Knowledge Check

```{r ch10-quiz, echo=FALSE}
quiz(
  caption = "Chapter 10 — Check Your Understanding",

  question(
    "What is the closest R equivalent of SAS's `ODS OUTPUT ParameterEstimates = est;`?",
    answer("`glm()`"),
    answer("`summary()`"),
    answer("A tidier such as `broom::tidy(fit)` or `parameters::model_parameters(fit)`", correct = TRUE),
    answer("`print(fit)`"),
    allow_retry = TRUE
  ),

  question(
    "Why does `broom::tidy()` omit the Placebo row for a three-level treatment factor?",
    answer("It is a bug in the package."),
    answer("Placebo's coefficient is mathematically zero (the reference level), so no row is computed for it.", correct = TRUE),
    answer("Because `broom` filters out treatment arms named `Placebo`."),
    answer("Because the model failed to converge."),
    allow_retry = TRUE
  ),

  question(
    "Which package injects the reference row, recovers variable labels, and parses the term back into variable + level — all without changing the column-name dictionary used by `broom`?",
    answer("`{parameters}`"),
    answer("`{broom.helpers}`", correct = TRUE),
    answer("`{cardx}`"),
    answer("`{gtsummary}`"),
    allow_retry = TRUE
  ),

  question(
    "What is the name of the `broom::tidy()` column that `parameters::model_parameters()` calls `Coefficient`?",
    answer("`term`"),
    answer("`std.error`"),
    answer("`estimate`", correct = TRUE),
    answer("`statistic`"),
    allow_retry = TRUE
  ),

  question(
    "You forgot `exponentiate = TRUE`. What will your Odds Ratio column look like?",
    answer("All ones"),
    answer("Small numbers near zero (negative for protective effects, positive for risk) — they are log-odds, not odds ratios", correct = TRUE),
    answer("Reproducible but inflated to scientific notation"),
    answer("NA throughout"),
    allow_retry = TRUE
  ),

  question(
    "Which argument of `cardx::ard_regression()` lets you replace the upstream tidier (for example, swap in `parameters::model_parameters()` instead of `broom.helpers::tidy_plus_plus()`)?",
    answer("`exponentiate`"),
    answer("`tidy_fun`", correct = TRUE),
    answer("`statistic`"),
    answer("`variable`"),
    allow_retry = TRUE
  ),

  question(
    "What does the `reference_row = TRUE` flag in a tidied data frame tell a downstream layout tool to do?",
    answer("Skip the row entirely."),
    answer("Render the cell as `(Reference)` and suppress the CI/p-value, instead of formatting the synthetic estimate as a number.", correct = TRUE),
    answer("Highlight the row in red."),
    answer("Re-run the model with that level as the new reference."),
    allow_retry = TRUE
  ),

  question(
    "A SAS programmer is used to PROC LOGISTIC printing the reference class automatically. Which R function is the closest behavioral analog for clinical reporting?",
    answer("`broom::tidy()`"),
    answer("`broom.helpers::tidy_plus_plus()`", correct = TRUE),
    answer("`coef()`"),
    answer("`summary()`"),
    allow_retry = TRUE
  ),

  question(
    "You want to produce a forest plot that has *only* non-reference levels. Which `cardx` call is most appropriate?",
    answer("`ard_regression(fit)` with the default tidier"),
    answer("`ard_regression(fit, tidy_fun = broom::tidy)` (or `ard_regression_basic(fit)`)", correct = TRUE),
    answer("`ard_categorical(fit)`"),
    answer("`ard_continuous(fit)`"),
    allow_retry = TRUE
  ),

  question(
    "Why is exponentiating the estimate manually with `exp(coef(fit))` dangerous for clinical reporting?",
    answer("It is computationally slow."),
    answer("It exponentiates the point estimate but typically forgets the CI bounds, producing a printed CI that does not bracket the printed OR.", correct = TRUE),
    answer("`exp()` is not vectorized."),
    answer("It rounds the result automatically."),
    allow_retry = TRUE
  )
)
```

---

## Chapter 10 Summary {.congratulations}

You now understand:

- The two competing model-tidier ecosystems in R — **`{broom}`** and **`{parameters}`** — and the column-name dictionary that translates between them (`term` vs `Parameter`, `estimate` vs `Coefficient`, and so on).
- Why `broom::tidy()` silently **drops the reference row** for categorical predictors and what that means for clinical OR / HR tables.
- How **`broom.helpers::tidy_plus_plus()`** acts as a Rosetta Stone: it keeps `broom`'s column names, reinjects the reference row, recovers variable labels, parses terms into `variable + level`, and offers a single consistent `exponentiate` flag.
- The five things `tidy_plus_plus()` adds on top of `broom::tidy()` — reference rows, labels, term parsing, contrast metadata, exponentiation.
- The most common reporting bug in this space — exponentiating the estimate without the CI — and why letting the tidier handle exponentiation eliminates it.
- The role of the **`tidy_fun`** argument in `cardx::ard_regression()` as the single seam you ever need to swap out the upstream tidier without touching the rest of the pipeline.
- When to pick `broom`-only vs `tidy_plus_plus` vs `parameters` based on the model class and the downstream display.
- A practical pitfalls catalog for diagnosing missing reference rows, mis-exponentiated effects, mismatched CI methods, and column-name surprises.
- Where each piece sits in the bigger picture: fitted model → tidier → ARD → table — and that the choice of tidier locks in everything else.

**Next:** Chapter 11 (if present) builds on this regression foundation; otherwise return to the curriculum index. You now have everything you need to ship regression results in any tidier dialect and translate fluently between them.
