---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 5: Models, broom, and parameters"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.2
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Deep dive into statistical model tidying. Understand the roles of broom, parameters, broom.helpers, and cardx::ard_regression as the bridge from fitted R model objects to ARD-shaped rows.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(broom)
library(broom.helpers)
library(parameters)
library(survival)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )

# Derived dataset with a binary response flag for logistic exercises
adsl_resp <- adsl %>%
  mutate(COMPLFL = ifelse(EOSSTT == "COMPLETED", 1L, 0L))

# Pre-fit a logistic regression for use in many exercises
fit_logistic <- glm(
  COMPLFL ~ TRT01A + AGE,
  data    = adsl_resp,
  family  = binomial()
)

# Pre-fit a linear regression (PROC REG / PROC GLM analog) for use in exercises
fit_lm <- lm(
  AGE ~ TRT01A + SEX,
  data = adsl
)

# Pre-fit a Cox proportional-hazards model on Overall Survival
adtte_os <- adtte %>% filter(PARAMCD == "OS")
fit_cox <- coxph(
  Surv(AVAL, 1 - CNSR) ~ TRT01A + AGE,
  data = adtte_os
)
```

---

## Chapter 5: Models, broom, and parameters

### 5.1 Why Model Tidying Exists — The Problem It Solves

When you fit a statistical model in R, the object you get back is not a tidy table. It is a deeply nested *list* — sometimes ten or fifteen named slots deep — containing the coefficients, the residuals, the design matrix, the variance-covariance matrix, the call, the convergence diagnostics, and metadata about the family and link.

Try it: type `str(fit_logistic)` on any fitted model in R, and you will see a wall of output that scrolls off the screen. To pull out "the coefficient for TRT01A and its p-value," you would historically have written something like:

```r
summary(fit_logistic)$coefficients["TRT01AXanomeline High Dose", c("Estimate", "Pr(>|z|)")]
```

That is brittle, hard to read, and different for every model class (`lm` uses `Pr(>|t|)`, `glm` uses `Pr(>|z|)`, `coxph` uses `Pr(>|z|)` but in a different slot, `lmer` uses no p-value column at all, and so on).

> **SAS analog:** in SAS, every PROC writes its results to *named* `ODS OUTPUT` datasets — `PROC LOGISTIC` writes `ParameterEstimates`, `PROC REG` writes `ParameterEstimates`, `PROC PHREG` writes `ParameterEstimates`, `PROC MIXED` writes `SolutionF`. The dataset names overlap, but the *column names* in each dataset differ (`Estimate`/`StdErr`/`Probt` here, `Estimate`/`StdErr`/`Probz` there, etc.), forcing the programmer to write per-PROC post-processing every time. R has the same disease in a different costume.

The tidier ecosystem is the cure: a single set of column names (`term`, `estimate`, `std.error`, `statistic`, `p.value`, `conf.low`, `conf.high`) that work for every supported model class. Once your code is written against those columns, swapping `lm` for `glm` for `coxph` for `lmer` becomes a one-line change.

---

### 5.2 The Anatomy of a Fitted R Model

Before you can appreciate what `tidy()` does, look at the raw shape of a model object.

A fitted `glm` (logistic regression) is an S3 list with roughly twenty slots. The interesting ones are:

| Slot                  | What it stores                                                       |
|-----------------------|----------------------------------------------------------------------|
| `coefficients`        | Named numeric vector — the point estimates on the link scale         |
| `residuals`           | Working residuals at convergence                                     |
| `fitted.values`       | Predictions on the response scale                                    |
| `family`              | `family()` object — name, link, variance function                    |
| `linear.predictors`   | Predictions on the linear-predictor (link) scale                     |
| `deviance` / `null.deviance` | Model fit and intercept-only deviance                         |
| `df.residual`         | Residual degrees of freedom                                          |
| `aic`                 | Akaike Information Criterion                                         |
| `model`               | The model frame (the actual data used)                               |
| `terms`               | The formula expanded into terms metadata                             |
| `qr`                  | QR decomposition of the design matrix (needed for vcov)              |
| `call`                | Captured function call (used by `update()`)                          |

`summary(fit_logistic)` adds **more** slots — most importantly `coefficients` (a *matrix*, not the original vector, with columns for `Estimate`, `Std. Error`, `z value`, `Pr(>|z|)`).

The variance-covariance matrix, needed for confidence intervals, lives behind `vcov(fit_logistic)`. The CIs themselves are computed by `confint(fit_logistic)`, which by default does a profile likelihood for `glm` (slow) and Wald for `lm` (fast).

> **SAS analog:** the things in `coefficients` correspond to the `Estimate` column of `ODS OUTPUT ParameterEstimates`; the SE/z/p columns correspond to the rest of that dataset. The variance-covariance matrix corresponds to `ODS OUTPUT CovB`. Convergence and fit diagnostics live in `ODS OUTPUT FitStatistics`, `ConvergenceStatus`, `ModelInfo`, etc. — each its own dataset. R puts all of these into one nested object instead of a directory of datasets.

The tidiers' job is to flatten that nested object into a small set of rectangular tibbles.

---

### 5.3 The R Tidier Ecosystem — Three Packages, Three Jobs

There is not one tidier package; there are three that compose. Understanding which does what is the difference between debugging happily and chasing bugs across package boundaries.

| Package           | Function                  | What it returns                                                                 |
|-------------------|---------------------------|---------------------------------------------------------------------------------|
| `{broom}`         | `tidy(model)`             | One row per coefficient: `term`, `estimate`, `std.error`, `statistic`, `p.value` |
| `{broom}`         | `glance(model)`           | One row per model: fit statistics (AIC, BIC, deviance, R², log-likelihood, etc.) |
| `{broom}`         | `augment(model, data)`    | The original data plus per-observation columns (`.fitted`, `.resid`, `.hat`, …)  |
| `{parameters}`    | `model_parameters(model)` | Similar to `tidy()` but with richer metadata and more model classes covered      |
| `{broom.helpers}` | `tidy_plus_plus(model)`   | `tidy()` output **plus** reference rows, labels, contrasts, term grouping        |
| `{cardx}`         | `ard_regression(model)`   | `tidy_plus_plus()` output cast into the ARD schema you already know              |

```
   +-----------+   tidy()         +-------------+
   |   model   |  -------------->  |  flat tibble |   (broom)
   +-----------+                   +-------------+
        |                                 |
        | glance()                        | tidy_plus_plus()
        v                                 v   (broom.helpers)
   +-------------+                   +-----------------+
   |  one-row    |                   |  flat tibble +  |
   |  fit stats  |                   |  reference rows |
   +-------------+                   |  + var labels   |
                                     +-----------------+
                                             |
                                             | ard_regression() (cardx)
                                             v
                                     +-----------------+
                                     |   ARD rows      |
                                     |   (cards shape) |
                                     +-----------------+
```

> **SAS analog:** `tidy()` is `ODS OUTPUT ParameterEstimates`. `glance()` is `ODS OUTPUT FitStatistics`/`ModelInfo`. `augment()` is `OUTPUT OUT=` adding `_PREDICTED_`, `_RESIDUAL_`, etc. `broom.helpers` is the SAS programmer's "add a header row for the reference level so the table reads cleanly" macro. `ard_regression()` is the formal ARD wrapper around that macro.

---

### 5.4 `broom::tidy()` — The One-Row-Per-Coefficient Workhorse

`tidy()` is the most-used function in the whole ecosystem. It dispatches on the class of the model object and returns a tibble with a standard column vocabulary.

For a `glm`:

```r
broom::tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE)
#> # A tibble: 4 x 7
#>   term                       estimate std.error statistic p.value conf.low conf.high
#>   <chr>                         <dbl>     <dbl>     <dbl>   <dbl>    <dbl>     <dbl>
#> 1 (Intercept)                   ...      ...       ...     ...      ...       ...
#> 2 TRT01AXanomeline High Dose    ...      ...       ...     ...      ...       ...
#> 3 TRT01AXanomeline Low Dose     ...      ...       ...     ...      ...       ...
#> 4 AGE                           ...      ...       ...     ...      ...       ...
```

The important arguments:

| Argument        | What it does                                                                      |
|-----------------|-----------------------------------------------------------------------------------|
| `conf.int`      | If `TRUE`, add `conf.low`/`conf.high` columns. Default `FALSE`.                   |
| `conf.level`    | Confidence level for the CI. Default `0.95`.                                      |
| `exponentiate`  | If `TRUE`, exponentiate `estimate` and CI bounds. For logistic → OR; Cox → HR.    |
| `effects`       | For mixed models: `"fixed"`, `"random"`, or `"ran_pars"` selects which terms.     |

#### 5.4.1 Why `exponentiate = TRUE` exists

Most modeling functions in R return coefficients on the **link scale**, not the response scale:

- `glm(family = binomial)` returns log-odds. Exponentiate → odds ratios.
- `glm(family = poisson)` returns log-rate. Exponentiate → rate ratios.
- `coxph()` returns log-hazard. Exponentiate → hazard ratios.
- `lm()` returns the response directly. Do **not** exponentiate; it makes no sense.

The flag belongs on the *tidy* call (or on `ard_regression`), not after the fact. Doing it manually with `exp(estimate)` is a common source of bugs because programmers forget to also exponentiate the CI bounds — and exponentiating *only* the point estimate produces a number that does not match the CI you also print.

> **SAS analog:** `PROC LOGISTIC` automatically prints both "Analysis of Maximum Likelihood Estimates" (log-odds) and "Odds Ratio Estimates" (exponentiated). `PROC PHREG` prints "Analysis of Maximum Likelihood Estimates" and "Hazard Ratios". `PROC REG` prints only the unexponentiated table because the response scale is already meaningful. R puts both behind a single boolean flag.

#### 5.4.2 The naming convention for `term`

The `term` column uses R's internal factor encoding rules:

- Continuous covariates → the variable name (`AGE`, `BMI`).
- Categorical covariates → `<var><level>` for each non-reference level (`TRT01AXanomeline High Dose`).
- Interactions → `<term1>:<term2>` (`TRT01AXanomeline High Dose:SEXM`).

That naming is fine for analysis but ugly in a clinical table. Fixing it is the job of `broom.helpers` (section 5.6).

> **SAS analog:** the `Parameter` column of `ODS OUTPUT ParameterEstimates` from `PROC LOGISTIC` uses the same `<varname><level>` convention by default (or the `ClassLevel` column if you ask for `CLASS / PARAM=GLM`).

---

### 5.5 `broom::glance()` — The One-Row-Per-Model Fit Summary

Where `tidy()` returns one row per coefficient, `glance()` returns **exactly one row** per model. The columns vary by model class — that is by design — and contain whatever fit statistics are conventional for that family.

For a `glm`:

```r
broom::glance(fit_logistic)
#> # A tibble: 1 x 8
#>   null.deviance df.null logLik   AIC   BIC deviance df.residual  nobs
#>           <dbl>   <int>  <dbl> <dbl> <dbl>    <dbl>       <int> <int>
#> 1          ...     ...    ...   ...   ...      ...         ...   ...
```

For an `lm`: `r.squared`, `adj.r.squared`, `sigma`, `statistic` (F), `p.value`, `df`, `logLik`, `AIC`, `BIC`, `deviance`, `df.residual`, `nobs`.

For a `coxph`: `n`, `nevent`, `statistic.log`, `p.value.log`, `statistic.sc`, `p.value.sc`, `statistic.wald`, `p.value.wald`, `r.squared`, `concordance`, `AIC`, `BIC`.

`glance()` is what you use to report the **goodness-of-fit footer** of a regression table: "n = 254, deviance = 312.4, AIC = 318.4."

> **SAS analog:** the `ODS OUTPUT FitStatistics=` dataset. In `PROC LOGISTIC` this contains AIC, SC (BIC), -2LogL; in `PROC REG` it contains R-Square, Adj R-Sq; in `PROC PHREG` it contains -2 LOG L, AIC, SBC, plus the three score / Wald / log-likelihood χ² tests. `glance()` collapses those into one named row.

---

### 5.6 `broom::augment()` — Adding Per-Observation Columns

`augment()` is the third of broom's trio. It takes the model and the original data and returns the data with per-observation diagnostics appended:

| Column     | Meaning                                                              |
|------------|----------------------------------------------------------------------|
| `.fitted`  | Predicted value (response scale)                                     |
| `.resid`   | Residual                                                             |
| `.hat`     | Leverage / hat-value                                                 |
| `.sigma`   | Estimated standard deviation excluding this observation              |
| `.cooksd`  | Cook's distance — influence measure                                  |
| `.std.resid` | Standardized residual                                              |

You will rarely use `augment()` to build an ARD — it produces *per-row* diagnostics, not summary statistics. It belongs more to model diagnostics and plotting workflows. But it is good to know it exists.

> **SAS analog:** the `OUTPUT OUT= dataset PREDICTED= RESIDUAL= COOKSD= H=` statement in `PROC REG`, `PROC GLM`, `PROC LOGISTIC`. Same idea, identical purpose.

---

### 5.7 `parameters::model_parameters()` — The easystats Alternative

`{parameters}` from the **easystats** suite is an alternative to `broom::tidy()`. It covers a wider range of model classes (about 230 vs. broom's ~100), and it returns slightly richer metadata.

```r
parameters::model_parameters(fit_logistic, exponentiate = TRUE)
#> # Fixed Effects
#>
#> Parameter                  | Odds Ratio |   SE |       95% CI |     z |     p
#> ----------------------------------------------------------------------------
#> (Intercept)                |       ...  | ...  | [..., ...]  |  ...  | ...
#> TRT01A [Xanomeline High..]  |       ...  | ...  | [..., ...]  |  ...  | ...
```

Notice:

- The print method is opinionated and friendly (it renames `estimate` → `Odds Ratio` when you exponentiate a logistic model).
- The underlying object is still a data frame — `as.data.frame(model_parameters(fit_logistic))` gives you the flat tibble.
- Columns are spelled `Parameter`, `Coefficient`, `SE`, `CI_low`, `CI_high`, `z`, `p` rather than `term`, `estimate`, `std.error`, `conf.low`, `conf.high`, `statistic`, `p.value`.

`{cardx}` will happily consume either one — you pass `tidy_fun = broom::tidy` or `tidy_fun = parameters::model_parameters` to `ard_regression()` and it adapts. The default is `broom.helpers::tidy_plus_plus()` (which itself wraps `broom::tidy()`).

| When to prefer `broom::tidy()`             | When to prefer `parameters::model_parameters()`  |
|--------------------------------------------|--------------------------------------------------|
| You use the standard model classes (lm, glm, coxph, lme4) | You hit a model class broom does not support |
| You want fewest dependencies                | You want robust-SE / bootstrap CIs out of the box |
| You will pass the result to `ard_regression()` | You want a pretty stand-alone print method |

> **SAS analog:** there isn't one for `{parameters}` specifically — the closest equivalent is using `PROC GLIMMIX` with `EMPIRICAL=` for robust SEs, then post-processing the `ODS OUTPUT ParameterEstimates` to rename and format. The cleanliness of `model_parameters()` is what motivated easystats in the first place.

---

### 5.8 `broom.helpers::tidy_plus_plus()` — The Clinical Reporting Layer

`broom.helpers` is the *unsung hero* of clinical R reporting. It sits between `broom::tidy()` and `cardx::ard_regression()` and adds five things that you would otherwise have to add manually for every table:

1. **Variable labels.** If your data frame has `attr(x, "label")` strings (a SAS-style label), `broom.helpers` finds them and puts them on the variable rows.
2. **Categorical reference rows.** When a `TRT01A` factor has levels `Placebo / Low / High` and `Placebo` is the reference, base `tidy()` simply omits Placebo. `broom.helpers` *adds back* a row with `estimate = 1` (when exponentiated) or `estimate = 0` (when not) and a flag `reference_row = TRUE` so the table can show it as the baseline.
3. **Term grouping.** It tags every row with the **original variable name** in a column called `variable` — so when you have a factor that produced three terms, you can still group those three terms back under one variable header.
4. **Contrast labels.** It records what kind of contrast was used (treatment, sum, Helmert, …) so the table knows what to footnote.
5. **N per term.** It records the per-coefficient sample size, useful for footnotes like "n = 84 in TRT01A=High Dose."

```r
broom.helpers::tidy_plus_plus(fit_logistic, exponentiate = TRUE, conf.int = TRUE) |>
  dplyr::select(variable, label, term, reference_row, estimate, conf.low, conf.high, p.value)
```

You will see one row per *level* of every factor — including the reference level — plus one row per continuous variable. That row count is exactly what a publication-style regression table needs.

> **SAS analog:** the `CLASS / PARAM=REF REF='Placebo'` plus a hand-written DATA step that adds the reference row for display. `broom.helpers` makes that automatic.

---

### 5.9 `cardx::ard_regression()` — The ARD Wrapper

`ard_regression()` is `tidy_plus_plus()` wrapped to emit the **ARD schema** you learned in Chapter 1: `variable`, `variable_level`, `stat_name`, `stat_label`, `stat`, `stat_fmt`, `fmt_fun`, etc.

The advantage over calling `tidy_plus_plus()` directly is that the output now plays in the same `bind_ard()` / `apply_fmt_fun()` / `unlist_ard_columns()` pipeline as every other ARD in your project — no special-case code for regression rows.

```r
cardx::ard_regression(
  x            = fit_logistic,
  exponentiate = TRUE,
  conf.int     = TRUE
)
```

Returns ARD rows where:

- `variable` = the original column name in your data.
- `variable_level` = the factor level (or `NA` for continuous).
- `stat_name` ∈ {`estimate`, `std.error`, `statistic`, `p.value`, `conf.low`, `conf.high`, `n_obs`, `n_event`, `reference_row`, `label`, ...}.
- `stat` is the value (list-column).

The full argument list worth knowing:

| Argument          | Purpose                                                                                  |
|-------------------|------------------------------------------------------------------------------------------|
| `x`               | The fitted model object.                                                                 |
| `tidy_fun`        | Which tidier to use under the hood. Default `broom.helpers::tidy_with_broom_or_parameters`. |
| `exponentiate`    | Exponentiate point estimates and CI bounds. Passed through to the tidier.                |
| `conf.int`        | Include confidence intervals.                                                            |
| `conf.level`      | Confidence level (default `0.95`).                                                       |
| `add_estimate_to_reference_rows` | If `TRUE`, the reference row gets `1` (exponentiated) or `0` (not).       |
| `stats_to_remove` | Optional character vector to drop unwanted `stat_name` values from the output.           |
| `fmt_fun`         | Same nested-list format you learned in Chapter 3.                                        |

#### 5.9.1 `ard_regression_basic()` — when you only want the tidy() rows

Sometimes you do *not* want reference rows, labels, or grouping — for example, when you are building a forest plot programmatically and want one row per non-reference term, no more, no less. `cardx::ard_regression_basic()` skips the `broom.helpers` layer and emits ARD rows that come straight from `broom::tidy()` (or whatever `tidy_fun` you supply).

| Function                  | Adds reference rows? | Adds labels? | Variable grouping? |
|---------------------------|----------------------|--------------|--------------------|
| `ard_regression()`        | Yes (by default)     | Yes          | Yes                |
| `ard_regression_basic()`  | No                   | No           | No (term-level only) |

> **SAS analog:** `ard_regression()` is the formatted "Odds Ratio Estimates" table you see in the PDF; `ard_regression_basic()` is the raw `ODS OUTPUT ParameterEstimates` dataset.

---

### 5.10 Supported Model Families and Their SAS PROC Analogs

The whole point of standardization is that one set of code handles many model types. Here is the comprehensive mapping you will use day-to-day.

| R fit call                                       | Model type                          | Closest SAS PROC                                  |
|--------------------------------------------------|-------------------------------------|---------------------------------------------------|
| `lm(y ~ x)`                                      | Linear regression                   | `PROC REG`, `PROC GLM`                            |
| `glm(y ~ x, family = binomial)`                  | Logistic regression                 | `PROC LOGISTIC`                                   |
| `glm(y ~ x, family = poisson)`                   | Poisson regression for rates        | `PROC GENMOD DIST=POISSON`                        |
| `glm(y ~ x, family = quasipoisson)`              | Overdispersed Poisson               | `PROC GENMOD DIST=POISSON SCALE=DEVIANCE`         |
| `MASS::glm.nb(y ~ x)`                            | Negative binomial regression        | `PROC GENMOD DIST=NEGBIN`                         |
| `MASS::polr(y ~ x)`                              | Ordinal (proportional-odds) logistic | `PROC LOGISTIC ... / LINK=CLOGIT`                |
| `nnet::multinom(y ~ x)`                          | Multinomial (nominal) logistic      | `PROC LOGISTIC ... / LINK=GLOGIT`                 |
| `survival::coxph(Surv(time, event) ~ x)`         | Cox proportional hazards            | `PROC PHREG`                                      |
| `survival::survreg(Surv(...) ~ x)`               | Parametric survival (AFT)           | `PROC LIFEREG`                                    |
| `lme4::lmer(y ~ x + (1 | id))`                   | Linear mixed-effects                | `PROC MIXED`                                      |
| `lme4::glmer(y ~ x + (1 | id), family = binomial)` | Generalized linear mixed-effects | `PROC GLIMMIX`                                    |
| `nlme::gls(y ~ x, correlation = ...)`            | GLS with correlation structure      | `PROC MIXED REPEATED`                             |
| `geepack::geeglm(y ~ x, id = id, family = ...)`  | GEE                                 | `PROC GENMOD REPEATED`                            |

Every one of these returns a fitted object that `broom::tidy()` recognises, and therefore that `ard_regression()` handles, with no per-class code on your part.

```
   fit <- lm(...) | glm(...) | coxph(...) | lmer(...) | polr(...) | ...
                                |
                                |  (one call works for all)
                                v
                       ard_regression(fit, ...)
                                |
                                v
                          ARD rows (standard schema)
```

> **SAS analog:** imagine if every modeling PROC's `ODS OUTPUT ParameterEstimates` dataset had identical column names, factor-level labelling, and reference-row handling. That uniformity is what the R tidier ecosystem gives you for free.

#### 5.10.1 Model-class specifics worth knowing

- **`lm`** — the simplest case. No `exponentiate`. CI defaults to Wald and is computed instantly.
- **`glm`** — `exponentiate = TRUE` is almost always what you want for binomial/Poisson. The default CI is *profile likelihood* (slow for big models but more accurate than Wald near boundaries). Pass `conf.method = "wald"` to force Wald.
- **`coxph`** — `exponentiate = TRUE` produces hazard ratios. The fitted object also carries the proportional-hazards assumption diagnostics, accessible via `survival::cox.zph()` separately.
- **`lmer` / `glmer`** — by default `tidy()` returns only fixed effects. Pass `effects = c("fixed", "ran_pars")` to also get random-effect variance components. There is no p-value column from `lmer` itself (a deliberate choice by the lme4 authors); the `{lmerTest}` package adds Satterthwaite-approximated p-values.
- **`polr`** — gives proportional-odds coefficients. Exponentiate to get cumulative odds ratios.
- **`multinom`** — produces one set of coefficients per non-reference response level. The `y.level` column distinguishes them.

---

### 5.11 The Reference-Row Convention in Detail

Reference rows are the single most-asked-about feature of `ard_regression()` for SAS programmers, because SAS does *not* add them automatically.

Consider a logistic regression with `TRT01A` as a 3-level factor (`Placebo / Xanomeline Low Dose / Xanomeline High Dose`) and `Placebo` chosen as the reference. The raw `tidy()` output has three rows (`Intercept`, `TRT01AXanomeline High Dose`, `TRT01AXanomeline Low Dose`) — Placebo never appears.

A publication-style regression table needs *four* rows for `TRT01A`:

```
   TRT01A
     Placebo                  1.00 (ref)         —
     Xanomeline Low Dose      1.34 (0.92, 1.95)  0.123
     Xanomeline High Dose     1.78 (1.21, 2.62)  0.003
```

That extra "Placebo  1.00 (ref)  —" row is the **reference row**. `ard_regression()` produces it by default. Each reference row in the ARD has:

- `stat_name = "reference_row"`, `stat = TRUE` to flag it.
- `stat_name = "estimate"`, `stat = 1` (when `exponentiate = TRUE`) or `stat = 0` (when not).
- `stat_name = "p.value"`, `stat = NA`.
- `stat_name = "conf.low"`, `stat = NA`, similar for `conf.high`.

The downstream renderer (`gtsummary`, `tfrmt`, your own code) reads `reference_row` and decides how to display it ("1.00 (ref)", "—", whatever your SAP specifies).

To **suppress** reference rows entirely, use `ard_regression_basic()` instead, or pass `add_estimate_to_reference_rows = FALSE` (which keeps the rows but leaves their estimate as `NA`).

> **SAS analog:** in `PROC LOGISTIC` you would `ODS OUTPUT ParameterEstimates=pe;` then write a DATA step that, for each `ClassVar`, finds the reference level and inserts a synthetic row with `Estimate = 0` (or `1` after exponentiation). `broom.helpers` does that for you.

---

### 5.12 Confidence Interval Methods — Wald, Profile, Robust

`conf.int = TRUE` is binary, but the *method* used to compute the CI varies by model class.

| Model        | Default CI method      | When to override                                      |
|--------------|------------------------|-------------------------------------------------------|
| `lm`         | Wald (`stats::confint.lm`) | Almost never — Wald is exact for `lm`.            |
| `glm`        | Profile likelihood     | Override to Wald (`conf.method = "wald"`) for speed on very large models. |
| `coxph`      | Wald                   | Profile available via `survival::confint.coxph(parm, method = "profile")`. |
| `lmer`       | Wald                   | `conf.method = "profile"` or `"boot"` for accuracy; both are slow. |

Robust / sandwich SEs (the SAS `EMPIRICAL=` / `COVB(SANDWICH)` analogs) are not enabled by default in any of these. To use them, pre-process the model with `sandwich::vcovHC()` or refit with `{rms}` / `{geepack}`, then pass the robust-vcov result via the `tidy_fun` argument.

> **SAS analog:** `PROC LOGISTIC` defaults to Wald CIs in the "Wald Confidence Limits" column; profile-likelihood CIs require the `CLPARM=PL` option. R reverses that default for `glm`: profile is the default, Wald requires an opt-in.

---

### 5.13 Customizing the Tidier via `tidy_fun`

The escape hatch when defaults are not enough.

```r
ard_regression(
  x        = fit_glm_with_clustered_data,
  tidy_fun = \(fit, ...) broom::tidy(fit, conf.int = TRUE, ...) |>
                          dplyr::mutate(
                            std.error = sqrt(diag(sandwich::vcovCL(fit, cluster = ~ subject_id)))
                          )
)
```

Any function that takes a model and returns a tibble with the standard column names (`term`, `estimate`, `std.error`, ...) is a valid `tidy_fun`. This is how you splice in:

- Robust / sandwich SEs (`sandwich::vcovHC` or `vcovCL`).
- Penalized-model CIs (`rms::lrm` outputs).
- Bayesian posterior summaries (`broom.mixed::tidy.brmsfit`).

The contract is the column vocabulary. Anything that meets it, works.

---

### 5.14 Formatting Regression ARDs

Identical mechanics to Chapter 3:

```r
ard_regression(
  x            = fit_logistic,
  exponentiate = TRUE,
  conf.int     = TRUE,
  fmt_fun      = list(
    estimate   = 2,    # OR to 2 decimals
    conf.low   = 2,
    conf.high  = 2,
    p.value    = 3     # p-values to 3 decimals
  )
) |>
  apply_fmt_fun()
```

Or update after the fact:

```r
ard_reg |>
  update_ard_fmt_fun(p.value = 4) |>
  apply_fmt_fun()
```

The `stat_fmt` column is the formatted string; `stat` remains the unrounded numeric. Layout engines pick whichever they need — typically `stat_fmt` for the rendered cell, `stat` for re-computation or rounding-sensitive QC.

> **SAS analog:** the `FORMAT` statement applied during `PROC REPORT` rendering. ARD separates the unrounded value (`stat`) from the rounded display string (`stat_fmt`) so both are auditable.

---

### 5.15 Stacking Model ARDs with Other ARDs

Because `ard_regression()` returns the standard ARD schema, you can `bind_ard()` it onto descriptive ARDs from `{cards}` and inferential ARDs from `{cardx}`. A typical "demographics + regression" mega-table is one stack:

```r
ard_descrip <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A
)

ard_model <- ard_regression(
  x = glm(COMPLFL ~ TRT01A + AGE + SEX, data = adsl_resp, family = binomial),
  exponentiate = TRUE,
  conf.int     = TRUE
)

ard_all <- bind_ard(ard_descrip, ard_model)
```

`ard_all` is one tall dataset feeding one layout call. The fact that the descriptive `AGE` block and the regression `AGE` row share the same `variable = "AGE"` value is exactly what lets the layout tool merge them visually under one variable heading.

---

### 5.16 Common Pitfalls

A short list, ranked by how often they bite:

| Symptom                                                       | Cause                                                                 | Fix                                                            |
|---------------------------------------------------------------|----------------------------------------------------------------------|----------------------------------------------------------------|
| OR / HR look like very small numbers (~ −0.5 to 0.5)           | Forgot `exponentiate = TRUE` — those are log-odds / log-hazards.       | Add `exponentiate = TRUE` to `ard_regression()`.               |
| CI bounds don't bracket the point estimate after exponentiation | Exponentiated the estimate manually but not the CI.                   | Always pass `exponentiate = TRUE` through `tidy()`/`ard_regression()`. |
| Reference row missing from the table                          | Used `ard_regression_basic()` instead of `ard_regression()`.          | Use `ard_regression()` unless you specifically want raw rows.   |
| Wrong baseline level appears as reference                     | R uses the first factor level as reference. Yours is alphabetically last. | `relevel(factor, ref = "Placebo")` before fitting.        |
| p-values empty for `lmer` fits                                | `lme4` deliberately omits p-values.                                   | Load `{lmerTest}` (`library(lmerTest)` before `lmer()`).        |
| `glm` CI is slow on a 100k-row dataset                        | Profile likelihood is the default and it refits the model per term.   | Pass `conf.method = "wald"` via `tidy_fun`.                    |
| Factor level labels show up as `TRT01AXanomeline High Dose`   | Raw `broom::tidy()` does not parse the level out.                     | Use `ard_regression()` (which uses `broom.helpers`) — the `variable_level` column gets the clean level. |
| Variable labels missing in the rendered table                 | The ADaM column has no `label` attribute.                             | Set it: `attr(adsl$AGE, "label") <- "Age (years)"` before fitting. |

---

## Exercise 5.1 — Tidy a Logistic Regression Manually with `broom::tidy()`

The model `fit_logistic` (response flag `COMPLFL` vs. `TRT01A + AGE`) is pre-fitted in your environment. Use `broom::tidy()` to flatten it into a clean data frame with exponentiated coefficients (Odds Ratios) and 95% confidence intervals.

```{r ch5-ex1, exercise=TRUE, exercise.lines=10}
# Tidy the pre-loaded fit_logistic model
# Hint: call tidy() with conf.int = TRUE and exponentiate = TRUE

fit_logistic %>%
  ___(conf.int = ___, exponentiate = ___)
```

```{r ch5-ex1-hint-1}
# Call tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE)
```

```{r ch5-ex1-solution}
tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE)
```

```{r ch5-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "conf.low" %in% names(.result), "Excellent! broom::tidy() converted the model into a flat tibble of one row per coefficient with standard column names (`term`, `estimate`, `std.error`, `statistic`, `p.value`, `conf.low`, `conf.high`)."),
  fail_if(~ TRUE, "Try calling tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE).")
)
```

> **Read the rows:** `estimate` is the Odds Ratio (because of `exponentiate = TRUE`), `std.error` is on the log-odds scale (not the OR scale — this is the SE used to construct CIs *before* exponentiation, then exponentiated), `statistic` is the Wald z-value, `p.value` is the two-sided p, `conf.low`/`conf.high` are the 95% CI on the OR scale.
>
> **SAS check:** these are the same six numbers you would read off "Odds Ratio Estimates" + "Analysis of Maximum Likelihood Estimates" in `PROC LOGISTIC`. If you ran the same data through both, the OR and CI should agree to several decimal places.

---

## Exercise 5.2 — Generate a Regression ARD with `cardx::ard_regression()`

Instead of tidying manually and then mapping the result to an ARD schema by hand, pass the fitted model directly to `cardx::ard_regression()`. The function tidies, adds categorical reference rows, attaches variable labels, and structures everything as a standard ARD.

```{r ch5-ex2, exercise=TRUE, exercise.lines=12}
# Convert fit_logistic results to an ARD using ard_regression()
# Hint: pass fit_logistic, set exponentiate = TRUE and conf.int = TRUE

ard_reg_results <- ard_regression(
  x = ___,
  exponentiate = ___,
  conf.int = ___
)

# Look at the results
ard_reg_results %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_fmt)
```

```{r ch5-ex2-hint-1}
# Set x = fit_logistic, exponentiate = TRUE, conf.int = TRUE
```

```{r ch5-ex2-solution}
ard_reg_results <- ard_regression(
  x = fit_logistic,
  exponentiate = TRUE,
  conf.int = TRUE
)
ard_reg_results %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_fmt)
```

```{r ch5-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result) && "estimate" %in% .result$stat_name, "Correct! cardx::ard_regression() created a complete, standard ARD representation of the regression — one variable per heading, reference rows added, and `stat_fmt` populated by `apply_fmt_fun()`."),
  fail_if(~ TRUE, "Check that you specify x = fit_logistic, exponentiate = TRUE, and conf.int = TRUE in ard_regression() and format the output.")
)
```

> **Spot the reference row:** scroll the result and look for `TRT01A` with `Placebo` in `variable_level` — its `estimate` is 1 and `p.value` is `NA`. That is the synthetic reference row that `broom.helpers` added behind the scenes. The SAS equivalent would be a DATA-step `INSERT` you'd have to write yourself after `ODS OUTPUT`.

---

## Exercise 5.3 — Goodness-of-Fit with `broom::glance()`

Where `tidy()` gives one row per coefficient, `glance()` gives exactly one row per model — the fit statistics that belong in a regression-table footer (n, AIC, BIC, deviance, log-likelihood).

```{r ch5-ex3, exercise=TRUE, exercise.lines=8}
# Get the one-row model summary for fit_logistic.
# Hint: glance(fit_logistic)

fit_logistic %>%
  ___()
```

```{r ch5-ex3-hint-1}
glance(fit_logistic)
```

```{r ch5-ex3-solution}
glance(fit_logistic)
```

```{r ch5-ex3-check}
grade_result(
  pass_if(~ is.data.frame(.result) && nrow(.result) == 1 && "AIC" %in% names(.result), "Right — `glance()` always returns exactly one row. For a glm you get AIC, BIC, null.deviance, deviance, df.null, df.residual, logLik, and nobs."),
  fail_if(~ TRUE, "Call glance(fit_logistic). The result should be a one-row tibble.")
)
```

> **SAS analog:** these are the same numbers SAS writes to `ODS OUTPUT FitStatistics=` for `PROC LOGISTIC`: AIC, SC (= BIC), -2 Log L. `glance()` is one tidy row containing all of them; the SAS dataset is a long stacked dataset with one row per statistic — same numbers, different shape.

---

## Exercise 5.4 — Tidy the same model with `parameters::model_parameters()`

`{parameters}` from easystats is an alternative to `{broom}`. It covers more model classes and has slightly different column names. The output is a flat data frame either way — you just have to know which column vocabulary is which.

```{r ch5-ex4, exercise=TRUE, exercise.lines=8}
# Use parameters::model_parameters() to tidy fit_logistic with exponentiated coefficients.
# Hint: model_parameters(fit_logistic, exponentiate = TRUE)

fit_logistic %>%
  ___(exponentiate = ___)
```

```{r ch5-ex4-hint-1}
model_parameters(fit_logistic, exponentiate = TRUE)
```

```{r ch5-ex4-solution}
model_parameters(fit_logistic, exponentiate = TRUE)
```

```{r ch5-ex4-check}
grade_result(
  pass_if(~ inherits(.result, "data.frame") && "Parameter" %in% names(.result), "Yes — `model_parameters()` returns a data frame whose column names are `Parameter`, `Coefficient` (or `Odds Ratio`), `SE`, `CI_low`, `CI_high`, `z`, `p`. Same numbers, different vocabulary from `broom::tidy()`."),
  fail_if(~ TRUE, "Call model_parameters(fit_logistic, exponentiate = TRUE).")
)
```

> **When to use which:** for the standard model families (`lm`, `glm`, `coxph`, `lme4`) you can pick freely — pass either as `tidy_fun` to `ard_regression()`. For unusual models (`brms`, `rstanarm`, `mmrm`, etc.) `parameters` often supports model classes that `broom` does not.

---

## Exercise 5.5 — Linear Regression (`lm`) — the `PROC REG` Analog

The pre-fitted `fit_lm` is a linear regression of `AGE` on `TRT01A + SEX`. Linear regression returns coefficients on the response scale already — **do not exponentiate**. Generate an ARD with confidence intervals.

```{r ch5-ex5, exercise=TRUE, exercise.lines=12}
# Build an ARD from the linear model fit_lm.
# Hint: ard_regression(fit_lm, conf.int = TRUE)  -- no exponentiate for lm.

ard_lm <- ard_regression(
  x = ___,
  conf.int = ___
)

ard_lm %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt)
```

```{r ch5-ex5-hint-1}
ard_regression(x = fit_lm, conf.int = TRUE)
```

```{r ch5-ex5-solution}
ard_lm <- ard_regression(
  x = fit_lm,
  conf.int = TRUE
)
ard_lm %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt)
```

```{r ch5-ex5-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "estimate" %in% .result$stat_name && "conf.low" %in% .result$stat_name, "Perfect — same ARD shape as the logistic case, but the coefficients are on the response scale (age in years per unit predictor). No exponentiation."),
  fail_if(~ TRUE, "Call ard_regression(x = fit_lm, conf.int = TRUE).")
)
```

> **SAS analog:** `PROC REG DATA=adsl; CLASS TRT01A SEX; MODEL AGE = TRT01A SEX / CLB;` or the same model in `PROC GLM`. The "Parameter Estimates" table with the `CLB` option gives you the same numbers `ard_regression()` returns here.

---

## Exercise 5.6 — Cox Proportional Hazards (`coxph`) — the `PROC PHREG` Analog

`fit_cox` is pre-fitted in your environment: `coxph(Surv(AVAL, 1 - CNSR) ~ TRT01A + AGE, data = adtte_os)`. Cox returns coefficients on the **log-hazard** scale, so you want `exponentiate = TRUE` to get Hazard Ratios.

```{r ch5-ex6, exercise=TRUE, exercise.lines=12}
# Build an ARD from the Cox fit, with HRs and 95% CIs.

ard_cox <- ard_regression(
  x = ___,
  exponentiate = ___,
  conf.int = ___
)

ard_cox %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt)
```

```{r ch5-ex6-hint-1}
ard_regression(x = fit_cox, exponentiate = TRUE, conf.int = TRUE)
```

```{r ch5-ex6-solution}
ard_cox <- ard_regression(
  x = fit_cox,
  exponentiate = TRUE,
  conf.int = TRUE
)
ard_cox %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, variable_level, stat_name, stat_fmt)
```

```{r ch5-ex6-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "estimate" %in% .result$stat_name, "Great — the same ARD shape works for Cox models. The `estimate` rows now hold Hazard Ratios (because `exponentiate = TRUE`), and `TRT01A` Placebo appears with `estimate = 1` as the reference."),
  fail_if(~ TRUE, "Call ard_regression(x = fit_cox, exponentiate = TRUE, conf.int = TRUE).")
)
```

> **SAS analog:** `PROC PHREG DATA=adtte_os; CLASS TRT01A (REF='Placebo'); MODEL AVAL*CNSR(1) = TRT01A AGE / RL;` — the "Hazard Ratios" table is what `ard_regression()` returns above, including the reference row for Placebo.

---

## Exercise 5.7 — `ard_regression()` vs. `ard_regression_basic()`

`ard_regression_basic()` skips the `broom.helpers` layer: no reference rows, no variable labels, no grouping. Compare the two side by side on the same `fit_logistic` and observe which rows disappear.

```{r ch5-ex7, exercise=TRUE, exercise.lines=12}
# Build both ARDs from fit_logistic.
# Hint: ard_regression_basic() takes the same first two arguments.

ard_full  <- ard_regression(x = fit_logistic, exponentiate = TRUE, conf.int = TRUE)
ard_basic <- ___(x = fit_logistic, exponentiate = TRUE, conf.int = TRUE)

# Compare row counts
list(
  rows_full  = nrow(ard_full),
  rows_basic = nrow(ard_basic)
)
```

```{r ch5-ex7-hint-1}
ard_regression_basic(x = fit_logistic, exponentiate = TRUE, conf.int = TRUE)
```

```{r ch5-ex7-solution}
ard_full  <- ard_regression(x = fit_logistic, exponentiate = TRUE, conf.int = TRUE)
ard_basic <- ard_regression_basic(x = fit_logistic, exponentiate = TRUE, conf.int = TRUE)
list(
  rows_full  = nrow(ard_full),
  rows_basic = nrow(ard_basic)
)
```

```{r ch5-ex7-check}
grade_result(
  pass_if(~ is.list(.result) && .result$rows_full > .result$rows_basic, "Exactly — the full `ard_regression()` returns more rows because `broom.helpers` adds the reference row for `Placebo` plus label / grouping metadata. `ard_regression_basic()` skips those and gives you only what `broom::tidy()` produced."),
  fail_if(~ TRUE, "Call ard_regression_basic() with the same arguments and compare nrow() of both ARDs.")
)
```

> **When to use the basic form:** building a forest plot programmatically (you want exactly one row per non-reference term), or feeding the ARD into a custom layout where the reference rows would be more noise than signal. For standard tabular display, prefer the full `ard_regression()`.

---

## Chapter 5 Knowledge Check

```{r ch5-quiz, echo=FALSE}
quiz(
  caption = "Chapter 5 — Check Your Understanding",

  question(
    "What does `broom::tidy()` do?",
    answer("Plots regression diagnostics"),
    answer("Converts complex statistical model coefficient outputs into flat, tidy data frames", correct = TRUE),
    answer("Cleans missing values from datasets"),
    answer("Formats R code syntax"),
    allow_retry = TRUE
  ),

  question(
    "Which package does `{cardx}` use under the hood to standardize outputs from both `{broom}` and `{parameters}` and to add categorical reference rows?",
    answer("ggplot2"),
    answer("broom.helpers", correct = TRUE),
    answer("dplyr"),
    answer("tidyr"),
    allow_retry = TRUE
  ),

  question(
    "How do you get Odds Ratios (exponentiated coefficients) from a logistic model in `ard_regression()`?",
    answer("Set odds_ratio = TRUE"),
    answer("Set exponentiate = TRUE", correct = TRUE),
    answer("Pass exp = TRUE"),
    answer("Perform the exponentiation manually after generating the ARD"),
    allow_retry = TRUE
  ),

  question(
    "Which `{broom}` function returns exactly one row per model containing fit statistics like AIC, BIC, and deviance?",
    answer("`tidy()`"),
    answer("`glance()`", correct = TRUE),
    answer("`augment()`"),
    answer("`summary()`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{broom}` function appends per-observation diagnostics (`.fitted`, `.resid`, `.hat`, `.cooksd`) to the original data?",
    answer("`tidy()`"),
    answer("`glance()`"),
    answer("`augment()`", correct = TRUE),
    answer("`model_parameters()`"),
    allow_retry = TRUE
  ),

  question(
    "Which SAS PROC is the closest analog to `survival::coxph()`?",
    answer("`PROC LOGISTIC`"),
    answer("`PROC LIFETEST`"),
    answer("`PROC PHREG`", correct = TRUE),
    answer("`PROC GLM`"),
    allow_retry = TRUE
  ),

  question(
    "Which SAS PROC is the closest analog to `glm(..., family = poisson)`?",
    answer("`PROC LOGISTIC`"),
    answer("`PROC GENMOD DIST=POISSON`", correct = TRUE),
    answer("`PROC REG`"),
    answer("`PROC GLIMMIX`"),
    allow_retry = TRUE
  ),

  question(
    "Why should you never exponentiate the `estimate` of an `lm()` fit?",
    answer("Because `lm()` returns coefficients on the response scale already — exponentiating destroys the interpretation.", correct = TRUE),
    answer("Because R prevents it with an error."),
    answer("Because the CIs would no longer match — but the estimate itself would be fine."),
    answer("There is no reason; you can do it freely."),
    allow_retry = TRUE
  ),

  question(
    "What is the practical difference between `ard_regression()` and `ard_regression_basic()`?",
    answer("They are aliases; the output is identical."),
    answer("The basic version runs `broom::tidy()` only; the full version layers `broom.helpers::tidy_plus_plus()` on top to add reference rows, labels, and term grouping.", correct = TRUE),
    answer("The basic version is for `lm`, the full version is for `glm`."),
    answer("The basic version drops the CI columns."),
    allow_retry = TRUE
  ),

  question(
    "Your logistic model uses a 3-level `TRT01A` factor with `Placebo` as the reference. How many rows for `TRT01A` does `ard_regression()` produce by default?",
    answer("1 row (just the omnibus test)"),
    answer("2 rows (the two non-reference levels)"),
    answer("3 rows — one per level including the reference row", correct = TRUE),
    answer("0 rows; the reference is dropped entirely"),
    allow_retry = TRUE
  ),

  question(
    "Which argument do you pass to `ard_regression()` if you want to use `parameters::model_parameters()` instead of the default broom-based tidier?",
    answer("`tidier = parameters::model_parameters`"),
    answer("`tidy_fun = parameters::model_parameters`", correct = TRUE),
    answer("`engine = \"parameters\"`"),
    answer("There is no way; the tidier is hard-coded."),
    allow_retry = TRUE
  ),

  question(
    "Why might `glm()` confidence intervals run slowly on a very large dataset, and what is the fix?",
    answer("R is single-threaded; there is no fix."),
    answer("The default CI is profile-likelihood, which refits the model per term. Switch to Wald via `conf.method = \"wald\"` for speed.", correct = TRUE),
    answer("Profile CIs require an Internet connection."),
    answer("The slowness is a bug — file an issue."),
    allow_retry = TRUE
  )
)
```

---

## Chapter 5 Summary {.congratulations}

You now know how to:

- Explain why **model tidying** exists and what it solves — every R modeling package returns a nested list with class-specific slot names; tidying standardizes them into a small flat tibble vocabulary.
- Read the **anatomy** of a fitted model object — `coefficients`, `vcov`, `residuals`, `family`, `terms` — and where each tidier reaches in to extract.
- Use the **three broom workhorses**:
  - `tidy()` → one row per coefficient.
  - `glance()` → one row per model (fit statistics).
  - `augment()` → one row per observation (predictions and diagnostics).
- Use **`parameters::model_parameters()`** as an alternative tidier with broader model coverage and richer print methods.
- Use **`broom.helpers::tidy_plus_plus()`** to add variable labels, reference rows, term grouping, and contrast metadata — the layer that makes tidy output *clinical-table ready*.
- Generate ARDs from any fitted model with **`cardx::ard_regression()`**, and understand the difference between it and **`ard_regression_basic()`**.
- Map every common SAS modeling PROC to its R counterpart and feed the fitted object through `ard_regression()`:
  - `PROC REG` / `PROC GLM` → `lm()`
  - `PROC LOGISTIC` → `glm(family = binomial)` / `MASS::polr()` / `nnet::multinom()`
  - `PROC GENMOD DIST=POISSON` → `glm(family = poisson)` / `MASS::glm.nb()`
  - `PROC PHREG` → `survival::coxph()`
  - `PROC LIFEREG` → `survival::survreg()`
  - `PROC MIXED` → `lme4::lmer()` / `nlme::gls()`
  - `PROC GLIMMIX` → `lme4::glmer()`
  - `PROC GENMOD REPEATED` → `geepack::geeglm()`
- Choose **`exponentiate = TRUE`** correctly: yes for logistic/Poisson/Cox, no for linear regression.
- Recognize and add **reference rows** so categorical factors display "Placebo  1.00 (ref)" alongside the non-reference levels.
- Pick the right **CI method** (Wald vs. profile likelihood vs. robust), and pass a custom **`tidy_fun`** when defaults are not enough.
- **Format** regression ARDs with the same `fmt_fun` / `apply_fmt_fun()` / `update_ard_fmt_fun()` machinery you learned for descriptive ARDs.
- **Bind** regression ARDs onto descriptive and inferential ARDs to feed a single end-to-end layout call.
- Diagnose the **common pitfalls** — forgetting `exponentiate`, missing reference rows, missing variable labels, slow profile CIs, and absent `lme4` p-values.

**Next:** Chapter 6 — Consuming ARD (rendering publication-ready tables with `{gtsummary}::tbl_ard_summary()`).
