---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 3: Stacking and Formatting"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Master combining multiple ARD objects using bind_ard and customizing decimal formats with fmt_fun in cards.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
```

---

## Chapter 3: Stacking and Formatting

### 3.1 Why Stacking — One Table, Many ARDs

A real demographics table is not the output of one function. It is the output of many small `{cards}` calls — one for the header N, one for AGE, one for SEX, one for RACE, possibly one for safety flags, possibly one for missingness — all stitched together into a single, layout-ready dataset.

```
   ard_total_n()         -> N=86, N=84, N=84       (header row)
   ard_continuous(AGE)   -> n, mean, sd, median... (Age block)
   ard_continuous(BMI)   -> n, mean, sd, median... (BMI block)
   ard_categorical(SEX)  -> n, p per level         (Sex block)
   ard_categorical(RACE) -> n, p per level         (Race block)
   ard_dichotomous(SAFFL)-> n, p for "Y"           (Safety pop row)
                |
                |  bind_ard()  <-- this chapter
                v
   +-----------------------------------------+
   |  ONE STACKED ARD                        |
   |  - same column schema across all rows   |
   |  - duplicates checked                   |
   |  - ready for a layout tool to consume   |
   +-----------------------------------------+
```

> **SAS analog:** writing several PROCs that each produce an `ODS OUTPUT` dataset and then `SET`-ing those datasets together in a DATA step. Stacking is the `{cards}` formalization of that pattern, with the added guarantee that the column schema is identical for every contributor.

The chapter splits into two halves: **stacking** (sections 3.2–3.5) and **formatting** (sections 3.6–3.10).

---

### 3.2 `bind_ard()` — More Than Just `bind_rows()`

Why not just call `dplyr::bind_rows()`? Because an ARD has invariants that a generic row-binder will not protect:

- The same `(group1_level, variable, variable_grp_level, stat_name)` quadruple should never appear twice — if it does, downstream tools cannot decide which value to display, and the silent duplication leads to visually-identical-but-numerically-divergent QC findings.
- All input data frames must carry the same standard columns; mismatched schemas should not be silently coerced.
- The `stat` list-column must survive the bind without being flattened or coerced.

`bind_ard()` enforces all three. It is a deliberately strict, validating combiner — not a permissive concatenation.

```r
ard_demog <- bind_ard(
  ard_total_n_trt,
  ard_age,
  ard_sex,
  ard_race,
  ard_saffl
)
```

The result is one tall ARD with all the rows needed to render a full demographics section. Same `group1_level` values everywhere (because every input was grouped by `TRT01A`), same column names, same data types in each column.

> **SAS analog:** `bind_ard()` is the equivalent of a DATA-step `SET` that also runs `PROC COMPARE` invariants on its inputs before producing the union. You never get a free `SET` of incompatible structures — and that strictness is the point.

---

### 3.3 What Counts as a Duplicate

A duplicate is *not* "the same number appears twice." A duplicate is "the same row identity appears twice." The row identity is the combination of:

```
   group1 + group1_level
 + group2 + group2_level   (if present)
 + variable
 + variable_grp + variable_grp_level
 + stat_name
```

If those columns match across two rows, those two rows are claiming to be the same statistic, computed two different ways. `bind_ard()` will flag it.

Common ways the situation arises in practice:

| Cause                                                              | Fix                                                                   |
|--------------------------------------------------------------------|-----------------------------------------------------------------------|
| You called `ard_continuous(AGE)` twice with different formatting   | Drop one call; configure formatting once in the surviving call.       |
| You bound an overall ARD into a per-arm ARD where both used `TRT01A` | Stratify only the per-arm one; produce the overall via a separate call without `by`. |
| You mixed `ard_categorical(SEX)` and `ard_dichotomous(SEX)`        | Pick one — they overlap on the same `(variable, level)` keys.         |

> **SAS analog:** the same defensive instinct as running `PROC SORT NODUPKEY` after a `SET` to make sure no `(BY, _STAT_)` row appears twice.

---

### 3.4 `ard_stack()` — One Call for Many ARDs

When you find yourself writing the same boilerplate four times — same `data`, same `by` — `ard_stack()` collapses it into one call:

```r
ard_demog <- ard_stack(
  data       = adsl,
  ard_continuous(variables = c(AGE, BMIBL)),
  ard_categorical(variables = c(SEX, RACE)),
  .by        = TRT01A,
  .total_n   = TRUE,
  .missing   = TRUE,
  .attributes = TRUE
)
```

Notice the leading dots on the wrapper-level arguments (`.by`, `.total_n`, `.missing`, `.attributes`, `.overall`). They distinguish wrapper-controlled options from the `by`/`variables` that would otherwise be re-passed into each inner function. The result is one stacked ARD, validated by the same `bind_ard()` machinery under the hood.

#### 3.4.1 What each wrapper flag does

| Argument       | When `TRUE`...                                                                |
|----------------|-------------------------------------------------------------------------------|
| `.by`          | Stratification variable passed down to every inner call.                       |
| `.total_n`     | Automatically adds an `ard_total_n()` block — no separate call needed.         |
| `.missing`     | Automatically adds an `ard_missing()` block per variable.                      |
| `.overall`     | Adds a no-stratification block — the "Total" column data, alongside per-arm.   |
| `.attributes`  | Picks up variable labels stored in the input frame (e.g., `label(AGE)="Age, years"`) and uses them for `variable_label`. |

Use `ard_stack()` for demographics-style tables where the per-call boilerplate would dominate. Use `bind_ard()` directly when the inputs are heterogeneous (different `by` variables, different filters, different populations).

> **SAS analog:** `ard_stack()` is the macro that wraps your `%MACRO build_demog_arm`. It is convenience around a well-defined pattern; the underlying tools still work on their own.

---

### 3.5 Choosing Between `bind_ard()` and `ard_stack()`

```
                Is every inner call working on the
                same data, same by, same population?
                          /          \
                        YES           NO
                         |             |
                         v             v
                   ard_stack()      bind_ard()
                  (less typing,    (full control,
                   one call)        heterogeneous inputs)
```

A typical study program uses both — `ard_stack()` for the demographics section, then `bind_ard()` to join the demographics ARD with a separately-built efficacy ARD that has a different `by` (e.g., arm × visit instead of just arm).

---

## Exercise 3.1 — Combine Multiple ARD Outputs

Calculate continuous summaries for `AGE` and categorical summaries for `SEX`, and combine them using `bind_ard()`.

``` {r ch3-ex1, exercise=TRUE, exercise.lines=15}
# 1. Generate continuous ARD for AGE
ard_age <- ard_continuous(adsl, variables = AGE, by = TRT01A)

# 2. Generate categorical ARD for SEX
ard_sex <- ard_categorical(adsl, variables = SEX, by = TRT01A)

# 3. Bind them together into one demographic ARD
# Hint: use bind_ard()

ard_demog <- ___(ard_age, ard_sex)
ard_demog
```

``` {r ch3-ex1-hint-1}
# Call bind_ard with your two ARD data frames
bind_ard(ard_age, ard_sex)
```

``` {r ch3-ex1-solution}
bind_ard(ard_age, ard_sex)
```

``` {r ch3-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("AGE", "SEX") %in% .result$variable), "Correct! bind_ard stacked both tables. Downstream tools can now render a unified demographics table from this single ARD object."),
  fail_if(~ TRUE, "Ensure you are calling bind_ard(ard_age, ard_sex).")
)
```

> **Inspect:** the `variable` column now contains both `"AGE"` and `"SEX"`. The `context` column tells you which `{cards}` function produced each block. The schema is identical across both — that uniformity is the reason `bind_ard()` could combine them safely.

---

### 3.6 Why Formatting Is a Separate Step

In SAS you typically attach a `FORMAT` to a variable and forget about it; the format follows the variable into every output. In ARD the rule is the opposite: formatting is an *explicit*, recorded transformation from `stat` to `stat_fmt`. Both versions persist side-by-side, and the formatting rules live on the dataset row, not on a global variable attribute.

```
                stat               fmt_fun             stat_fmt
              (numeric)          (decimals/fn)        (character)
              75.21    -------->   1 decimal  ------->  "75.2"
              8.12     -------->   2 decimals ------->  "8.12"
              0.616    -------->  pct, 1 dec  ------->  "61.6%"
              0.00234  -------->  scientific  ------->  "2.3e-3"
                |                                          |
                |                                          |
                +---- both stay in the ARD ----------------+
```

Two payoffs:

1. **Auditability.** A reviewer can re-derive every `stat_fmt` from `stat` because the rule is on the row, not implicit in a SAS proc somewhere.
2. **Re-rendering.** Switching the entire table from one decimal place to two for the means is a single argument change — not a hunt through layout code.

> **SAS analog:** the difference between `PUT(AGE, 5.1)` evaluated once at display time vs. storing both the raw `AGE` and the formatted string `"75.2"` as two persistent columns in the dataset. ARD prefers the latter.

---

### 3.7 The `fmt_fun` Specification

`fmt_fun` is a nested list. The outer level is keyed by `variable`, and the inner level is keyed by `stat_name`. Three forms are accepted at the inner level.

| Form                                | Meaning                                                                  |
|-------------------------------------|--------------------------------------------------------------------------|
| `mean = 1`                          | "Round mean to 1 decimal place." Shorthand.                              |
| `mean = ~ format(round(.x, 1), nsmall = 1)` | Plug in any custom R function; receives the raw value as `.x`.   |
| `p = label_percent(accuracy = 0.1)` | Pre-built scales-style formatter that returns a character string.        |

A complete spec for a typical AGE block:

```r
fmt_fun = list(
  AGE = list(
    n        = 0,    # integer
    mean     = 1,    # one decimal
    sd       = 2,    # two decimals
    median   = 1,    # one decimal
    p25      = 0,    # quartiles as integers
    p75      = 0,
    min      = 0,
    max      = 0
  )
)
```

Multiple variables get multiple top-level keys:

```r
fmt_fun = list(
  AGE    = list(n = 0, mean = 1, sd = 2),
  BMIBL  = list(n = 0, mean = 2, sd = 2)
)
```

> **SAS analog:** the inner list is the moral equivalent of attaching a `FORMAT` to a SAS variable, but the format is keyed by *both* the variable and the *statistic* so that within the AGE block the mean rounds to 1 dp and the SD rounds to 2 dp — something a global SAS `FORMAT AGE 5.1;` cannot express.

---

### 3.8 Clinical Decimal Conventions — A Survival Cheat Sheet

These conventions are nearly universal across pharma SAPs. Memorize them; downstream reviewers will assume them.

| Statistic                        | Typical decimals | Example                |
|----------------------------------|------------------|------------------------|
| Counts (`n`, `N`)                | 0                | `86`                   |
| Percentages (`p`)                | 1                | `61.6%`                |
| Mean                             | 1                | `75.2`                 |
| Standard deviation (`sd`)        | 2                | `8.12`                 |
| Median, quartiles                | 1                | `76.0`, `(70, 81)`     |
| Min, Max                         | 0 (integers) or 1 (real-valued)  | `52, 89` or `52.0, 89.0` |
| p-value                          | 3-4 with `<0.001` threshold | `0.234`, `<0.001` |
| Hazard ratio, odds ratio         | 2-3              | `0.78`                 |
| Confidence interval bounds       | match the point estimate | `(0.65, 0.95)` |

The rule is *one more decimal than the precision the reader cares about, and never more than the data justifies.* A weight measured to the nearest kilogram should not produce a mean reported to four decimals.

---

### 3.9 `apply_fmt_fun()` — The Formatting Pipeline

Specifying `fmt_fun` is only half the work. To actually fill in the `stat_fmt` column, you pipe the ARD through `apply_fmt_fun()`:

```
   ard_continuous(... fmt_fun = list(...))
                |
                |  produces stat (raw) and the formatting rule
                v
        apply_fmt_fun()
                |
                |  evaluates each rule, writes stat_fmt
                v
   ARD with both stat and stat_fmt populated
```

The two-step design is deliberate. You can:

- Compute once and re-format many times by re-piping through `apply_fmt_fun()` with a different `fmt_fun`.
- Inspect raw `stat` values before formatting (great for QC).
- Stack many partially-formatted ARDs before deciding on a unified formatting pass at the end.

> **SAS analog:** writing a `FORMAT` statement and then running the dataset through `PROC PRINT` is the moment the format actually executes. `apply_fmt_fun()` is the explicit "now apply the format" step.

---

### 3.10 List-Columns and `unlist_ard_columns()`

The `stat` column is a **list-column** because not every statistic returns a single number. A 95% CI returns a length-2 vector. A vector of quantiles returns a vector. By storing each row's value as a list element, `{cards}` keeps the raw R object alive without forcing a coercion.

The same is true of `stat_fmt`. Even though a formatted string is "just text", it lives in a list-column to mirror `stat` exactly.

This is convenient for the *computation* but inconvenient for *display* — many downstream tools (`gt`, `flextable`, `kable`, generic `dplyr::select()` pipes) prefer plain atomic vectors. `unlist_ard_columns()` is the bridge:

```r
ard_demog %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, group1_level, stat_name, stat_fmt)
```

Two behaviors worth knowing:

- **Default behavior** unlists *all* list-columns it finds; passing `columns = stat_fmt` flattens only that one.
- **NULL safety:** standard R `unlist()` silently drops `NULL` elements, which shrinks the row count and produces a misaligned vector. `unlist_ard_columns()` replaces `NULL` with `NA` first, preserving row alignment. This is the kind of bug that wastes hours if you reach for base R.

> **SAS analog:** there is no real one — SAS does not have list-columns. The closest mental model is a variable whose value can be a delimited string of multiple numbers; you would have to `SCAN()` it apart for display. `unlist_ard_columns()` is the `SCAN()` step, except it is safe and standardized.

---

## Exercise 3.2 — Apply Custom Formatting Rules

Calculate continuous summaries for `AGE` grouped by `TRT01A`, but apply custom formatting:
- `n` = 0 decimal places
- `mean` = 1 decimal place
- `sd` = 2 decimal places

``` {r ch3-ex2, exercise=TRUE, exercise.lines=15}
# Generate continuous summaries with custom formatting rules
# Hint: pass a nested list to fmt_fun mapping stat_names to decimal integers

ard_age_fmt <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  fmt_fun = list(
    AGE = list(
      n = ___,
      mean = ___,
      sd = ___
    )
  )
)

# Apply formatting function to add 'stat_fmt' column and select columns
ard_age_fmt %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_fmt)
```

``` {r ch3-ex2-hint-1}
# Set n = 0, mean = 1, sd = 2
# fmt_fun = list(AGE = list(n = 0, mean = 1, sd = 2))
```

``` {r ch3-ex2-solution}
ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  fmt_fun = list(
    AGE = list(
      n = 0,
      mean = 1,
      sd = 2
    )
  )
) %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_fmt)
```

``` {r ch3-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result) && .result$stat_fmt[.result$stat_name == "mean"][[1]] == "74.4", "Awesome! You customized the decimals and applied the formatting function to generate the stat_fmt column."),
  fail_if(~ TRUE, "Double-check your nested list structure in fmt_fun: list(AGE = list(n = 0, mean = 1, sd = 2)) and verify you use apply_fmt_fun() and select(variable, stat_name, stat_fmt).")
)
```

> **Notice the pipeline:** `ard_continuous()` produced `stat` and recorded the rule, `apply_fmt_fun()` materialized `stat_fmt`, `unlist_ard_columns()` flattened the list-column so `select()` could behave normally. Four steps, one direction, no surprises.

---

## Exercise 3.3 — One Call With `ard_stack()`

Reproduce the AGE-plus-SEX demographic ARD from Exercise 3.1, but in a single `ard_stack()` call. Also turn on `.total_n` so the header denominator row is included automatically.

``` {r ch3-ex3, exercise=TRUE, exercise.lines=15}
# Hint: pass ard_continuous(variables = AGE) and ard_categorical(variables = SEX)
# directly as arguments to ard_stack(). Use .by for stratification.

ard_demog_stack <- ard_stack(
  data = adsl,
  ard_continuous(variables = ___),
  ard_categorical(variables = ___),
  .by      = TRT01A,
  .total_n = TRUE
)

ard_demog_stack
```

``` {r ch3-ex3-hint-1}
ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by      = TRT01A,
  .total_n = TRUE
)
```

``` {r ch3-ex3-solution}
ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by      = TRT01A,
  .total_n = TRUE
)
```

``` {r ch3-ex3-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("AGE", "SEX") %in% .result$variable) && "N" %in% .result$stat_name, "Excellent! One ard_stack() call produced the AGE block, the SEX block, and the total-N header row — all stacked and validated."),
  fail_if(~ TRUE, "Pass AGE to ard_continuous() and SEX to ard_categorical(), set .by = TRT01A and .total_n = TRUE.")
)
```

> **Compare with Exercise 3.1:** same rows, fewer lines, automatic total-N. This is the typical production pattern.

---

## Chapter 3 Knowledge Check

``` {r ch3-quiz, echo=FALSE}
quiz(
  caption = "Chapter 3 — Check Your Understanding",

  question(
    "Which function in `{cards}` is used to combine multiple ARD data frames vertically?",
    answer("dplyr::bind_rows()"),
    answer("bind_ard()", correct = TRUE),
    answer("stack_ard()"),
    answer("ard_merge()"),
    allow_retry = TRUE
  ),

  question(
    "What happens by default if `bind_ard()` detects duplicate statistic rows across inputs?",
    answer("It silently ignores the duplicates"),
    answer("It throws an error/warning to prevent inconsistent results", correct = TRUE),
    answer("It merges them by calculating the average"),
    answer("It deletes the entire dataset"),
    allow_retry = TRUE
  ),

  question(
    "In which column does `{cards}` store the rounded, formatted display string (after calling `apply_fmt_fun()`)?",
    answer("stat"),
    answer("stat_fmt", correct = TRUE),
    answer("stat_label"),
    answer("context"),
    allow_retry = TRUE
  ),

  question(
    "Which combination of ARD columns defines a 'duplicate' row that `bind_ard()` will reject?",
    answer("Any two rows with the same `stat_fmt` string."),
    answer("Any two rows with the same `variable` and `group1_level`."),
    answer("Any two rows with the same `group1_level`, `variable`, `variable_grp_level`, and `stat_name`.", correct = TRUE),
    answer("Any two rows produced by the same `{cards}` function."),
    allow_retry = TRUE
  ),

  question(
    "When should you reach for `ard_stack()` instead of `bind_ard()`?",
    answer("When the inner calls share the same `data`, `by`, and population — `ard_stack()` removes the boilerplate.", correct = TRUE),
    answer("When you have a single `ard_continuous()` call to make."),
    answer("When the inputs come from different studies."),
    answer("Never — `bind_ard()` is always preferred."),
    allow_retry = TRUE
  ),

  question(
    "Why does `{cards}` store statistics in list-columns instead of plain numeric vectors?",
    answer("To make the code look more functional."),
    answer("Because some statistics (CIs, quantile vectors) return more than one value, and list-columns preserve that shape.", correct = TRUE),
    answer("Because base R requires it for `bind_rows()`."),
    answer("To stop users from running calculations on the raw `stat` values."),
    allow_retry = TRUE
  ),

  question(
    "Why use `unlist_ard_columns()` rather than base R `unlist()` to flatten `stat_fmt`?",
    answer("It is faster."),
    answer("It replaces `NULL` elements with `NA`, preserving row alignment instead of silently dropping rows.", correct = TRUE),
    answer("It is a syntactic alias — they behave identically."),
    answer("Because `unlist()` is deprecated."),
    allow_retry = TRUE
  ),

  question(
    "Inside `fmt_fun = list(AGE = list(mean = 1, sd = 2))`, what does the inner number represent?",
    answer("The number of statistic rows to keep."),
    answer("The number of decimal places used when formatting that statistic.", correct = TRUE),
    answer("The number of treatment arms."),
    answer("The position of the statistic in the table."),
    allow_retry = TRUE
  )
)
```

---

## Chapter 3 Summary {.congratulations}

You now know:

- Why stacking exists: a real clinical table is the union of many small ARDs, and the union has to be safe.
- How to combine ARDs with **`bind_ard()`** and what invariants it enforces (schema, duplicates, list-column preservation).
- What counts as a duplicate row in an ARD and how to avoid producing duplicates in the first place.
- How to collapse many `{cards}` calls into one with **`ard_stack()`**, including the meaning of its dotted arguments (`.by`, `.total_n`, `.missing`, `.overall`, `.attributes`).
- When to prefer `ard_stack()` over `bind_ard()` and vice versa.
- Why formatting is a separate, explicit step in the ARD workflow — auditability and re-rendering.
- How to write a `fmt_fun` specification keyed by variable and statistic, using integer-shorthand or full functions.
- The clinical decimal conventions reviewers will expect.
- The **`apply_fmt_fun()`** pipeline that materializes `stat_fmt` from `stat`.
- Why `stat` and `stat_fmt` are list-columns and how to flatten them safely with **`unlist_ard_columns()`**.

**Next:** Chapter 4 — Extended Stats with `{cardx}` (hypothesis tests, Kaplan-Meier curves, and other inferential outputs that follow the same ARD shape).
