---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 5: Models, broom, and parameters"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Deep dive into statistical model tidying. Understand the roles of broom, parameters, and cardx::ard_regression.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(broom)
library(parameters)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")

# Pre-fit model for use in exercises
fit_logistic <- glm(
  COMPLFL ~ TRT01A + AGE, 
  data = adsl %>% mutate(COMPLFL = ifelse(COMPLFL == "Y", 1, 0)), 
  family = binomial()
)
```

---

## Chapter 5: Models, broom, and parameters

### 5.1 R Model Extraction and the Tidier Ecosystem

Statistical models in R are complex, deeply nested lists containing information about coefficients, residuals, convergence, and family. Slicing these lists manually is tedious. 

The R community developed the **Tidier Ecosystem** to solve this:
- **`{broom}`**: Provides `tidy()`, which converts a model's coefficients into a standard tibble with columns like `term`, `estimate`, and `p.value`.
- **`{parameters}`**: A robust alternative from the `{easystats}` suite that extracts detailed parameter tables (via `model_parameters()`) for an even wider range of models.
- **`{broom.helpers}`**: Acts as an abstraction layer to standardize column headers and add reference levels for clinical reporting.
- **`{cardx}`**: Leverages these packages to generate regression ARDs using `ard_regression()`.

---

## Exercise 5.1 — Tidy a Logistic Regression Model manually

The model `fit_logistic` (Response flag `COMPLFL` vs. Treatment and Age) is already fitted in your environment. Let's use `broom::tidy()` to flatten its outputs into a clean data frame with exponentiated coefficients (to get Odds Ratios) and confidence intervals.

```{r ch5-ex1, exercise=TRUE, exercise.lines=10}
# Tidy the pre-loaded fit_logistic model
# Hint: call tidy() with conf.int = TRUE and exponentiate = TRUE

fit_logistic %>%
  ___(conf.int = ___, exponentiate = ___)
```

```{r ch5-ex1-hint-1}
# Call tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE)
```

```{r ch5-ex1-solution}
tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE)
```

```{r ch5-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "conf.low" %in% names(.result), "Excellent! broom::tidy() converted the model to a flat tibble, providing Odds Ratios and Confidence Intervals in standard columns."),
  fail_if(~ TRUE, "Try calling tidy(fit_logistic, conf.int = TRUE, exponentiate = TRUE).")
)
```

---

## Exercise 5.2 — Generate a Regression ARD using `ard_regression`

Instead of tidying manually and mapping results to an ARD schema, we can pass our model directly to `cardx::ard_regression()`. This will run the tidier, add categorical reference rows, and structure everything as a standard ARD data frame.

```{r ch5-ex2, exercise=TRUE, exercise.lines=12}
# Convert fit_logistic results to an ARD using ard_regression()
# Hint: Pass fit_logistic, set exponentiate = TRUE and conf.int = TRUE

ard_reg_results <- ard_regression(
  x = ___,
  exponentiate = ___,
  conf.int = ___
)

# Look at the results
ard_reg_results %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_fmt)
```

```{r ch5-ex2-hint-1}
# Set x = fit_logistic, exponentiate = TRUE, conf.int = TRUE
```

```{r ch5-ex2-solution}
ard_reg_results <- ard_regression(
  x = fit_logistic,
  exponentiate = TRUE,
  conf.int = TRUE
)
ard_reg_results %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_fmt)
```

```{r ch5-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result) && "estimate" %in% .result$stat_name, "Correct! cardx::ard_regression() created a complete, standard ARD representation of the regression model, incorporating reference rows and formatting variables."),
  fail_if(~ TRUE, "Check that you specify x = fit_logistic, exponentiate = TRUE, and conf.int = TRUE in ard_regression() and format the output.")
)
```

---

## Chapter 5 Knowledge Check

```{r ch5-quiz, echo=FALSE}
quiz(
  caption = "Chapter 5 — Check Your Understanding",

  question(
    "What does `broom::tidy()` do?",
    answer("Plots regression diagnostics"),
    answer("Converts complex statistical model coefficient outputs into flat, tidy data frames", correct = TRUE),
    answer("Cleans missing values from datasets"),
    answer("Formats R code syntax"),
    allow_retry = TRUE
  ),

  question(
    "Which package does `{cardx}` use under the hood to standardize outputs from both `{broom}` and `{parameters}`?",
    answer("ggplot2"),
    answer("broom.helpers", correct = TRUE),
    answer("dplyr"),
    answer("tidyr"),
    allow_retry = TRUE
  ),

  question(
    "How do you get Odds Ratios (exponentiated coefficients) from a logistic model in `ard_regression()`?",
    answer("Set odds_ratio = TRUE"),
    answer("Set exponentiate = TRUE", correct = TRUE),
    answer("Pass exp = TRUE"),
    answer("Perform the exponentiation manually after generating the ARD"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 5 Summary {.congratulations}

You now understand:
- The role of **model tidying** in flattening nested statistical objects.
- How **`{broom}`** (`tidy()`) and **`{parameters}`** (`model_parameters()`) extract model estimates.
- How **`{broom.helpers}`** standardizes coefficient columns and adds categorical reference rows.
- How to generate regression ARDs using **`cardx::ard_regression()`**.

**Next:** Chapter 6 — Consuming ARD (Rendering tables with `{gtsummary}`)
