---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 4: Extended Stats with cardx"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Explore how cardx extends the ARD framework with functions for t-tests, Fisher's exact tests, and Kaplan-Meier survival models.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(survival)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
```

---

## Chapter 4: Extended Stats with cardx

### 4.1 Why `{cardx}` Exists — A Two-Package Design

`{cards}` deliberately stops at descriptive summaries (counts, means, percentages, missingness). It carries almost no dependencies and never asks R to load a statistical-modeling engine. That keeps it fast, light, and easy to validate.

But a clinical table almost always needs *inferential* statistics too: p-values, confidence intervals, hazard ratios, Kaplan-Meier estimates, ANOVA F-tests. Those calculations require packages like `{stats}`, `{survival}`, `{broom.helpers}`, `{emmeans}`, and others. Loading them all into `{cards}` would defeat the "small core" design.

`{cardx}` ("cards eXtended") is the answer: a separate package that takes the dependencies on those modeling engines and wraps them in the exact same ARD shape you already learned. *Same columns, same `stat`/`stat_fmt` pair, same list-columns, same `apply_fmt_fun()` pipeline.* From a layout engine's point of view, an `{cardx}` row is indistinguishable from a `{cards}` row.

```
                +----------------------+
                |  {cards}  (light)    |
                |  - ard_continuous    |
                |  - ard_categorical   |
                |  - ard_dichotomous   |
                |  - ard_missing       |
                |  - ard_total_n       |
                +-----------+----------+
                            |
                            |  same ARD schema
                            v
                +-----------+----------+
                |  {cardx}  (heavy)    |
                |  - ard_stats_*       |   wraps stats::t.test, etc.
                |  - ard_survival_*    |   wraps survival::survfit
                |  - ard_regression    |   wraps broom.helpers
                |  - ard_emmeans_*     |   wraps emmeans
                |  - ard_effectsize_*  |   wraps effectsize
                +----------------------+
                            |
                            v
              bind_ard() / ard_stack() it
              with descriptive ARDs into
              one layout-ready dataset.
```

> **SAS analog mental model:** in SAS, `PROC FREQ`, `PROC MEANS`, `PROC LIFETEST`, `PROC LOGISTIC`, and `PROC GLM` are all separate procedures with separate keyword grammars and separate `ODS OUTPUT` dataset names. Imagine if every one of them emitted an `ODS OUTPUT` dataset with identical column names, validated by the same dictionary. That is what `{cards}` + `{cardx}` deliver together.

---

### 4.2 The `{cardx}` Function Family — One Stop for Every Common Test

`{cardx}` is organized into prefixes by what it wraps. Recognizing the prefix tells you which kind of statistical engine is doing the work underneath.

| Prefix                  | Engine wrapped                      | Typical use                                     |
|-------------------------|-------------------------------------|-------------------------------------------------|
| `ard_stats_*`           | base R `stats::` functions          | t-test, Wilcoxon, Chi-square, Fisher, ANOVA     |
| `ard_survival_*`        | `survival::`                        | Kaplan-Meier, log-rank, survival differences    |
| `ard_regression*`       | `broom.helpers::tidy_plus_plus()`   | Any model: lm, glm, coxph, lmer, ...            |
| `ard_emmeans_*`         | `emmeans::`                         | LS-means, contrasts, pairwise differences       |
| `ard_effectsize_*`      | `effectsize::`                      | Cohen's d, Hedges' g                            |
| `ard_smd_smd()`         | `smd::`                             | Standardized mean differences (propensity work) |
| `ard_continuous_ci()`   | `stats::`                           | Confidence intervals around descriptive means   |
| `ard_categorical_ci()`  | `stats::` / proportions             | Confidence intervals around proportions         |
| `ard_incidence_rate()`  | exact Poisson                       | Adverse event incidence rates                   |
| `ard_survey_svy*`       | `survey::`                          | Survey-weighted variants of the same tests      |

The next sub-sections walk through the families a SAS programmer is most likely to recognize, each paired with its closest `PROC` analog. The two existing exercises (t-test, KM) are followed by two new ones (Fisher's, log-rank).

---

### 4.3 The Shape of an Inferential ARD Row

Inferential outputs add a few new `stat_name` values that descriptive ARDs do not have:

| `stat_name`     | Meaning                                                       |
|-----------------|---------------------------------------------------------------|
| `estimate`      | The point estimate (mean difference, OR, HR, prob, etc.)      |
| `statistic`     | The test statistic (t, χ², F, Z, log-rank statistic)          |
| `parameter`     | Degrees of freedom (or other model parameter)                 |
| `p.value`       | The p-value                                                   |
| `conf.low` / `conf.high` | Confidence interval bounds                           |
| `method`        | A human-readable description of the test that was run         |

The first five are numeric and live in `stat` as before. The `method` row is *metadata*: a character string describing the test, with `stat_fmt = NULL` because there is nothing to format. That metadata row is what causes the `unlist_ard_columns()` issue you will see in section 4.15 — base R `unlist()` drops `NULL` and your row count silently shrinks.

> **SAS analog:** in the `OUT=` dataset of `PROC TTEST` or in `ODS OUTPUT TTests=` you see exactly these quantities (`tValue`, `DF`, `Probt`, `Estimate`, `LowerCL`, `UpperCL`, `Method`). `{cardx}` standardizes the naming so all tests use the same column vocabulary.

---

### 4.4 Two-Sample Mean Comparisons — t-test and Friends

#### 4.4.1 `ard_stats_t_test()` — the `PROC TTEST` equivalent

Compares the means of one numeric variable across exactly two groups.

```r
ard_stats_t_test(
  data      = adsl_2groups,
  variables = AGE,
  by        = TRT01A
)
```

Returns rows for `estimate` (mean difference), `statistic` (the t value), `parameter` (df), `p.value`, `conf.low`/`conf.high`, and `method`.

> **SAS analog:** `PROC TTEST DATA=adsl_2groups; CLASS TRT01A; VAR AGE; RUN;` — same setup, same outputs, ARD shape instead of two SAS datasets.

If you pass more than two groups to `by`, `ard_stats_t_test()` will error. Pre-filter the data to exactly the two arms you want to compare.

#### 4.4.2 `ard_stats_paired_t_test()` — within-subject mean change

For paired designs (baseline vs. post-treatment within the same subject), the paired variant compares the mean *change* against zero:

```r
ard_stats_paired_t_test(
  data      = adlb_diff,
  variables = CHG,
  by        = AVISIT,
  id        = USUBJID
)
```

> **SAS analog:** `PROC TTEST; PAIRED BASELINE*POST;` — same idea, ARD shape.

#### 4.4.3 `ard_stats_wilcox_test()` — the non-parametric alternative

When normality is in doubt (small sample, skewed labs, count outcomes), use the Wilcoxon rank-sum test:

```r
ard_stats_wilcox_test(
  data      = adsl_2groups,
  variables = AGE,
  by        = TRT01A
)
```

> **SAS analog:** `PROC NPAR1WAY WILCOXON; CLASS TRT01A; VAR AGE; RUN;`.

#### 4.4.4 `ard_stats_t_test_onesample()` — one-sample test

Tests whether the mean of a single variable differs from a fixed value (default zero):

```r
ard_stats_t_test_onesample(
  data      = adlb_chg,
  variables = CHG,
  mu        = 0
)
```

> **SAS analog:** `PROC TTEST; VAR CHG; H0 = 0;`.

---

### 4.5 Multi-Group Mean Comparisons — ANOVA and Kruskal-Wallis

#### 4.5.1 `ard_stats_oneway_test()` — the F-test (Welch-corrected)

When you have three or more arms and want a single overall F-test:

```r
ard_stats_oneway_test(
  data      = adsl,
  variables = AGE,
  by        = TRT01A
)
```

> **SAS analog:** `PROC GLM; CLASS TRT01A; MODEL AGE = TRT01A; RUN;` — but only the omnibus F line, not the full Type-III table.

#### 4.5.2 `ard_stats_aov()` and `ard_stats_anova()` — the full ANOVA table

`ard_stats_aov()` wraps base `aov()` and returns the classical ANOVA table (sums of squares, F, p) for a fitted model. `ard_stats_anova()` accepts an already-fitted model and `tidy()`s its ANOVA decomposition.

> **SAS analog:** `PROC ANOVA;` / `PROC GLM;` Type I/III tables. Same kinds of rows; one row per term in the model, each with its F statistic and p-value.

#### 4.5.3 `ard_stats_kruskal_test()` — non-parametric multi-group

```r
ard_stats_kruskal_test(
  data      = adsl,
  variables = AGE,
  by        = TRT01A
)
```

> **SAS analog:** `PROC NPAR1WAY WILCOXON; CLASS TRT01A; VAR AGE;` — the chi-square approximation it reports under "Kruskal-Wallis Test" is the same statistic.

---

### 4.6 Effect Sizes — Cohen's d, Hedges' g, SMD

Effect sizes quantify *how big* a difference is, independent of sample size. Required by ICH E9(R1) for many summaries and by HTA agencies.

| Function                                  | Effect size                                                   |
|-------------------------------------------|---------------------------------------------------------------|
| `ard_effectsize_cohens_d()`               | Cohen's d (standardized mean difference, two-sample)          |
| `ard_effectsize_hedges_g()`               | Hedges' g (small-sample-corrected Cohen's d)                  |
| `ard_effectsize_paired_cohens_d()`        | Paired Cohen's d                                              |
| `ard_smd_smd()`                           | Standardized mean difference — used in propensity-score balance tables |

> **SAS analog:** typically hand-computed from `PROC TTEST` outputs in SAS. `{cardx}` collapses that into one call that returns the standard ARD row with `estimate`, `conf.low`, `conf.high`, and `method`.

---

### 4.7 Categorical Tests — Chi-Square, Fisher, McNemar

#### 4.7.1 `ard_stats_chisq_test()` — Pearson's χ²

The default test for treatment-arm by response, treatment-arm by sex, AE preferred-term by arm, etc.

```r
ard_stats_chisq_test(
  data      = adsl,
  variables = SEX,
  by        = TRT01A
)
```

> **SAS analog:** `PROC FREQ; TABLES SEX*TRT01A / CHISQ;`. Returns the `statistic` (χ²), `parameter` (df), `p.value`, and `method` rows.

#### 4.7.2 `ard_stats_fisher_test()` — Fisher's exact test

Required for sparse tables (any expected cell count < 5) and the default test in many adverse-event tables. Identical call signature; the engine switches under the hood.

```r
ard_stats_fisher_test(
  data      = adsl,
  variables = SEX,
  by        = TRT01A
)
```

> **SAS analog:** `PROC FREQ; TABLES SEX*TRT01A / FISHER;` or the `EXACT FISHER;` statement in `PROC FREQ`.

#### 4.7.3 `ard_stats_mcnemar_test()` — paired categorical (binary)

For matched / paired binary data (case-control matched on covariates, before/after binary measurements):

```r
ard_stats_mcnemar_test(
  data      = adam_paired,
  variables = c(RESP_PRE, RESP_POST)
)
```

> **SAS analog:** `PROC FREQ; TABLES RESP_PRE*RESP_POST / AGREE;` and reading the McNemar's row.

#### 4.7.4 `ard_stats_prop_test()` — comparing two proportions

A direct test of whether two proportions differ, with confidence interval on the difference:

```r
ard_stats_prop_test(
  data      = adae_summary,
  variables = AE_RESP,
  by        = TRT01A
)
```

> **SAS analog:** `PROC FREQ; TABLES TRT01A*AE_RESP / RISKDIFF;` and the corresponding `Pr(z)` column.

---

### 4.8 Stratified Categorical — Mantel-Haenszel / CMH

#### 4.8.1 `ard_stats_mantelhaen_test()` — the CMH test

The Cochran-Mantel-Haenszel test is the workhorse stratified categorical test in clinical trials (e.g., response rate comparison stratified by region or by baseline severity). The function name in current `{cardx}` is `ard_stats_mantelhaen_test()` — the older alias `ard_cmh_test()` may also appear in older code paths.

```r
ard_stats_mantelhaen_test(
  data      = adsl,
  variables = RESPFL,
  by        = TRT01A,
  strata    = REGION
)
```

> **SAS analog:** `PROC FREQ; TABLES REGION*TRT01A*RESPFL / CMH;` — and reading the Mantel-Haenszel χ² row from `ODS OUTPUT CMH=`. `{cardx}` returns the same `statistic`, `parameter`, `p.value`, `method` rows in standard ARD shape.

The CMH test is the standard answer to "give me a treatment p-value adjusted for the strata I designed the trial around."

---

### 4.9 Proportions, Rates, and Counts

#### 4.9.1 `ard_categorical_ci()` — proportion with confidence interval

Adds CI rows around a binomial proportion (Clopper-Pearson by default; many methods available):

```r
ard_categorical_ci(
  data      = adsl,
  variables = RESPFL,
  by        = TRT01A,
  method    = "exact"
)
```

> **SAS analog:** `PROC FREQ; TABLES RESPFL / BINOMIAL(EXACT);`.

#### 4.9.2 `ard_continuous_ci()` — CI around a mean

Adds `conf.low`/`conf.high` rows for the mean of a continuous variable:

```r
ard_continuous_ci(
  data      = adsl,
  variables = AGE,
  by        = TRT01A
)
```

> **SAS analog:** `PROC MEANS CLM;` — that `LowerCL` and `UpperCL` pair, but in ARD shape.

#### 4.9.3 `ard_incidence_rate()` — exposure-adjusted rate

Adverse-event incidence per 100 person-years (or any denominator unit). Wraps an exact Poisson CI under the hood.

> **SAS analog:** hand-computed from `PROC GENMOD DIST=POISSON` outputs or `PROC FREQ` over person-time. `{cardx}` collapses that into one ARD-shaped call.

#### 4.9.4 `ard_stats_poisson_test()` — Poisson rate test

Compares an observed event count to an expected one under a Poisson assumption.

> **SAS analog:** `PROC GENMOD DIST=POISSON; MODEL events / offset = log_pyears;` and reading the rate p-value.

---

### 4.10 LS-Means and Contrasts via `ard_emmeans_*`

Pre-specified comparisons in mixed-model designs (the staple of MMRM analyses for continuous endpoints) live in `{emmeans}`. `{cardx}` wraps three of its most useful outputs:

| Function                       | Produces                                                             |
|--------------------------------|----------------------------------------------------------------------|
| `ard_emmeans_emmeans()`        | LS-means with standard errors and CIs at each level of a factor.     |
| `ard_emmeans_mean_difference()`| Pairwise LS-mean differences (e.g., active − placebo).               |
| `ard_emmeans_contrast()`       | Arbitrary user-defined linear contrasts.                             |

> **SAS analog:** `LSMEANS TRT01A / CL DIFF;` and `LSMESTIMATE ... ;` statements in `PROC MIXED` / `PROC GLIMMIX`. Same outputs, ARD-shaped rows.

---

### 4.11 Survival Analysis — KM and Log-Rank

#### 4.11.1 `ard_survival_survfit()` — Kaplan-Meier point estimates

KM survival probabilities at specified time points, per treatment arm.

```r
ard_survival_survfit(
  x         = adtte_os,
  y         = Surv(AVAL, 1 - CNSR),
  variables = TRT01A,
  times     = c(6, 12)
)
```

The result has one block of rows per (treatment arm × time point), with `stat_name` values such as `n.risk`, `n.event`, `estimate` (survival probability), `std.error`, `conf.low`, `conf.high`. The time point lives in `variable_level`.

> **SAS analog:** `PROC LIFETEST DATA=adtte_os; TIME AVAL*CNSR(1); STRATA TRT01A; RUN;` — the Survival Estimates table, sliced to the time points you want.

A few notes on the encoding of censorship:

- `Surv(time, event_indicator)` expects **event=1, censored=0**.
- ADaM convention puts that the other way around: `CNSR = 1` means censored.
- Hence `Surv(AVAL, 1 - CNSR)` — flip the indicator before feeding `Surv()`.
- The same trap exists in SAS: `TIME AVAL*CNSR(1);` tells SAS that `CNSR = 1` is the censored value, *not* the event value. Same conceptual flip, expressed differently.

#### 4.11.2 `ard_survival_survdiff()` — log-rank test

The omnibus test of "do the KM curves differ?":

```r
ard_survival_survdiff(
  data      = adtte_os,
  y         = Surv(AVAL, 1 - CNSR),
  variables = TRT01A
)
```

Returns `statistic` (log-rank χ²), `parameter` (df), `p.value`, `method`.

> **SAS analog:** the "Test of Equality over Strata" table from `PROC LIFETEST` — specifically the `Log-Rank` row.

#### 4.11.3 `ard_survival_survfit_diff()` — survival probability difference

Difference in survival probability at a given time point between two arms, with CI:

```r
ard_survival_survfit_diff(
  x         = adtte_os,
  y         = Surv(AVAL, 1 - CNSR),
  variables = TRT01A,
  times     = 12
)
```

> **SAS analog:** there is no direct PROC counterpart; this is typically hand-computed from two `PROC LIFETEST` runs in SAS workflows.

---

### 4.12 Regression Model ARDs — `ard_regression()` Preview

Chapter 5 dives deep into model tidying. The headline: `ard_regression()` accepts an already-fitted model object (`lm`, `glm`, `coxph`, `lmer`, `polr`, ...) and returns an ARD with one block of rows per coefficient — including the standard estimate / SE / statistic / p.value / conf.low / conf.high columns.

```r
fit <- glm(AEFL ~ TRT01A + AGE + SEX, data = adsl, family = binomial)
ard_regression(fit, exponentiate = TRUE, conf.int = TRUE)
```

> **SAS analog:** `PROC LOGISTIC; CLASS TRT01A SEX; MODEL AEFL = TRT01A AGE SEX;` and the "Analysis of Maximum Likelihood Estimates" table → exponentiated to "Odds Ratio Estimates."

---

## Exercise 4.1 — Perform a Two-Sample t-test

Let's compare the mean age between the Placebo and Xanomeline High Dose groups. A t-test compares exactly two groups, so we must filter out the Low Dose group first.

```{r ch4-ex1, exercise=TRUE, exercise.lines=15}
# Filter adsl to Placebo and Xanomeline High Dose only
adsl_2groups <- adsl %>%
  filter(TRT01A %in% c("Placebo", "Xanomeline High Dose"))

# Perform t-test comparing AGE by TRT01A and format as ARD
# Hint: use ard_stats_t_test()

ard_ttest <- ___(
  data = adsl_2groups,
  variables = AGE,
  by = TRT01A
)

# Look at the statistical results
ard_ttest %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(stat_name, stat_label, stat_fmt)
```

```{r ch4-ex1-hint-1}
# Call ard_stats_t_test on adsl_2groups
ard_stats_t_test(adsl_2groups, variables = AGE, by = TRT01A)
```

```{r ch4-ex1-solution}
ard_ttest <- ard_stats_t_test(
  data = adsl_2groups,
  variables = AGE,
  by = TRT01A
)
ard_ttest %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(stat_name, stat_label, stat_fmt)
```

```{r ch4-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result), "Correct! ard_stats_t_test generated standard test rows, and apply_fmt_fun() formatted them into stat_fmt."),
  fail_if(~ TRUE, "Ensure you are using ard_stats_t_test() on adsl_2groups and selecting the correct columns via apply_fmt_fun().")
)
```

> **Read the rows:** `estimate` is the mean difference between the two arms, `statistic` is the t-value, `parameter` is the degrees of freedom, `p.value` is the two-sided p, `conf.low`/`conf.high` are the 95% CI on the difference, and `method` records that this was Welch's t-test. The same six rows would come back from any other `ard_stats_*` test, just with different numbers.

---

## Exercise 4.2 — Calculate Kaplan-Meier survival estimates

In oncology trials, Overall Survival (OS) is a key parameter. We need to calculate KM survival estimates at month time points (6 and 12 months) from the `adtte` dataset.

The `ard_survival_survfit()` function accepts:
- `x`: The input data frame.
- `y`: The survival response object, created with `Surv(time, event)` (or `survival::Surv(time, event)`).
- `variables`: The grouping/strata variable (e.g. `TRT01A`).
- `times`: The numeric vector of time points for estimates.

```{r ch4-ex2, exercise=TRUE, exercise.lines=15}
# Filter for Overall Survival (OS)
adtte_os <- adtte %>% filter(PARAMCD == "OS")

# Calculate survival estimates at Months 6 and 12, grouped by TRT01A
# Hint: use ard_survival_survfit() and specify times = c(6, 12)

ard_km_rates <- ard_survival_survfit(
  x = adtte_os,
  y = Surv(AVAL, 1 - CNSR),
  variables = TRT01A,
  times = ___
)

# View results
ard_km_rates %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable_level, stat_name, stat_fmt)
```

```{r ch4-ex2-hint-1}
# Set times = c(6, 12)
```

```{r ch4-ex2-solution}
ard_km_rates <- ard_survival_survfit(
  x = adtte_os,
  y = Surv(AVAL, 1 - CNSR),
  variables = TRT01A,
  times = c(6, 12)
)
ard_km_rates %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable_level, stat_name, stat_fmt)
```

```{r ch4-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result), "Incredible! You generated Kaplan-Meier survival statistics at specified time intervals. These are mapped to rows where variable_level denotes the timepoint (e.g. 6 or 12)."),
  fail_if(~ TRUE, "Check that you specify times = c(6, 12) in your ard_survival_survfit call and format the output.")
)
```

> **Notice the `variable_level` column:** the time point appears there, not in a separate column. This is the same long-format design pattern as descriptive ARDs — adding a new time point adds rows, not columns.

---

## Exercise 4.3 — Fisher's Exact Test for SEX × TRT01A

Compare the distribution of `SEX` across treatment arms in the safety population using Fisher's exact test — the default test for sparse cells in adverse-event tables.

```{r ch4-ex3, exercise=TRUE, exercise.lines=15}
# Filter to the safety population
adsl_safe <- adsl %>% filter(SAFFL == "Y")

# Run a Fisher's exact test for SEX by TRT01A and format the result.
# Hint: use ard_stats_fisher_test() with the same variables/by pattern.

ard_fisher <- ard_stats_fisher_test(
  data = adsl_safe,
  variables = ___,
  by = ___
)

ard_fisher %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_label, stat_fmt)
```

```{r ch4-ex3-hint-1}
ard_stats_fisher_test(adsl %>% filter(SAFFL == "Y"),
                      variables = SEX, by = TRT01A)
```

```{r ch4-ex3-solution}
ard_fisher <- ard_stats_fisher_test(
  data = adsl %>% filter(SAFFL == "Y"),
  variables = SEX,
  by = TRT01A
)
ard_fisher %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(variable, stat_name, stat_label, stat_fmt)
```

```{r ch4-ex3-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "p.value" %in% .result$stat_name, "Nice! Fisher's exact test ran and the ARD now contains the p.value row alongside estimate and method — same shape as the t-test ARD."),
  fail_if(~ TRUE, "Pass SEX to variables and TRT01A to by inside ard_stats_fisher_test().")
)
```

> **Same shape, different test:** notice the output schema is identical to the t-test exercise above — the only differences are which `stat_name` values appear and which method string is recorded. That uniformity is precisely why a layout engine can render any `{cardx}` output without knowing which test produced it.

---

## Exercise 4.4 — Log-Rank Test on Overall Survival

The log-rank test asks "do the KM curves for the treatment arms differ?". It is the standard omnibus efficacy test for time-to-event endpoints.

```{r ch4-ex4, exercise=TRUE, exercise.lines=15}
# Use the OS subset you built for Exercise 4.2
adtte_os <- adtte %>% filter(PARAMCD == "OS")

# Run the log-rank test for TRT01A.
# Hint: ard_survival_survdiff() takes data=, y=Surv(...), variables=TRT01A.

ard_logrank <- ard_survival_survdiff(
  data = adtte_os,
  y = Surv(AVAL, 1 - CNSR),
  variables = ___
)

ard_logrank %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(stat_name, stat_label, stat_fmt)
```

```{r ch4-ex4-hint-1}
ard_survival_survdiff(
  data = adtte %>% filter(PARAMCD == "OS"),
  y = Surv(AVAL, 1 - CNSR),
  variables = TRT01A
)
```

```{r ch4-ex4-solution}
ard_logrank <- ard_survival_survdiff(
  data = adtte %>% filter(PARAMCD == "OS"),
  y = Surv(AVAL, 1 - CNSR),
  variables = TRT01A
)
ard_logrank %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(stat_name, stat_label, stat_fmt)
```

```{r ch4-ex4-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "p.value" %in% .result$stat_name, "Perfect — the log-rank test ran and the ARD captured statistic, df, p.value, and method in the standard shape."),
  fail_if(~ TRUE, "Pass TRT01A to variables and Surv(AVAL, 1 - CNSR) to y.")
)
```

> **SAS check:** the p-value row corresponds to the `Pr > Chi-Square` value next to `Log-Rank` in the "Test of Equality over Strata" table from `PROC LIFETEST`. If you ran the same data through both, the two p-values should agree to many decimal places.

---

### 4.13 Formatting `{cardx}` Outputs

#### 4.13.1 Setting `fmt_fun` at calculation time

Every `{cardx}` function takes the same nested `fmt_fun` argument you learned in Chapter 3. Two common patterns:

```r
# Round p-values to 3 decimals, the rest to defaults
ard_stats_t_test(
  data      = adsl_2groups,
  variables = AGE,
  by        = TRT01A,
  fmt_fun   = list(AGE = list(p.value = 3, estimate = 1))
)
```

The keys at the inner level are `stat_name` values like `p.value`, `estimate`, `statistic`, `conf.low`, `conf.high`. The outer keys are the `variable` names.

#### 4.13.2 Updating an existing ARD with `update_ard_fmt_fun()`

When you have already calculated an ARD and want to re-format it without re-running the test:

```r
ard_ttest %>%
  update_ard_fmt_fun(p.value = 4) %>%   # bump p.value to 4 decimals
  apply_fmt_fun()
```

This is the cheap way to switch between submission-grade formatting (typically 3-4 decimals on p-values) and exploratory-grade formatting (looser) without recomputing.

#### 4.13.3 The `NULL` `stat_fmt` gotcha for `method` rows

Test ARDs always include a `method` row whose `stat` is a character string ("Welch Two Sample t-test") and whose `stat_fmt` is `NULL` — there is nothing to format. If you flatten with base R `unlist()`, those `NULL`s are silently dropped, your vector becomes shorter than the ARD's row count, and `select()` / `bind_cols()` fall over.

Always use `unlist_ard_columns(columns = stat_fmt)`. It replaces `NULL` with `NA` and preserves the row alignment that downstream layout code depends on.

```
   Without unlist_ard_columns():
     row count of stat_fmt: 6
     row count of ARD:      7  <-- mismatch -> downstream error

   With unlist_ard_columns():
     row count of stat_fmt: 7  <-- NA fills the method row, alignment kept
     row count of ARD:      7
```

> **SAS analog:** the same trap exists when you subset `ODS OUTPUT` datasets in a DATA step and drop "method" or "label" rows accidentally. The `unlist_ard_columns()` wrapper is the equivalent of always using `IF NOT MISSING(_LABEL_);` *carefully* — i.e., never.

---

### 4.14 Stacking Inferential with Descriptive

The big payoff: `{cardx}` rows can be `bind_ard()`ed with `{cards}` rows because they share the same schema. A typical demographics-plus-p-value table is one stack:

```r
ard_demog <- ard_stack(
  data = adsl_2groups,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by      = TRT01A,
  .total_n = TRUE
)

ard_pvals <- bind_ard(
  ard_stats_t_test(adsl_2groups, variables = AGE, by = TRT01A),
  ard_stats_fisher_test(adsl_2groups, variables = SEX, by = TRT01A)
)

ard_final <- bind_ard(ard_demog, ard_pvals)
```

`ard_final` is one tall ARD that contains every cell needed to render a demographics table with a "p-value" column attached — descriptive rows, inferential rows, header denominator rows, all in the same shape.

```
                +--------------------+
                |  ard_demog rows    |   <-- {cards}: descriptive
                +--------------------+
                +--------------------+
                |  ard_pvals rows    |   <-- {cardx}: inferential
                +--------------------+
                            |
                            v
                +--------------------+
                |  ard_final         |   <-- one input, one layout call
                +--------------------+
```

The fact that this works without renaming a single column is the entire point of the two-package design.

---

## Chapter 4 Knowledge Check

```{r ch4-quiz, echo=FALSE}
quiz(
  caption = "Chapter 4 — Check Your Understanding",

  question(
    "Why is `{cardx}` developed as a separate package from `{cards}`?",
    answer("Because `{cards}` is written in Python and `{cardx}` is in R"),
    answer("To keep `{cards}` dependency-light and fast, isolating heavy statistical dependencies like `{survival}` inside `{cardx}`", correct = TRUE),
    answer("There is no technical reason; they are duplicate packages"),
    answer("To support SAS integration"),
    allow_retry = TRUE
  ),

  question(
    "How does `ard_survival_survfit()` represent different time points in the ARD structure?",
    answer("In separate columns (time_6, time_12)"),
    answer("In the `variable_level` column as separate row records", correct = TRUE),
    answer("In the `group1_level` column"),
    answer("As list elements within a single row"),
    allow_retry = TRUE
  ),

  question(
    "Your ADaM `CNSR` column is `1 = censored`. What survival response do you pass to `ard_survival_survfit()`?",
    answer("`Surv(AVAL, CNSR)`"),
    answer("`Surv(AVAL, 1 - CNSR)`", correct = TRUE),
    answer("`Surv(AVAL)`"),
    answer("`Surv(CNSR, AVAL)`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{cardx}` function is the equivalent of `PROC FREQ; TABLES SEX*TRT01A / CHISQ;` in SAS?",
    answer("`ard_stats_t_test()`"),
    answer("`ard_stats_chisq_test()`", correct = TRUE),
    answer("`ard_survival_survdiff()`"),
    answer("`ard_regression()`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{cardx}` function corresponds most directly to `PROC LIFETEST` with the `STRATA TRT01A;` log-rank test of equality?",
    answer("`ard_survival_survfit()`"),
    answer("`ard_survival_survdiff()`", correct = TRUE),
    answer("`ard_stats_oneway_test()`"),
    answer("`ard_regression()`"),
    allow_retry = TRUE
  ),

  question(
    "Why must you use `unlist_ard_columns(columns = stat_fmt)` instead of base R `unlist()` on a `{cardx}` test result?",
    answer("Because `unlist()` is deprecated."),
    answer("Because test ARDs contain a `method` row whose `stat_fmt` is `NULL`, and base `unlist()` silently drops those rows, breaking row alignment.", correct = TRUE),
    answer("Because `unlist()` only works on `{cards}` objects, not `{cardx}` objects."),
    answer("They are identical — using either one is fine."),
    allow_retry = TRUE
  ),

  question(
    "Which `{cardx}` function corresponds to SAS's `PROC FREQ; TABLES REGION*TRT01A*RESPFL / CMH;` stratified test?",
    answer("`ard_stats_chisq_test()`"),
    answer("`ard_stats_fisher_test()`"),
    answer("`ard_stats_mantelhaen_test()`", correct = TRUE),
    answer("`ard_stats_kruskal_test()`"),
    allow_retry = TRUE
  ),

  question(
    "Which `{cardx}` function would you reach for to mimic SAS `PROC NPAR1WAY WILCOXON;` when comparing two arms?",
    answer("`ard_stats_t_test()`"),
    answer("`ard_stats_wilcox_test()`", correct = TRUE),
    answer("`ard_stats_chisq_test()`"),
    answer("`ard_stats_kruskal_test()` only"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 4 Summary {.congratulations}

You now know how to:

- Explain why `{cardx}` is split from `{cards}` and what design payoff that split delivers.
- Recognize the `{cardx}` function prefixes (`ard_stats_*`, `ard_survival_*`, `ard_regression`, `ard_emmeans_*`, `ard_effectsize_*`) and what each family wraps.
- Map every common SAS PROC for inferential statistics to its `{cardx}` counterpart:
  - `PROC TTEST` → `ard_stats_t_test()`, `ard_stats_paired_t_test()`, `ard_stats_t_test_onesample()`
  - `PROC NPAR1WAY` → `ard_stats_wilcox_test()`, `ard_stats_kruskal_test()`
  - `PROC GLM` / `PROC ANOVA` → `ard_stats_oneway_test()`, `ard_stats_aov()`, `ard_stats_anova()`
  - `PROC FREQ CHISQ` → `ard_stats_chisq_test()`
  - `PROC FREQ EXACT` → `ard_stats_fisher_test()`
  - `PROC FREQ AGREE` → `ard_stats_mcnemar_test()`
  - `PROC FREQ CMH` → `ard_stats_mantelhaen_test()`
  - `PROC MEANS CLM` → `ard_continuous_ci()`; binomial CI → `ard_categorical_ci()`
  - `PROC LIFETEST` Survival Estimates → `ard_survival_survfit()`
  - `PROC LIFETEST` Log-Rank → `ard_survival_survdiff()`
  - `LSMEANS` / `LSMESTIMATE` in `PROC MIXED` → `ard_emmeans_emmeans()`, `ard_emmeans_mean_difference()`, `ard_emmeans_contrast()`
- Read the inferential ARD row shape: `estimate`, `statistic`, `parameter`, `p.value`, `conf.low`, `conf.high`, `method`.
- Format `{cardx}` outputs both at calculation time (`fmt_fun = list(...)`) and after the fact (`update_ard_fmt_fun()` + `apply_fmt_fun()`).
- Safely flatten test ARDs with `unlist_ard_columns(columns = stat_fmt)` to avoid the `NULL`-drop bug from base R `unlist()`.
- Bind inferential ARDs with descriptive ones using `bind_ard()` because the schema is shared — one stacked ARD drives a demographics-plus-p-value table end to end.

**Next:** Chapter 5 — Models, `{broom}`, and `{parameters}` (the deep dive on regression tidying that the `ard_regression()` preview only sketched).
