---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 8: Advanced Workflows — Custom Statistics and CDISC ARS Alignment"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.2
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Write your own statistical functions, plug them into the cards ARD machinery,
  and translate every column of a flat ARD into the CDISC Analysis Results
  Standard (ARS) JSON schema — with every step mapped to its SAS analog.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)
library(tibble)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
data("adlb",  package = "pharmaverseadam")

# Pre-computed sample ARD row used by Exercise 8.2
ard_row_demo <- tibble(
  group1       = "TRT01A",
  group1_level = "Placebo",
  variable     = "AGE",
  stat_name    = "mean",
  stat         = list(75.21),
  stat_fmt     = "75.2"
)
```

---

## Chapter 8: Advanced Workflows

### 8.1 Why This Chapter Exists — The Last Two Gaps

By the end of Chapter 7 you can take an ADaM dataset, build an ARD from it, layout that ARD as a `{gtsummary}` table, style it through `{flextable}`, and ship it as a regulator-grade Word or RTF deliverable. That is a complete reporting pipeline.

But two needs keep showing up in real clinical projects, and neither is solved by the built-in `{cards}` function family:

1. **The statistic you need is not on the list.** PK/PD studies want geometric means. Biomarker work wants coefficient of variation. Safety wants exposure-adjusted incidence rates. Oncology wants median follow-up via the reverse Kaplan-Meier. None of those are in the default `statistic` argument of `ard_continuous()`.
2. **The regulator wants the dataset, not the table.** Modern submissions under the CDISC **Analysis Results Standard (ARS)** ask for a machine-readable JSON file that *describes* every analysis result — its operation, its grouping, its population, its raw and formatted values — alongside (or instead of) the rendered PDF. That JSON has a specific schema. Your ARD already has every field the schema needs, but the column names are different and the shape is flat-versus-nested.

This chapter closes both gaps:

```
   +--------------------------+        +-------------------------+
   |  YOUR ARD                |        |  TWO ADVANCED USES      |
   |  - standard schema       |        |                         |
   |  - n, mean, sd, p, ...   |  -+--> |  (1) Extend it with     |
   |                          |   |    |      CUSTOM statistics  |
   +--------------------------+   |    |      (sections 8.3-8.11)|
                                  |    |                         |
                                  +--> |  (2) Translate it to    |
                                       |      CDISC ARS JSON     |
                                       |      (sections 8.12-8.18)|
                                       +-------------------------+
```

You will not need any new external package — `{cards}` already exposes the customization seams, and `{jsonlite}` (which you have used since Chapter 1) does the JSON serialization.

---

### 8.2 SAS Programmer's Quick Map

If you have spent years writing SAS code, the two big ideas in this chapter both have direct analogs.

| SAS concept                                              | `{cards}` / R equivalent                              | Note                                                                   |
|----------------------------------------------------------|-------------------------------------------------------|------------------------------------------------------------------------|
| `PROC FCMP` user-defined function                        | An R function passed into `statistic = list(...)`     | Write the math once, register it as a named statistic.                 |
| `%MACRO geom_mean(var=);`                                | `geom_mean <- function(x) exp(mean(log(x[x>0])))`     | R functions are first-class objects — no `%LET`, no `%LOCAL`.          |
| `%LET DECIMALS = 1;` then `PUT(x, 5.1)`                  | `fmt_fun = list(VAR = list(stat = 1))`                | The "1" means *1 decimal place*; it travels next to the raw stat.      |
| `ODS OUTPUT MEANS=means_out;`                            | The ARD itself                                        | The ARD is the standardized "ODS OUTPUT" the regulator can consume.    |
| `PROC EXPORT ... DBMS=CSV;`                              | `jsonlite::toJSON(ard, auto_unbox = TRUE)`            | One call, one file, one wire format.                                   |
| Pinnacle 21 / Define-XML for SDTM/ADaM                   | CDISC ARS for *analysis results*                      | Standardizes the *outputs* (numbers) the way Define-XML standardizes the *inputs* (variables). |
| Re-running the program in a validation environment      | Diffing two ARDs row by row                           | Pure data comparison — no fragile PDF text extraction.                 |
| `PROC COMPARE base=qc compare=prod;`                     | `dplyr::anti_join(qc_ard, prod_ard, by = key_cols)`   | Same idea, but every ARD has the same key columns by construction.     |

**Key mental shift:** in SAS, an *extension* often means a new macro library wired into a new program; in R the extension is a function you wrote inline and passed by name. There is no macro variable scope to manage, no `%MACRO ... %MEND` block to remember to close.

---

### 8.3 What Is a Custom Statistic, Really?

A statistic, from `{cards}`'s point of view, is a function with one rule: *given a vector of numbers, return a value that can be stored in the `stat` list-column.* That value can be:

- a single number — `mean(x)` returns `length 1` numeric;
- a vector of numbers — `quantile(x, c(0.25, 0.75))` returns `length 2` numeric;
- a list with names — useful when the statistic naturally has parts (e.g., CI lower, CI upper).

That is the entire contract. If your R function satisfies it, `{cards}` will compute it for every group, store the result in `stat`, format it using the rule you pass via `fmt_fun`, place the formatted text in `stat_fmt`, and stamp the row with the name you registered.

```
                       INPUT: x  (a numeric vector for one group)
                                       |
                                       v
                       +-------------------------------+
                       |  YOUR FUNCTION                |
                       |    f(x) -> a value            |
                       +-------------------------------+
                                       |
                                       v
                       +-------------------------------+
                       |  {cards}                      |
                       |    stat       = f(x)          |
                       |    stat_fmt   = fmt(f(x))     |
                       |    stat_name  = "gmean"       |
                       |    stat_label = "Geo. Mean"   |
                       +-------------------------------+
```

> **SAS analog:** in `PROC FCMP` you write `FUNCTION GMEAN(x); RETURN(...); ENDSUB;` and then call `GMEAN(AGE)` inside a data step. In R it is `geom_mean <- function(x) ...` and then you pass `geom_mean` *by name* — no quoting, no macro-invocation syntax — into the `statistic =` argument.

#### 8.3.1 The minimum viable custom statistic

```r
geom_mean <- function(x) {
  exp(mean(log(x[x > 0]), na.rm = TRUE))
}
```

Three things to notice for a SAS reader:

- **No declarations of length, type, or label.** R infers all of that from the value the function returns.
- **`x[x > 0]`** is logical-index subsetting. It is the R equivalent of a SAS data-step `IF x > 0 THEN OUTPUT;`. Geometric mean is undefined for zero/negative values, so we drop them.
- **`na.rm = TRUE`** tells `mean()` to ignore missing values. In SAS, `MEAN()` does this automatically; in R, you must say so explicitly. This is one of the most common bugs when a SAS programmer first writes R.

That one-liner is now ready to plug into `ard_continuous()`.

---

### 8.4 Plugging a Custom Statistic into `ard_continuous()`

`ard_continuous()` accepts an argument called `statistic` that controls *which* statistics get computed for *which* variables. The shape of that argument is a *nested named list*:

```
statistic = list(
  <variable_name> = list(
    <stat_name_1> = <function_or_existing_name>,
    <stat_name_2> = <function_or_existing_name>,
    ...
  )
)
```

The outer key is the variable; the inner keys become `stat_name` in the ARD; the inner values are either:

- a built-in name like `"mean"` or `"sd"`, **or**
- an R function you wrote (or a `~`-formula expression that returns a value).

A minimal call that computes a geometric mean of `AGE` per treatment arm:

```r
ard_custom <- ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = list(
    AGE = list(
      gmean = geom_mean      # <-- this is the named custom slot
    )
  ),
  fmt_fun = list(
    AGE = list(
      gmean = 1              # <-- one decimal place in stat_fmt
    )
  )
)
```

What you get back is a perfectly ordinary ARD — same columns as Chapter 1 — with `stat_name = "gmean"` rows for each treatment arm. Downstream layout code does not need to know it was a custom statistic.

> **SAS analog:** writing `PROC MEANS DATA=adsl; CLASS TRT01A; VAR AGE; OUTPUT OUT=out MEAN=mn GEOMEAN=gm; RUN;`. The geometric mean keyword lives in the statistics list; in `{cards}` your custom slot lives in the `statistic =` list. Identical idea, different syntax.

```
   +-----------------------------+
   |  statistic = list(          |
   |    AGE = list(              |
   |      n     = "n",      <----+--- built-in
   |      mean  = "mean",   <----+--- built-in
   |      gmean = geom_mean <----+--- YOUR function
   |    )                        |
   |  )                          |
   +-----------------------------+
```

#### 8.4.1 Where each piece ends up in the ARD

| Place in `statistic = list(...)`           | Becomes...                       |
|--------------------------------------------|----------------------------------|
| Outer name (`AGE`)                         | The `variable` column.           |
| Inner name (`gmean`)                       | The `stat_name` column.          |
| Inner function (`geom_mean`)               | Called once per group; the return value becomes `stat`. |
| `fmt_fun$AGE$gmean = 1`                    | The formatting rule applied to produce `stat_fmt`. |

Nothing else changes — `group1`, `group1_level`, `variable`, `context`, the list-column behavior — all identical to a default `ard_continuous()` call.

---

## Exercise 8.1 — Run a Custom Geometric Mean Statistic

Write a geometric mean function and pass it to `ard_continuous()` to summarize `AGE` grouped by `TRT01A`. Confirm the resulting ARD has a `gmean` row per treatment arm.

```{r ch8-ex1, exercise=TRUE, exercise.lines=22}
# 1. Define the geometric mean function (ignores non-positive and NA values)
geom_mean <- function(x) {
  exp(mean(log(x[x > 0]), na.rm = TRUE))
}

# 2. Run ard_continuous and register your function under the name "gmean"
#    Hint: pass it inside statistic = list(AGE = list(gmean = ...))

ard_custom <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  statistic = list(
    AGE = list(
      gmean = ___
    )
  ),
  fmt_fun = list(
    AGE = list(
      gmean = 1
    )
  )
)

# Look at the result
ard_custom %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable, stat_name, stat_fmt)
```

```{r ch8-ex1-hint-1}
# Set gmean = geom_mean inside the inner list
```

```{r ch8-ex1-solution}
geom_mean <- function(x) {
  exp(mean(log(x[x > 0]), na.rm = TRUE))
}

ard_custom <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  statistic = list(
    AGE = list(
      gmean = geom_mean
    )
  ),
  fmt_fun = list(
    AGE = list(
      gmean = 1
    )
  )
)
ard_custom %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable, stat_name, stat_fmt)
```

```{r ch8-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result) && "gmean" %in% .result$stat_name, "Correct. You registered a custom statistic by name, and {cards} treated it identically to a built-in: same schema, same formatting pipeline, same list-column."),
  fail_if(~ TRUE, "Set gmean = geom_mean inside the nested statistic list, and supply the matching fmt_fun entry.")
)
```

> **Read the output:** the `stat_name` column contains exactly the literal `"gmean"` you typed. That string is the key by which `{gtsummary}` (or any downstream consumer) will find the row when it builds a table. Choose stat names that are short, lowercase, and stable — they become part of your project's API.

---

### 8.5 Multiple Custom Statistics in One Call

You almost never need *just one* custom statistic. PK summaries typically want at minimum `n`, geometric mean, geometric SD, and the geometric coefficient of variation. All four can be registered in the same `statistic =` list:

```r
geom_mean <- function(x)  exp(mean(log(x[x > 0]), na.rm = TRUE))
geom_sd   <- function(x)  exp(sd(log(x[x > 0]), na.rm = TRUE))
geom_cv   <- function(x)  sqrt(exp(sd(log(x[x > 0]), na.rm = TRUE)^2) - 1) * 100

ard_pk <- ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = list(
    AGE = list(
      n      = "n",                # mix built-ins...
      gmean  = geom_mean,          # ...with custom...
      gsd    = geom_sd,
      gcv_pc = geom_cv
    )
  ),
  fmt_fun = list(
    AGE = list(
      n      = 0,                  # integer
      gmean  = 1,                  # 1 decimal
      gsd    = 2,                  # 2 decimals
      gcv_pc = function(x) sprintf("%.1f%%", x)   # 1 dp with %
    )
  )
)
```

Three things worth flagging:

- The inner-list names (`gmean`, `gsd`, `gcv_pc`) are *your choice* and will appear in `stat_name`. Pick conventions and stick with them.
- `fmt_fun` accepts either an integer (number of decimals) **or** a function that takes a number and returns a formatted string. The `function(x) sprintf("%.1f%%", x)` style is the most flexible; it is the R equivalent of writing a SAS `PROC FORMAT VALUE` rule with a `%` suffix.
- Built-ins and custom stats coexist in the same list; the ARD output is one block of rows, one row per `(group1_level, variable, stat_name)` triple, no matter where the math came from.

> **SAS analog:** declaring multiple statistics keywords (`MEAN STD MEDIAN P25 P75`) in a single `PROC MEANS`. The `statistic =` list is the long-form, function-aware version of that keyword list.

---

### 8.6 Custom Statistics with Parameters — Closures

Sometimes a statistic has a tunable knob. The 95th percentile is a special case of "the *p*-th percentile". The truncated mean has a trim level. You do not want to write `p95`, `p97_5`, `p99` as three near-identical functions. R lets you write *one* function that *returns* a function — a **closure** — and pass each closure into a different slot.

```r
make_pctile <- function(p) {
  function(x) stats::quantile(x, probs = p, na.rm = TRUE, names = FALSE)
}

ard_pcts <- ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = list(
    AGE = list(
      p95   = make_pctile(0.95),
      p97_5 = make_pctile(0.975),
      p99   = make_pctile(0.99)
    )
  ),
  fmt_fun = list(
    AGE = list(p95 = 1, p97_5 = 1, p99 = 1)
  )
)
```

`make_pctile(0.95)` returns a *function* that already has `p = 0.95` baked in. `{cards}` calls that returned function with each group's `x` vector. The result is three rows per treatment arm — `p95`, `p97_5`, `p99` — all produced by the same one-line factory.

> **SAS analog:** parameterizing a `%MACRO PCTILE(p=);` and calling `%PCTILE(p=0.95)`, `%PCTILE(p=0.975)`, `%PCTILE(p=0.99)`. The R closure is the lambda-friendly equivalent and lives inside the same expression as the call — no separate `%MEND` cleanup.

---

### 8.7 Returning a Vector vs a Scalar — the List-Column Shape

`stat` is a *list-column*. That means each row of the ARD can hold an R object of any shape inside `stat`. This is by design: a single statistic sometimes naturally has multiple components.

```
   +----------------------------------------------+
   |  stat_name  =  "n"                           |
   |  stat       =  list( 86L )                   |   length-1 integer
   +----------------------------------------------+
   |  stat_name  =  "ci"                          |
   |  stat       =  list( c(70.1, 80.3) )         |   length-2 numeric (lower, upper)
   +----------------------------------------------+
   |  stat_name  =  "summary"                     |
   |  stat       =  list(                         |
   |     list(mean = 75.21, sd = 8.12, n = 86) )  |   nested named list
   +----------------------------------------------+
```

`{cards}` does not enforce a specific shape; it just stores whatever you return. The downstream consumer is responsible for understanding it.

Practical rule for clinical reporting:

- For a **single scalar** statistic (mean, median, p-value), return a single number.
- For a **pair** that conceptually travels together (CI lower + upper, range min + max), return a `length-2` numeric. Format both with one `fmt_fun` rule, then a layout step joins them as `"(70.1, 80.3)"`.
- For a **structured** statistic (HR + CI + p, all from one Cox model), prefer returning a *named list* so downstream code can pick by name.

> **SAS analog:** the difference between a `PROC MEANS OUTPUT OUT=` dataset (one column per statistic, scalar each) and a `PROC LIFETEST OUTSURV=` dataset (multiple columns with paired meanings — survival, lower CL, upper CL). The list-column is `{cards}`'s general-purpose container.

---

### 8.8 Custom `stat_label` — Making Tables Readable

`stat_name` is for code; `stat_label` is for humans (Chapter 1, section 1.3). You can override the auto-generated label per statistic via the `stat_label` argument, which mirrors the structure of `statistic =`:

```r
ard_pk_labeled <- ard_continuous(
  data       = adsl,
  variables  = AGE,
  by         = TRT01A,
  statistic  = list(AGE = list(gmean = geom_mean, gsd = geom_sd)),
  fmt_fun    = list(AGE = list(gmean = 1, gsd = 2)),
  stat_label = list(
    AGE = list(
      gmean = "Geometric Mean",
      gsd   = "Geometric SD"
    )
  )
)
```

When `{gtsummary}` builds a table, the row label inside the AGE section will read `"Geometric Mean"`, not the raw `"gmean"`. The internal `stat_name = "gmean"` is preserved for code that needs to filter rows.

> **SAS analog:** the difference between a SAS variable name (`GMEAN`) and the `LABEL` you attach for `PROC REPORT` column headers (`"Geometric Mean"`).

---

### 8.9 Custom Formatters — `fmt_fun` in Depth

You have already seen integer formatters (`= 1` meaning "one decimal"). The full `fmt_fun` argument accepts four shapes:

| `fmt_fun` value                                          | Meaning                                                |
|----------------------------------------------------------|--------------------------------------------------------|
| an integer `n`                                           | Round to `n` decimal places via `formatC(x, digits = n, format = "f")`. |
| a function `function(x) ...`                             | Apply that function to the raw `stat` value.           |
| a `~`-formula like `~ sprintf("%.2f%%", .x * 100)`       | Same idea, formula syntax. `.x` is the placeholder.    |
| a `cards::label_round()` helper                          | Convenience for typical clinical rounding rules.       |

The function form is the most flexible and is the one you reach for when:

- you need a *percent sign*: `function(x) sprintf("%.1f%%", x * 100)`
- you need to *truncate small p-values*: `function(x) ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))`
- you need to *guard against `NaN`*: `function(x) ifelse(is.nan(x), "NE", sprintf("%.2f", x))`

```r
ard_pretty <- ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = list(AGE = list(gmean = geom_mean)),
  fmt_fun   = list(
    AGE = list(
      gmean = function(x) {
        if (is.na(x) || is.nan(x)) return("NE")    # Not Estimable
        sprintf("%.2f", x)
      }
    )
  )
)
```

Two takeaways:

- A formatter is *the only* place where rounding is allowed. Never round `stat` itself. The raw value must survive in full precision for QC.
- Formatters can be reused: define `fmt_pct1 <- function(x) sprintf("%.1f%%", x * 100)` once and pass `fmt_pct1` into every percentage slot across every variable.

> **SAS analog:** a custom `PROC FORMAT VALUE` paired with a `PUT()` call. The R formatter is the function-call version; you can reuse it across as many ARDs as you want without re-declaring a format library.

---

### 8.10 Combining Custom and Built-in Statistics

Real PK or biomarker tables mix default statistics (`n`, `mean`, `sd`, `min`, `max`) with custom statistics in one ARD. The pattern is exactly the same nested list you already know:

```r
ard_pk_full <- ard_continuous(
  data       = adsl,
  variables  = AGE,
  by         = TRT01A,
  statistic  = list(
    AGE = list(
      n      = "n",                       # built-in
      mean   = "mean",                    # built-in
      sd     = "sd",                      # built-in
      gmean  = geom_mean,                 # custom
      gsd    = geom_sd,                   # custom
      min    = "min",                     # built-in
      max    = "max"                      # built-in
    )
  ),
  fmt_fun    = list(
    AGE = list(n = 0, mean = 1, sd = 2, gmean = 1, gsd = 2, min = 0, max = 0)
  ),
  stat_label = list(
    AGE = list(
      n     = "n", mean = "Mean", sd = "SD",
      gmean = "Geo. Mean", gsd = "Geo. SD",
      min   = "Min", max = "Max"
    )
  )
)
```

The order of inner list entries controls the row order in the ARD. That, in turn, drives the row order inside the AGE section of any downstream table. You get *layout-level* row ordering "for free" by ordering the `statistic =` list — no separate sort step.

> **SAS analog:** the order of statistics keywords after `PROC MEANS` controls the column order of the resulting wide dataset. Here, because the ARD is long, the order controls *row* order — same idea, transposed.

---

### 8.11 Defensive Patterns — NA, Empty Groups, Edge Cases

Custom functions get called on whatever vector the data happens to produce. That vector can be:

- entirely `NA` (no subjects met the criteria),
- empty (zero rows),
- a single value (one subject in a tiny stratum),
- a vector with non-positive values (problematic for `log()`).

A robust custom statistic anticipates these. The pattern:

```r
geom_mean_safe <- function(x) {
  x <- x[!is.na(x)]            # drop NA explicitly
  x <- x[x > 0]                # drop non-positive (log undefined)
  if (length(x) == 0) return(NA_real_)     # nothing left -> NA
  exp(mean(log(x)))
}
```

Three guard rails, every one of which has a SAS analog:

- `is.na(x)` filtering → `IF NOT MISSING(x);` in a data step.
- `x[x > 0]` filtering → `IF x > 0;`.
- `if (length(x) == 0) return(NA_real_)` → returning a missing in a `PROC FCMP` function when no observations satisfy the criteria.

Returning `NA_real_` (not just `NA`) preserves the numeric type of the `stat` column. The formatter can then translate `NA_real_` to `"NE"` (Not Estimable), which is the standard clinical convention.

---

## Exercise 8.2 — Multiple Custom Statistics with Safe Fallbacks

Build an ARD that reports **n**, the **geometric mean** of `AGE`, and the **geometric coefficient of variation** (`gcv_pc`, in percent) for each treatment arm. Make both custom functions safe against `NA` and non-positive values, returning `NA_real_` when no usable observations remain.

```{r ch8-ex2-pk, exercise=TRUE, exercise.lines=30}
# Safe geometric mean
geom_mean_safe <- function(x) {
  x <- x[!is.na(x)]
  x <- x[x > 0]
  if (length(x) == 0) return(NA_real_)
  exp(mean(log(x)))
}

# Safe geometric CV (%)
geom_cv_safe <- function(x) {
  x <- x[!is.na(x)]
  x <- x[x > 0]
  if (length(x) < 2) return(NA_real_)
  sqrt(exp(stats::sd(log(x))^2) - 1) * 100
}

# Build the ARD: n, gmean, gcv_pc grouped by TRT01A
ard_pk <- ard_continuous(
  data      = adsl,
  variables = AGE,
  by        = TRT01A,
  statistic = list(
    AGE = list(
      n      = "n",
      gmean  = ___,
      gcv_pc = ___
    )
  ),
  fmt_fun   = list(
    AGE = list(n = 0, gmean = 1, gcv_pc = 1)
  )
)

ard_pk %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable, stat_name, stat_fmt)
```

```{r ch8-ex2-pk-hint-1}
# Pass the function objects by name (no parentheses):
#   gmean  = geom_mean_safe
#   gcv_pc = geom_cv_safe
```

```{r ch8-ex2-pk-solution}
geom_mean_safe <- function(x) {
  x <- x[!is.na(x)]
  x <- x[x > 0]
  if (length(x) == 0) return(NA_real_)
  exp(mean(log(x)))
}
geom_cv_safe <- function(x) {
  x <- x[!is.na(x)]
  x <- x[x > 0]
  if (length(x) < 2) return(NA_real_)
  sqrt(exp(stats::sd(log(x))^2) - 1) * 100
}
ard_pk <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  statistic = list(
    AGE = list(
      n      = "n",
      gmean  = geom_mean_safe,
      gcv_pc = geom_cv_safe
    )
  ),
  fmt_fun = list(
    AGE = list(n = 0, gmean = 1, gcv_pc = 1)
  )
)
ard_pk %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable, stat_name, stat_fmt)
```

```{r ch8-ex2-pk-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("n", "gmean", "gcv_pc") %in% .result$stat_name), "Correct. Three statistics — built-in and custom — coexist in the same ARD with consistent formatting."),
  fail_if(~ TRUE, "Pass geom_mean_safe and geom_cv_safe by name (no parentheses) into the inner list.")
)
```

> **Why the safe versions matter:** the moment you submit an ARD to a regulator, every row is a claim. A custom function that crashes when a group has zero positive values does not produce wrong numbers — but it produces *missing rows*, which downstream layout tools will silently render as blank cells. The defensive pattern guarantees every row exists and contains either a valid number or `NA_real_` → `"NE"`, never silence.

---

### 8.12 CDISC ARS — What It Is and Why It Exists

For two decades, CDISC has been standardizing the *inputs* to clinical reporting: SDTM standardized the raw data, ADaM standardized the analysis-ready data, Define-XML described both. Until recently, the *outputs* — the tables, listings, and figures — were standardized only in their visual layout (e.g., ICH E3 section numbering), never in their data shape.

The **Analysis Results Standard (ARS)** closes that gap. It is a CDISC model that says: *for every reported analysis result, store a machine-readable record of what was computed, on which population, with which grouping, and what the resulting raw and formatted values were.*

```
   +--------------------------+         +--------------------------+
   |     PAST                 |         |     NOW (with ARS)       |
   |  PDF / RTF tables        |   --->  |  PDF / RTF tables        |
   |  (visually inspected)    |         |        +                 |
   |                          |         |  ARS JSON dataset        |
   +--------------------------+         |  (machine-validated)     |
                                        +--------------------------+
```

The ARS dataset is a JSON document (or XML, but JSON is the practical default) that mirrors a clinical-trial-report TFL inventory at the data level. A regulator can:

- re-run the math from the raw values stored in the ARS,
- pool results across studies by joining on the standardized fields,
- cross-check a PDF cell against the ARS rawValue with one script.

The shape that matters for `{cards}` users is the ARS **`Analysis`** object and its child **`OperationResult`** records. Every row in a flat ARD corresponds to exactly one `OperationResult`.

> **SAS analog:** if Define-XML is the spec for the *input* side, ARS is the same kind of spec for the *output* side. Pinnacle 21 already validates the inputs; ARS lets the same tooling validate the outputs.

---

### 8.13 The ARS Object Hierarchy at a Glance

The ARS model is rich, but for our mapping purposes only four levels matter:

```
   ReportingEvent
     |
     +-- AnalysisOutputs            <-- a "table" or "figure" in the CSR
            |
            +-- Analysis            <-- one named analysis within that output
                   |
                   +-- AnalysisSet          <-- the population (e.g., Safety Pop)
                   +-- Grouping             <-- the by-variable (e.g., TRT01A)
                   +-- Operation            <-- the statistic (e.g., mean)
                   +-- results              <-- a list of OperationResult
                              |
                              +-- OperationResult     <-- ONE ARD ROW
                                    operationId      ("mean")
                                    resultGroups[]   (grouping + group ID + raw + formatted)
                                       groupingId   ("TRT01A")
                                       groupId      ("Placebo")
                                    rawValue        (75.21)
                                    formattedValue  ("75.2")
```

The crucial line is the last block: **an `OperationResult` is a single ARD row**. The fields are renamed, the shape is nested instead of flat, but the *information content* is identical.

---

### 8.14 The Field-by-Field Mapping

Here is the dictionary that turns any ARD into ARS JSON. Memorize this once and the rest is mechanical.

| ARD column           | ARS field                          | Notes                                                                            |
|----------------------|------------------------------------|----------------------------------------------------------------------------------|
| `stat_name`          | `operationId`                      | The programmatic name of the statistic — exactly what ARS calls the `Operation`. |
| `stat_label`         | `Operation.label` (one level up)   | The human-readable label of the operation.                                        |
| `group1`             | `resultGroups[i].groupingId`       | The *name* of the stratification variable (e.g., `"TRT01A"`).                    |
| `group1_level`       | `resultGroups[i].groupId`          | The *value* within that variable (e.g., `"Placebo"`).                            |
| `group2`, `group3`   | additional entries in `resultGroups[]` | Each strata becomes one element in the array.                                   |
| `variable`           | `Analysis.analysisVariable` (parent level) | Identifies the variable analyzed; lives on the parent Analysis, not the row.   |
| `stat`               | `rawValue`                         | The unrounded, full-precision number. Critical for downstream auditability.       |
| `stat_fmt`           | `formattedValue`                   | The display string that appeared in the PDF cell.                                |
| `context`            | (loose mapping to Operation type)  | Tells the consumer this row came from `ard_continuous`, `ard_categorical`, etc.   |

```
   ARD row (flat)                      OperationResult (nested)
   ----------------------------        --------------------------------------
   group1        = "TRT01A"            operationId   = "mean"
   group1_level  = "Placebo"           resultGroups  = [
   variable      = "AGE"                  { groupingId = "TRT01A",
   stat_name     = "mean"                   groupId    = "Placebo" }
   stat          = list(75.21)         ]
   stat_fmt      = "75.2"              rawValue       = 75.21
                                       formattedValue = "75.2"
```

That is the entire schema mismatch: ARD names its keys with R-friendly column names; ARS names them with the controlled vocabulary regulators expect. The values are the same.

> **SAS analog:** the difference between a SAS `ODS OUTPUT` dataset and a CDISC-conformant transport dataset (`.xpt`) generated by `LIBNAME XPORT`. Same data, different naming convention, plus structural metadata. ARS is the JSON-era version of the same idea.

---

### 8.15 Writing a Mapper Function

Given the dictionary above, a mapper is mechanical. Here is the minimal version that takes one ARD row (as a one-row tibble) and returns the equivalent `OperationResult` as an R list — ready for `jsonlite::toJSON()`:

```r
map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(
      list(
        groupingId = row$group1,
        groupId    = row$group1_level
      )
    ),
    rawValue       = row$stat[[1]],     # un-list the list-column
    formattedValue = row$stat_fmt
  )
}
```

Three subtleties to point out for a SAS reader:

- **`row$stat[[1]]`** — `stat` is a list-column, so each cell is itself a list. `[[1]]` extracts the first (and usually only) element. This is the moment the list-column "unwraps" into a plain number.
- **The nested `list( list(...) )`** is the JSON-array-of-objects pattern. `jsonlite::toJSON()` will turn `list(list(...))` into `[ { ... } ]` — an *array* containing one *object*. ARS expects `resultGroups` to be an array, even when there is only one element.
- **`auto_unbox = TRUE`** in the eventual `toJSON()` call collapses length-1 vectors into JSON scalars (`75.21` instead of `[75.21]`). Forgetting that flag is the single most common bug when generating ARS-style JSON from R.

---

## Exercise 8.3 — Map a Flat ARD Row to CDISC ARS JSON

Complete the mapper function so it pulls the right ARD column into each ARS field. The input is the `ard_row_demo` tibble pre-loaded in setup (one row, one statistic).

```{r ch8-ex2, exercise=TRUE, exercise.lines=22}
# A single-row ARD (already loaded as ard_row_demo)
ard_row <- ard_row_demo
ard_row

# Complete the mapper
map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(
      list(
        groupingId = row$___,       # the stratification variable name
        groupId    = row$___        # the level within that variable
      )
    ),
    rawValue       = row$stat[[1]],
    formattedValue = row$___        # the display string
  )
}

# Run the mapper
map_row_to_ars(ard_row)
```

```{r ch8-ex2-hint-1}
# group1 -> groupingId
# group1_level -> groupId
# stat_fmt -> formattedValue
```

```{r ch8-ex2-solution}
map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(
      list(
        groupingId = row$group1,
        groupId    = row$group1_level
      )
    ),
    rawValue       = row$stat[[1]],
    formattedValue = row$stat_fmt
  )
}
map_row_to_ars(ard_row_demo)
```

```{r ch8-ex2-check}
grade_result(
  pass_if(~ .result$operationId == "mean" && .result$resultGroups[[1]]$groupId == "Placebo" && .result$formattedValue == "75.2", "Excellent. The flat ARD row is now in ARS-conformant nested shape, ready for jsonlite::toJSON()."),
  fail_if(~ TRUE, "Make sure you map row$group1 -> groupingId, row$group1_level -> groupId, and row$stat_fmt -> formattedValue.")
)
```

---

### 8.16 From a Mapped Row to Valid ARS JSON

Once the mapper returns a per-row list, scaling to a whole ARD is one `lapply()` away, and the final JSON string is one `toJSON()` away:

```r
# Step 1: map every row of the ARD into an OperationResult list
op_results <- lapply(split(ard, seq_len(nrow(ard))), map_row_to_ars)

# Step 2: wrap them inside the surrounding Analysis object
ars_analysis <- list(
  id                = "AN.DEMOG.AGE.MEAN",
  label             = "Demographics — Mean AGE by Treatment Arm",
  analysisVariable  = "AGE",
  results           = op_results
)

# Step 3: serialize to a string and write it out
ars_json <- jsonlite::toJSON(ars_analysis, auto_unbox = TRUE, pretty = TRUE)
writeLines(ars_json, "ars_demog_age_mean.json")
```

```
   +-----------------------------+
   |  flat ARD (n rows)          |
   +---------------+-------------+
                   |
                   |  lapply(split(ard, row), map_row_to_ars)
                   v
   +-----------------------------+
   |  list of OperationResult    |
   +---------------+-------------+
                   |
                   |  wrap in Analysis object
                   v
   +-----------------------------+
   |  R list (ARS-shaped)        |
   +---------------+-------------+
                   |
                   |  jsonlite::toJSON(auto_unbox = TRUE, pretty = TRUE)
                   v
   +-----------------------------+
   |  ars_demog_age_mean.json    |
   +-----------------------------+
```

The output file is what a downstream consumer — a regulator's reviewer, a pooling tool, an automated QC script — will read. Because it is a structured document and not a PDF, every field is addressable: an analyst can ask "what was the mean AGE in the Placebo arm of study X?" and get the answer in one line of code, without parsing any visual layout.

> **SAS analog:** the difference between sending a regulator an RTF file and sending them a `.xpt` SAS transport file. Both contain the same information; only one is machine-validatable. ARS JSON is the modern equivalent — JSON instead of XPT, but the auditability gain is the same.

#### 8.16.1 Round-tripping — ARS back to a data frame

A reviewer who receives the JSON file can rebuild the ARD with two calls:

```r
ars_in <- jsonlite::read_json("ars_demog_age_mean.json")

ard_rebuilt <- do.call(rbind, lapply(ars_in$results, function(op) {
  data.frame(
    group1       = op$resultGroups[[1]]$groupingId,
    group1_level = op$resultGroups[[1]]$groupId,
    stat_name    = op$operationId,
    stat         = unlist(op$rawValue),
    stat_fmt     = op$formattedValue,
    stringsAsFactors = FALSE
  )
}))
```

The fact that this round-trip is *symmetric* — ARD goes out, ARS comes back, ARD reappears — is the whole point of having a standard. The numbers cannot drift across the trip.

---

### 8.17 Cross-Study Pooling — Why a Standard Pays Off

If every study in a development program emits an ARS JSON file for every analysis result, pooling becomes trivial:

```r
ars_study_a <- jsonlite::read_json("studyA/ars_demog.json")
ars_study_b <- jsonlite::read_json("studyB/ars_demog.json")

# rebuild ARDs from each, tag with study, bind
ard_pool <- bind_rows(
  ars_to_ard(ars_study_a) %>% mutate(study = "A"),
  ars_to_ard(ars_study_b) %>% mutate(study = "B")
)

ard_pool %>%
  filter(stat_name == "mean", variable == "AGE") %>%
  group_by(group1_level) %>%
  summarise(pooled_mean = weighted.mean(stat, w = n))
```

The pooling is *just a data step*. No PDF parsing, no re-running anyone's SAS macros, no re-implementing rounding rules.

> **SAS analog:** combining `_FREQ_` from two studies' `ODS OUTPUT` datasets after a meta-data harmonization step. ARS removes the harmonization step because the column names are guaranteed identical by spec.

---

### 8.18 Reproducibility, Snapshots, and Provenance

A submission-grade ARD answers three questions on demand:

1. **What was computed?** — the rows themselves, with `stat_name`, raw `stat`, formatted `stat_fmt`.
2. **From what?** — the source data, captured by hash or version, alongside the ARD.
3. **By whom and when?** — a provenance block: package versions, R version, run timestamp, user.

A minimal provenance wrapper:

```r
ard_with_provenance <- list(
  ard       = ard_demog,
  source    = list(
    dataset = "adsl",
    md5     = digest::digest(adsl, algo = "md5"),
    rows    = nrow(adsl)
  ),
  software  = list(
    r_version    = R.version.string,
    cards        = as.character(packageVersion("cards")),
    cardx        = as.character(packageVersion("cardx")),
    jsonlite     = as.character(packageVersion("jsonlite"))
  ),
  generated = list(
    user      = Sys.info()[["user"]],
    timestamp = format(Sys.time(), tz = "UTC", usetz = TRUE)
  )
)

saveRDS(ard_with_provenance, "ard_demog_provenance.rds")
```

This is the R-side equivalent of a SAS `%LET RUNDT = %SYSFUNC(DATETIME());` plus a `OPTIONS NOTES NUMBER PAGENO=1;` capture, but stored as data instead of in a log file. It is part of the ARD when the ARD is what you ship.

> **SAS analog:** the `_NULL_` step that captures `&SYSDATE9.`, `&SYSUSERID.`, and the version of every macro called, and writes them as a banner at the top of every output. The R version stores the same fields as a structured list bound to the ARD itself.

---

### 8.19 QC Pattern — Diffing Two ARDs

The single biggest practical payoff of the ARD-first workflow is automated QC. Two independent programmers produce two ARDs from the same ADaM; the QC test is a row-by-row diff.

```r
key_cols   <- c("group1", "group1_level", "variable",
                "variable_grp", "variable_grp_level", "stat_name")

# rows in production but missing from QC
anti_join(ard_prod, ard_qc, by = key_cols)

# rows whose numeric value disagrees beyond a tolerance
ard_prod %>%
  inner_join(ard_qc, by = key_cols, suffix = c("_p", "_q")) %>%
  mutate(diff = unlist(stat_p) - unlist(stat_q)) %>%
  filter(abs(diff) > 1e-9)
```

That is the entire test. It runs in milliseconds, gives a deterministic answer, and is independent of any visual rendering.

> **SAS analog:** `PROC COMPARE` between two `ODS OUTPUT` datasets, but with the column-naming question already settled by the ARD spec. You never need to write `RENAME=` clauses.

---

## Exercise 8.4 — Round-Trip an ARD Through ARS JSON

Build a tiny one-row ARD, map it to an ARS `OperationResult` list using `map_row_to_ars()` from Exercise 8.3, serialize it with `jsonlite::toJSON()`, then parse it back with `jsonlite::fromJSON()` and confirm the `formattedValue` survives unchanged.

```{r ch8-ex3, exercise=TRUE, exercise.lines=20}
# A single ARD row to round-trip
one_row <- tibble(
  group1       = "TRT01A",
  group1_level = "Xanomeline Low Dose",
  variable     = "AGE",
  stat_name    = "mean",
  stat         = list(74.4),
  stat_fmt     = "74.4"
)

map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(list(groupingId = row$group1, groupId = row$group1_level)),
    rawValue       = row$stat[[1]],
    formattedValue = row$stat_fmt
  )
}

# 1. Map -> 2. Serialize -> 3. Parse back
ars_list   <- map_row_to_ars(one_row)
ars_json   <- jsonlite::toJSON(ars_list, auto_unbox = ___, pretty = TRUE)
parsed     <- jsonlite::fromJSON(ars_json, simplifyVector = FALSE)

# Return the formattedValue from the parsed JSON
parsed$formattedValue
```

```{r ch8-ex3-hint-1}
# auto_unbox = TRUE collapses length-1 vectors into JSON scalars.
```

```{r ch8-ex3-solution}
one_row <- tibble(
  group1       = "TRT01A",
  group1_level = "Xanomeline Low Dose",
  variable     = "AGE",
  stat_name    = "mean",
  stat         = list(74.4),
  stat_fmt     = "74.4"
)
map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(list(groupingId = row$group1, groupId = row$group1_level)),
    rawValue       = row$stat[[1]],
    formattedValue = row$stat_fmt
  )
}
ars_list   <- map_row_to_ars(one_row)
ars_json   <- jsonlite::toJSON(ars_list, auto_unbox = TRUE, pretty = TRUE)
parsed     <- jsonlite::fromJSON(ars_json, simplifyVector = FALSE)
parsed$formattedValue
```

```{r ch8-ex3-check}
grade_result(
  pass_if(~ identical(.result, "74.4"), "Round-trip succeeded. The formattedValue is unchanged because the JSON encoding preserved it exactly."),
  fail_if(~ TRUE, "Set auto_unbox = TRUE so length-1 vectors become JSON scalars, then re-run.")
)
```

> **What this proves:** the ARS JSON is a faithful, lossless carrier of the ARD. The text `"74.4"` you stored survives serialization, byte-shipping, and re-parsing. This is the property that lets two systems — yours and the regulator's — disagree about *everything else* (operating system, R/SAS version, library paths) and still agree on what number was computed.

---

### 8.20 The Full Advanced-Workflow Pipeline

Put the two halves together and the picture is one diagram you can keep on a sticky note:

```
   +---------------------+
   |  ADaM dataset       |
   +----------+----------+
              |
              |  ard_continuous(..., statistic = list(... = list(your_fn = ...)))
              v
   +---------------------+
   |  custom-stat ARD    |   <-- sections 8.3 - 8.11
   +----------+----------+
              |
              |  apply_fmt_fun()  +  bind_ard()  (Chapter 3)
              v
   +---------------------+
   |  display-ready ARD  |   <-- still the standard schema
   +----------+----------+
              |
              |  layout via gtsummary  (Chapter 6)
              |  styled via flextable  (Chapter 7)
              |        |
              |        v
              |   .docx / .rtf  ----> regulator (human-readable)
              |
              |  map_row_to_ars()  +  jsonlite::toJSON()
              v
   +---------------------+
   |  ARS JSON           |   <-- sections 8.12 - 8.18
   +---------------------+   ----> regulator (machine-readable)
```

The two arrows out of the display-ready ARD — one to the styled document, one to the ARS JSON — are why the entire `{cards}` philosophy is worth the effort. One source of truth, two deliverables, zero possibility of drift.

> **SAS analog:** in SAS you would write a program that produces the RTF, and a *separate* program (typically a different macro) that produces the regulator's transport datasets. The two programs are independent, and divergence between them is a real and recurring risk. In the `{cards}` workflow, both deliverables come from the same in-memory ARD; divergence is structurally impossible.

---

## Exercise 8.5 — End-to-End Mini Pipeline

Tie everything together. Given the `adsl` dataset, compute a custom **geometric mean** of `AGE` per `TRT01A` arm, attach a clear `stat_label`, format it to 1 decimal, then map the first row of the result to an ARS `OperationResult` list. Report that list.

```{r ch8-ex5, exercise=TRUE, exercise.lines=35}
# 1. Define the custom statistic
geom_mean_safe <- function(x) {
  x <- x[!is.na(x) & x > 0]
  if (length(x) == 0) return(NA_real_)
  exp(mean(log(x)))
}

# 2. Build the ARD with a custom label and a 1-decimal format
ard_gm <- ard_continuous(
  data       = adsl,
  variables  = AGE,
  by         = TRT01A,
  statistic  = list(AGE = list(gmean = geom_mean_safe)),
  fmt_fun    = list(AGE = list(gmean = 1)),
  stat_label = list(AGE = list(gmean = "Geometric Mean"))
) %>% apply_fmt_fun()

# 3. Map the FIRST row to ARS shape
map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(list(groupingId = row$group1, groupId = row$group1_level)),
    rawValue       = row$stat[[1]],
    formattedValue = row$stat_fmt
  )
}

first_row <- ard_gm[___, ]   # take the first row of the ARD
map_row_to_ars(first_row)
```

```{r ch8-ex5-hint-1}
# Use 1 to take the first row: ard_gm[1, ]
```

```{r ch8-ex5-solution}
geom_mean_safe <- function(x) {
  x <- x[!is.na(x) & x > 0]
  if (length(x) == 0) return(NA_real_)
  exp(mean(log(x)))
}
ard_gm <- ard_continuous(
  data       = adsl,
  variables  = AGE,
  by         = TRT01A,
  statistic  = list(AGE = list(gmean = geom_mean_safe)),
  fmt_fun    = list(AGE = list(gmean = 1)),
  stat_label = list(AGE = list(gmean = "Geometric Mean"))
) %>% apply_fmt_fun()
map_row_to_ars <- function(row) {
  list(
    operationId  = row$stat_name,
    resultGroups = list(list(groupingId = row$group1, groupId = row$group1_level)),
    rawValue       = row$stat[[1]],
    formattedValue = row$stat_fmt
  )
}
first_row <- ard_gm[1, ]
map_row_to_ars(first_row)
```

```{r ch8-ex5-check}
grade_result(
  pass_if(~ is.list(.result) && .result$operationId == "gmean" && !is.null(.result$resultGroups), "Beautifully done. You used a custom statistic, formatted it, and exposed it through an ARS-shaped record — the full advanced workflow in one pipeline."),
  fail_if(~ TRUE, "Take the first row with ard_gm[1, ] and pass it through map_row_to_ars().")
)
```

---

## Chapter 8 Knowledge Check

```{r ch8-quiz, echo=FALSE}
quiz(
  caption = "Chapter 8 — Check Your Understanding",

  question(
    "How do you pass a custom statistic function to `ard_continuous()`?",
    answer("Write a new R package"),
    answer("Pass a named list of functions to the `statistic` argument", correct = TRUE),
    answer("Pass it to the `fmt_fun` argument"),
    answer("You cannot calculate custom statistics in `{cards}`"),
    allow_retry = TRUE
  ),

  question(
    "Where does the **name** you assign in the inner `statistic` list end up in the ARD?",
    answer("The `variable` column"),
    answer("The `stat_label` column"),
    answer("The `stat_name` column", correct = TRUE),
    answer("Nowhere — it is internal only"),
    allow_retry = TRUE
  ),

  question(
    "Why should a custom statistic function return `NA_real_` (rather than crash) when the input has no usable values?",
    answer("Because R cannot crash"),
    answer("So that every ARD row exists; missing values become `NE` in `stat_fmt`, never silently dropped rows", correct = TRUE),
    answer("To save memory in the resulting ARD"),
    answer("Because regulators reject any output containing NA"),
    allow_retry = TRUE
  ),

  question(
    "Which `fmt_fun` shape produces the string `\"75.2%\"` from the number `0.752`?",
    answer("`gmean = 1`"),
    answer("`gmean = function(x) sprintf(\"%.1f%%\", x * 100)`", correct = TRUE),
    answer("`gmean = \"percent\"`"),
    answer("`gmean = 0.75`"),
    allow_retry = TRUE
  ),

  question(
    "Which CDISC ARS object directly corresponds to a single row of a flat `{cards}` ARD?",
    answer("AnalysisSet"),
    answer("OperationResult", correct = TRUE),
    answer("OutputDisplay"),
    answer("WhereClause"),
    allow_retry = TRUE
  ),

  question(
    "When mapping an ARD row to an ARS `OperationResult`, which ARS field receives `group1_level`?",
    answer("`operationId`"),
    answer("`groupingId`"),
    answer("`groupId`", correct = TRUE),
    answer("`formattedValue`"),
    allow_retry = TRUE
  ),

  question(
    "Which ARS field receives the **unrounded** numeric value from the ARD `stat` list-column?",
    answer("`formattedValue`"),
    answer("`rawValue`", correct = TRUE),
    answer("`operationId`"),
    answer("`groupingId`"),
    allow_retry = TRUE
  ),

  question(
    "Why is `auto_unbox = TRUE` important when calling `jsonlite::toJSON()` on an ARS-shaped list?",
    answer("It compresses the JSON file"),
    answer("It collapses length-1 R vectors into JSON scalars; without it, `\"mean\"` becomes `[\"mean\"]` and the schema validation fails", correct = TRUE),
    answer("It removes NA values"),
    answer("It is purely cosmetic"),
    allow_retry = TRUE
  ),

  question(
    "Which SAS construct most closely matches an R closure factory like `make_pctile(p = 0.95)`?",
    answer("`PROC TRANSPOSE`"),
    answer("A `%MACRO PCTILE(p=);` invoked with `%PCTILE(p=0.95)`", correct = TRUE),
    answer("`PROC SQL` correlated subquery"),
    answer("`OPTIONS MPRINT;`"),
    allow_retry = TRUE
  ),

  question(
    "Two programmers produce two ARDs from the same ADaM and you want a deterministic QC check. What is the most direct test?",
    answer("Open both RTF files and visually compare"),
    answer("`anti_join()` on the standard ARD key columns and compare `stat` values within tolerance", correct = TRUE),
    answer("Re-run both programs and inspect the logs"),
    answer("Render both ARDs to gtsummary tables and compare HTML"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 8 Summary {.congratulations}

You now know how to:

- Define a **custom statistic** as an ordinary R function and register it inside `statistic = list(...)` for `ard_continuous()` — no package required.
- Mix **built-in and custom** statistics in one call, control the row order via the order of the inner list, and assign **`stat_label`** to make the eventual table human-readable.
- Drive formatting with **`fmt_fun`** — integers for decimal places, functions for percent signs, p-value truncation, and `NA → "NE"` substitution.
- Use **closures** (`make_pctile(p = 0.95)`) to generate parameterized statistic functions without copy-paste.
- Build **defensive custom functions** that survive empty groups, all-`NA` inputs, and edge values, returning `NA_real_` instead of crashing.
- Read every column of a flat ARD as a field in the **CDISC ARS** `OperationResult`/`Analysis` hierarchy: `stat_name → operationId`, `group1/group1_level → groupingId/groupId`, `stat → rawValue`, `stat_fmt → formattedValue`.
- Write a **mapper function** that turns one ARD row into an ARS-shaped R list, scale it to the whole ARD with `lapply()`, and serialize with `jsonlite::toJSON(auto_unbox = TRUE)`.
- **Round-trip** an ARD through ARS JSON and back without information loss.
- Attach **provenance metadata** (source data hash, package versions, run user/time) so the shipped ARD answers what, from what, and by whom.
- Use ARDs for **deterministic, scriptable QC** via row-by-row diffs on standard key columns — the modern equivalent of `PROC COMPARE`.

**Congratulations!** You have completed the `{cards}` & `{cardx}` interactive training curriculum. You now hold a complete, ARD-first reporting pipeline: from raw ADaM to custom statistics, from stacked ARDs to styled Word/RTF deliverables, and from R-side data structures to CDISC-compliant, regulator-ready JSON.
