---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 1: Intro to ARD and cards"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  A hands-on tutorial covering the flat Analysis Results Data (ARD) standard
  using the cards R package and real pharmaverse datasets.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)

# learnr global options
tutorial_options(
  exercise.timelimit = 60
)

# gradethis setup
gradethis_setup()

# Load pharmaverse ADaM datasets
data("adsl",  package = "pharmaverseadam")
data("adae",  package = "pharmaverseadam")
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
data("adlb",  package = "pharmaverseadam")

# Pre-filter datasets
adsl_safe <- adsl %>% filter(SAFFL == "Y")
adsl_itt  <- adsl %>% filter(!is.na(ARM) & ARM != "")
adae_teae <- adae %>% filter(TRTEMFL == "Y")

# Load sample ARD JSON
ard_json <- jsonlite::read_json("data/adsl_demog_ard.json")
```

---

## Chapter 1: Intro to ARD and cards

### 1.1 Tidy Results vs. Static Displays

Traditionally, tables, figures, and listings (TFLs) are rendered directly into formatted PDF or RTF files. If a reviewer needs to run an audit or re-calculate a percentage, they must manually extract numbers from text boxes or recreate the calculation code. 

**Analysis Results Data (ARD)** separates the calculation of numbers from the formatting of those numbers into table layout structures. An ARD is a simple, standardized, flat table where every computed statistic lives on its own row, capturing the raw numeric precision (`stat_value`) alongside its rounded, formatted display string (`stat_value_fmt`).

---

### 1.2 The `{cards}` Schema

The `{cards}` package is the foundational tool for generating ARDs in R. Every function in `{cards}` returns a data frame containing standard columns:

- `group1`, `group1_level`: The grouping factor name and value (e.g., `TRT01A` and `Placebo`).
- `variable`: The variable being summarized (e.g., `AGE`).
- `variable_grp`, `variable_grp_level`: The categorical variable sub-group and level (e.g., `SEX` and `F`).
- `stat_name`, `stat_label`: The statistical name (e.g., `mean`) and its display label (e.g., `Mean`).
- `stat_value`, `stat_value_fmt`: The raw, full-precision statistic (stored in a list-column) and its formatted text representation.

---

## Exercise 1.1 — Explore the Scale of the Input Data

Before generating an ARD, let's explore our input data. We are working with the `adsl` subject-level dataset from the `{pharmaverseadam}` package. Let's find out how many unique subjects and treatment arms are in the Safety Population (`SAFFL == 'Y'`).

```{r ch1-ex1, exercise=TRUE, exercise.lines=8}
# Count the number of subjects in the Safety Population (SAFFL == "Y") 
# and list the unique values of TRT01A (Treatment Arm).

adsl %>%
  filter(SAFFL == "Y") %>%
  distinct(TRT01A)
```

```{r ch1-ex1-check}
grade_result(
  pass_if(~ nrow(.result) == 3, "Correct! There are 3 treatment groups in the safety population. Each group will require separate statistic rows in our generated ARD."),
  fail_if(~ TRUE, "Try filtering where SAFFL == 'Y' and retrieving distinct TRT01A values.")
)
```

---

## Exercise 1.2 — Parse the ARD JSON structure

A pre-calculated ARD has been loaded as `ard_json` (parsed from `data/adsl_demog_ard.json`). Explore the structure of the first row of this ARD.

```{r ch1-ex2, exercise=TRUE, exercise.lines=6}
# Print the first item of the ard_json list to see the column structure.
# Hint: use double square brackets [[1]]

ard_json[[___]]
```

```{r ch1-ex2-hint-1}
# Access the first element using list indexing:
ard_json[[1]]
```

```{r ch1-ex2-solution}
ard_json[[1]]
```

```{r ch1-ex2-check}
grade_result(
  pass_if(~ .result$variable == "AGE" && .result$stat_name == "n", "Great! The first row represents the count (n) of the AGE variable for the Placebo group."),
  fail_if(~ TRUE, "Try retrieving the first element using: ard_json[[1]]")
)
```

---

## Chapter 1 Knowledge Check

```{r ch1-quiz, echo=FALSE}
quiz(
  caption = "Chapter 1 — Check Your Understanding",

  question(
    "What is the main goal of Analysis Results Data (ARD)?",
    answer("To replace SDTM and ADaM datasets"),
    answer("To separate statistical calculations from presentation formatting", correct = TRUE),
    answer("To generate RTF files directly from SAS"),
    answer("To store raw clinical trial study design files"),
    allow_retry = TRUE
  ),

  question(
    "Why does the ARD store BOTH stat_value and stat_value_fmt?",
    answer("Because stat_value is for SAS and stat_value_fmt is for R"),
    answer("stat_value is for double-precision calculations and QC; stat_value_fmt is the formatted text for display", correct = TRUE),
    answer("To save space in the R session"),
    answer("They are identical, kept only for backward compatibility"),
    allow_retry = TRUE
  ),

  question(
    "Which column in a standard `{cards}` ARD holds the unrounded statistic?",
    answer("stat_name"),
    answer("stat_value", correct = TRUE),
    answer("stat_value_fmt"),
    answer("variable_grp"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 1 Summary {.congratulations}

You now understand:
- The concept of **Analysis Results Data (ARD)**.
- Why separating calculations from presentation is essential for automation and quality control.
- The columns that compose the standard `{cards}` ARD schema.
- The difference between `stat_value` (raw value in a list-column) and `stat_value_fmt` (formatted character string).

**Next:** Chapter 2 — Core `{cards}` Operations (`ard_continuous`, `ard_categorical`, etc.)
