---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 4: Extended Stats with cardx"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Explore how cardx extends the ARD framework with functions for t-tests, Fisher's exact tests, and Kaplan-Meier survival models.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(survival)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
data("adtte_onco", package = "pharmaverseadam")
adtte <- adtte_onco %>%
  left_join(
    adsl %>%
      mutate(ITTFL = ifelse(!is.na(ARM) & ARM != "", "Y", "N")) %>%
      select(USUBJID, TRT01A, ITTFL),
    by = "USUBJID"
  )
```

---

## Chapter 4: Extended Stats with cardx

### 4.1 Statistical Testing & Survival Wrap-ups

`{cardx}` extends the `{cards}` ecosystem by wrapping standard R statistical analysis packages (like `{stats}` and `{survival}`). Rather than returning raw model outputs or complex vectors, `{cardx}` functions process the math and return a standardized ARD dataset.

Key statistical test functions include:
- `ard_stats_t_test()`: Run a t-test to compare means.
- `ard_stats_fisher_test()`: Run a Fisher's exact test for categorical variables.
- `ard_cmh_test()`: Cochran-Mantel-Haenszel test.
- `ard_survival_survfit()`: Extract Kaplan-Meier survival estimates at specific time points.

### 4.2 Formatting Statistics in cardx

Statistical test functions in `{cardx}` return multiple values (e.g., test statistics, confidence intervals, p-values). Formatting these values is managed in two ways:

1. **`apply_fmt_fun()`**: This function reads the formatting rules defined in the `fmt_fn` column of the ARD, applies them to the raw values in the `stat` column, and appends the **`stat_fmt`** column (which holds the display text).
2. **Modifying formatting rules:**
   - **During Calculation:** You can pass a named list to the `fmt_fn` argument mapping statistics to decimal integers or custom formatting functions:
     ```r
     ard_stats_t_test(
       data = adsl_2groups, 
       variables = AGE, 
       by = TRT01A,
       fmt_fn = list(p.value = 3, estimate = 1)
     )
     ```
   - **Post-Calculation:** You can modify the formatting rules of an already-computed ARD using **`update_ard_fmt_fun()`** before you call `apply_fmt_fun()`:
     ```r
     ard_ttest %>%
       update_ard_fmt_fun(p.value = 4) %>%  # Change p-value format to 4 decimals
       apply_fmt_fun()
     ```
3. **Safely flattening columns:**
   - Because hypothesis tests and models return metadata rows (like the statistical `method`) that aren't formatted, `stat_fmt` will contain `NULL` for those rows. Direct use of R's `unlist()` will drop these `NULL`s, resulting in vector length mismatches. Always use **`unlist_ard_columns()`** (or `unlist_ard_columns(columns = stat_fmt)`) to safely flatten the column while keeping the row count intact.

---

## Exercise 4.1 — Perform a Two-Sample t-test

Let's compare the mean age between the Placebo and Xanomeline High Dose groups. A t-test compares exactly two groups, so we must filter out the Low Dose group first.

```{r ch4-ex1, exercise=TRUE, exercise.lines=15}
# Filter adsl to Placebo and Xanomeline High Dose only
adsl_2groups <- adsl %>%
  filter(TRT01A %in% c("Placebo", "Xanomeline High Dose"))

# Perform t-test comparing AGE by TRT01A and format as ARD
# Hint: use ard_stats_t_test()

ard_ttest <- ___(
  data = adsl_2groups,
  variables = AGE,
  by = TRT01A
)

# Look at the statistical results
ard_ttest %>%
  apply_fmt_fun() %>%
  select(stat_name, stat_label, stat_fmt)
```

```{r ch4-ex1-hint-1}
# Call ard_stats_t_test on adsl_2groups
ard_stats_t_test(adsl_2groups, variables = AGE, by = TRT01A)
```

```{r ch4-ex1-solution}
ard_ttest <- ard_stats_t_test(
  data = adsl_2groups,
  variables = AGE,
  by = TRT01A
)
ard_ttest %>%
  apply_fmt_fun() %>%
  select(stat_name, stat_label, stat_fmt)
```

```{r ch4-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result), "Correct! ard_stats_t_test generated standard test rows, and apply_fmt_fun() formatted them into stat_fmt."),
  fail_if(~ TRUE, "Ensure you are using ard_stats_t_test() on adsl_2groups and selecting the correct columns via apply_fmt_fun().")
)
```

---

## Exercise 4.2 — Calculate Kaplan-Meier survival estimates

In oncology trials, Overall Survival (OS) is a key parameter. We need to calculate KM survival estimates at month time points (6 and 12 months) from the `adtte` dataset.

The `ard_survival_survfit()` function accepts:
- `x`: The input data frame.
- `y`: The survival response object, created with `Surv(time, event)` (or `survival::Surv(time, event)`).
- `variables`: The grouping/strata variable (e.g. `TRT01A`).
- `times`: The numeric vector of time points for estimates.

```{r ch4-ex2, exercise=TRUE, exercise.lines=15}
# Filter for Overall Survival (OS)
adtte_os <- adtte %>% filter(PARAMCD == "OS")

# Calculate survival estimates at Months 6 and 12, grouped by TRT01A
# Hint: use ard_survival_survfit() and specify times = c(6, 12)

ard_km_rates <- ard_survival_survfit(
  x = adtte_os,
  y = Surv(AVAL, 1 - CNSR),
  variables = TRT01A,
  times = ___
)

# View results
ard_km_rates %>%
  apply_fmt_fun() %>%
  select(group1_level, variable_grp_level, stat_name, stat_fmt)
```

```{r ch4-ex2-hint-1}
# Set times = c(6, 12)
```

```{r ch4-ex2-solution}
ard_km_rates <- ard_survival_survfit(
  x = adtte_os,
  y = Surv(AVAL, 1 - CNSR),
  variables = TRT01A,
  times = c(6, 12)
)
ard_km_rates %>%
  apply_fmt_fun() %>%
  select(group1_level, variable_grp_level, stat_name, stat_fmt)
```

```{r ch4-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result), "Incredible! You generated Kaplan-Meier survival statistics at specified time intervals. These are mapped to rows where variable_grp_level denotes the timepoint (e.g. 6 or 12)."),
  fail_if(~ TRUE, "Check that you specify times = c(6, 12) in your ard_survival_survfit call and format the output.")
)
```

---

## Chapter 4 Knowledge Check

```{r ch4-quiz, echo=FALSE}
quiz(
  caption = "Chapter 4 — Check Your Understanding",

  question(
    "Why is `{cardx}` developed as a separate package from `{cards}`?",
    answer("Because `{cards}` is written in Python and `{cardx}` is in R"),
    answer("To keep `{cards}` dependency-light and fast, isolating heavy statistical dependencies like `{survival}` inside `{cardx}`", correct = TRUE),
    answer("There is no technical reason; they are duplicate packages"),
    answer("To support SAS integration"),
    allow_retry = TRUE
  ),

  question(
    "How does `ard_survival_survfit()` represent different time points in the ARD structure?",
    answer("In separate columns (time_6, time_12)"),
    answer("In the `variable_grp_level` column as separate row records", correct = TRUE),
    answer("In the `group1_level` column"),
    answer("As list elements within a single row"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 4 Summary {.congratulations}

You now know how to:
- Use **`{cardx}`** functions to run statistical comparisons.
- Run a t-test and extract statistics using **`ard_stats_t_test()`**.
- Compute Kaplan-Meier survival rates at specific timepoints using **`ard_survival_survfit()`**.

**Next:** Chapter 5 — Models, `{broom}`, and `{parameters}` (Understanding regression model tidying)
