---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 8: Advanced Workflows"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Write custom statistical functions and explore the field-by-field mapping between cards ARDs and CDISC ARS structures.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
```

---

## Chapter 8: Advanced Workflows

### 8.1 Custom Statistics & Standards Alignment

In clinical programming, we often need to calculate non-standard statistics (like geometric means in PK/PD studies) or audit calculations for regulatory compliance. 

- **Custom Statistics**: You can pass a named list of custom functions to `ard_continuous(statistic = ...)` to calculate anything.
- **CDISC ARS Alignment**: Because `{cards}` ARD is structured, it maps directly, row-by-row, to the CDISC Analysis Results Standard (ARS) JSON schema's **`OperationResult`** objects, enabling automated regulatory validation and traceability.

---

## Exercise 8.1 — Run a Custom Geometric Mean Statistic

Let's write a simple geometric mean function: `geom_mean = function(x) exp(mean(log(x[x > 0]), na.rm = TRUE))`. We will pass it to `ard_continuous()` to summarize `AGE` grouped by `TRT01A`.

```{r ch8-ex1, exercise=TRUE, exercise.lines=20}
# 1. Define the geometric mean function
geom_mean <- function(x) {
  exp(mean(log(x[x > 0]), na.rm = TRUE))
}

# 2. Run ard_continuous using the custom function
# Hint: Pass your function as a named item inside statistic = list(AGE = list(gmean = ...))

ard_custom <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  statistic = list(
    AGE = list(
      gmean = ___
    )
  ),
  fmt_fn = list(
    AGE = list(
      gmean = 1
    )
  )
)

# Look at the results
ard_custom %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable, stat_name, stat_fmt)
```

```{r ch8-ex1-hint-1}
# Set gmean = geom_mean inside statistic
```

```{r ch8-ex1-solution}
geom_mean <- function(x) {
  exp(mean(log(x[x > 0]), na.rm = TRUE))
}

ard_custom <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  statistic = list(
    AGE = list(
      gmean = geom_mean
    )
  ),
  fmt_fn = list(
    AGE = list(
      gmean = 1
    )
  )
)
ard_custom %>%
  apply_fmt_fun() %>%
  unlist_ard_columns(columns = stat_fmt) %>%
  select(group1_level, variable, stat_name, stat_fmt)
```

```{r ch8-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result) && "gmean" %in% .result$stat_name, "Correct! You successfully calculated and formatted a custom statistic. This demonstrates how extensible the `{cards}` package is for non-standard clinical summaries."),
  fail_if(~ TRUE, "Check that you map gmean = geom_mean inside the nested statistic list and format the output.")
)
```

---

## Exercise 8.2 — Map a Flat ARD Row to CDISC ARS JSON

Each row in a `{cards}` ARD corresponds to a CDISC ARS `OperationResult` object inside `Analysis.results`. 

Write a simple mapper function that takes a single-row ARD dataframe (or list) and maps its elements to the CDISC ARS JSON list structure:
- `stat_name` -> `operationId`
- `group1` -> `groupingId`
- `group1_level` -> `groupId`
- `stat` -> `rawValue`
- `stat_fmt` -> `formattedValue`

```{r ch8-ex2, exercise=TRUE, exercise.lines=20}
# A single-row ARD dataframe
ard_row <- tibble(
  group1 = "TRT01A",
  group1_level = "Placebo",
  variable = "AGE",
  stat_name = "mean",
  stat = list(75.21),
  stat_fmt = "75.2"
)

# Complete the mapper function
map_row_to_ars <- function(row) {
  list(
    operationId = row$stat_name,
    resultGroups = list(
      list(
        groupingId = row$___,      # group1
        groupId = row$___          # group1_level
      )
    ),
    rawValue = row$stat[[1]],
    formattedValue = row$___       # stat_fmt
  )
}

# Run mapper
map_row_to_ars(ard_row)
```

```{r ch8-ex2-hint-1}
# Map group1 to groupingId, group1_level to groupId, and stat_fmt to formattedValue
```

```{r ch8-ex2-solution}
map_row_to_ars <- function(row) {
  list(
    operationId = row$stat_name,
    resultGroups = list(
      list(
        groupingId = row$group1,
        groupId = row$group1_level
      )
    ),
    rawValue = row$stat[[1]],
    formattedValue = row$stat_fmt
  )
}
map_row_to_ars(ard_row)
```

```{r ch8-ex2-check}
grade_result(
  pass_if(~ .result$operationId == "mean" && .result$resultGroups[[1]]$groupId == "Placebo", "Superb! You mapped the flat ARD row to the CDISC ARS logical schema. This row can now be serialized directly as compliance-conformant JSON for submission!"),
  fail_if(~ TRUE, "Ensure you are pulling row$group1, row$group1_level, and row$stat_fmt inside the function.")
)
```

---

## Chapter 8 Knowledge Check

```{r ch8-quiz, echo=FALSE}
quiz(
  caption = "Chapter 8 — Check Your Understanding",

  question(
    "How do you pass a custom statistic function to `ard_continuous()`?",
    answer("Write a new R package"),
    answer("Pass a named list of functions to the `statistic` argument", correct = TRUE),
    answer("Pass it to the `fmt_fn` argument"),
    answer("You cannot calculate custom statistics in `{cards}`"),
    allow_retry = TRUE
  ),

  question(
    "Which CDISC ARS object directly corresponds to a row of flat `{cards}` ARD containing a calculated value?",
    answer("AnalysisSet"),
    answer("OperationResult", correct = TRUE),
    answer("OutputDisplay"),
    answer("WhereClause"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 8 Summary {.congratulations}

You now know how to:
- Write and integrate **custom continuous statistics** functions in `{cards}`.
- Map flat ARD columns directly to CDISC ARS **`OperationResult`** JSON specifications.
- Understand the automated quality control benefits of standardized results metadata.

**Congratulations!** You have completed the `{cards}` & `{cardx}` interactive training curriculum!
