---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 3: Stacking and Formatting"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Master combining multiple ARD objects using bind_ard and customizing decimal formats with fmt_fn in cards.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(pharmaversesdtm)
library(dplyr)
library(cards)
library(cardx)
library(jsonlite)
library(glue)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

# Load datasets
data("adsl",  package = "pharmaverseadam")
```

---

## Chapter 3: Stacking and Formatting

### 3.1 Combining ARD Data Frames

Most clinical trial tables present more than one variable. In `{cards}`, you generate individual ARD data frames for each variable (or set of variables of the same type) and then bind them vertically to create a single, unified dataset.

- **`bind_ard()`**: Combines multiple ARD datasets. It validates the ARD structure and checks for duplicate rows (e.g., repeating the same analysis statistic for the same group and variable).
- **`ard_stack()`**: A wrapper that lets you perform continuous and categorical calculations in a single function call.

---

### 3.2 Formatting Clinical Statistics

Clinical standards dictate the exact number of decimal places for each statistic:
- Sample sizes (`n`) are typically formatted as integers (0 decimal places).
- Means and Medians are formatted to 1 decimal place.
- Standard deviations (`sd`) are formatted to 2 decimal places.

In `{cards}`, you specify these rules using the `fmt_fn` argument when calculating the ARD. The formatted string is generated in the `stat_fmt` column after calling `apply_fmt_fun()`.

- **`unlist_ard_columns()`**: Because `stat` and `stat_fmt` are list-columns, they can be difficult to work with in standard tables. `{cards}` provides `unlist_ard_columns()` to cleanly flatten these columns into standard vectors. By default, it replaces any `NULL` elements with `NA` (preventing standard R `unlist()` errors where `NULL`s are dropped and row counts mismatch). You can target specific columns by calling `unlist_ard_columns(columns = stat_fmt)`.

---

## Exercise 3.1 — Combine Multiple ARD Outputs

Calculate continuous summaries for `AGE` and categorical summaries for `SEX`, and combine them using `bind_ard()`.

``` {r ch3-ex1, exercise=TRUE, exercise.lines=15}
# 1. Generate continuous ARD for AGE
ard_age <- ard_continuous(adsl, variables = AGE, by = TRT01A)

# 2. Generate categorical ARD for SEX
ard_sex <- ard_categorical(adsl, variables = SEX, by = TRT01A)

# 3. Bind them together into one demographic ARD
# Hint: use bind_ard()

ard_demog <- ___(ard_age, ard_sex)
ard_demog
```

``` {r ch3-ex1-hint-1}
# Call bind_ard with your two ARD data frames
bind_ard(ard_age, ard_sex)
```

``` {r ch3-ex1-solution}
bind_ard(ard_age, ard_sex)
```

``` {r ch3-ex1-check}
grade_result(
  pass_if(~ is.data.frame(.result) && all(c("AGE", "SEX") %in% .result$variable), "Correct! bind_ard stacked both tables. Downstream tools can now render a unified demographics table from this single ARD object."),
  fail_if(~ TRUE, "Ensure you are calling bind_ard(ard_age, ard_sex).")
)
```

---

## Exercise 3.2 — Apply Custom Formatting Rules

Calculate continuous summaries for `AGE` grouped by `TRT01A`, but apply custom formatting:
- `n` = 0 decimal places
- `mean` = 1 decimal place
- `sd` = 2 decimal places

``` {r ch3-ex2, exercise=TRUE, exercise.lines=15}
# Generate continuous summaries with custom formatting rules
# Hint: pass a nested list to fmt_fn mapping stat_names to decimal integers

ard_age_fmt <- ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  fmt_fn = list(
    AGE = list(
      n = ___,
      mean = ___,
      sd = ___
    )
  )
)

# Apply formatting function to add 'stat_fmt' column and select columns
ard_age_fmt %>%
  apply_fmt_fun() %>%
  select(variable, stat_name, stat_fmt)
```

``` {r ch3-ex2-hint-1}
# Set n = 0, mean = 1, sd = 2
# fmt_fn = list(AGE = list(n = 0, mean = 1, sd = 2))
```

``` {r ch3-ex2-solution}
ard_continuous(
  data = adsl,
  variables = AGE,
  by = TRT01A,
  fmt_fn = list(
    AGE = list(
      n = 0,
      mean = 1,
      sd = 2
    )
  )
) %>%
  apply_fmt_fun() %>%
  select(variable, stat_name, stat_fmt)
```

``` {r ch3-ex2-check}
grade_result(
  pass_if(~ is.data.frame(.result) && "stat_fmt" %in% names(.result) && .result$stat_fmt[.result$stat_name == "mean"][[1]] == "74.4", "Awesome! You customized the decimals and applied the formatting function to generate the stat_fmt column."),
  fail_if(~ TRUE, "Double-check your nested list structure in fmt_fn: list(AGE = list(n = 0, mean = 1, sd = 2)) and verify you use apply_fmt_fun() and select(variable, stat_name, stat_fmt).")
)
```

---

## Chapter 3 Knowledge Check

``` {r ch3-quiz, echo=FALSE}
quiz(
  caption = "Chapter 3 — Check Your Understanding",

  question(
    "Which function in `{cards}` is used to combine multiple ARD data frames vertically?",
    answer("dplyr::bind_rows()"),
    answer("bind_ard()", correct = TRUE),
    answer("stack_ard()"),
    answer("ard_merge()"),
    allow_retry = TRUE
  ),

  question(
    "What happens by default if `bind_ard()` detects duplicate statistic rows across inputs?",
    answer("It silently ignores the duplicates"),
    answer("It throws an error/warning to prevent inconsistent results", correct = TRUE),
    answer("It merges them by calculating the average"),
    answer("It deletes the entire dataset"),
    allow_retry = TRUE
  ),

  question(
    "In which column does `{cards}` store the rounded, formatted display string (after calling `apply_fmt_fun()`)?",
    answer("stat"),
    answer("stat_fmt", correct = TRUE),
    answer("stat_label"),
    answer("context"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 3 Summary {.congratulations}

You now know:
- How to stack multiple ARD data frames using **`bind_ard()`**.
- How `bind_ard()` protects against data duplication.
- How to configure decimal places and formatting rules using the **`fmt_fn`** list inside calculation functions.

**Next:** Chapter 4 — Extended Stats with `{cardx}` (Hypothesis tests and Kaplan-Meier curves)
