---
title: "cards & cardx R Packages — Interactive Training"
subtitle: "Chapter 7: Exporting and MS Word Styling"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Learn how to bridge gtsummary to MS Word using flextable and officer to enforce strict clinical layouts, margins, headers, and footers.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(pharmaverseadam)
library(dplyr)
library(cards)
library(gtsummary)
library(flextable)
library(officer)

tutorial_options(
  exercise.timelimit = 60
)
gradethis_setup()

data("adsl", package = "pharmaverseadam")

# Pre-compute an ARD and table for exercises to save time
ard_demog <- ard_stack(
  data = adsl,
  ard_continuous(variables = AGE),
  ard_categorical(variables = SEX),
  .by = TRT01A,
  .missing = TRUE
)

base_table <- tbl_ard_summary(
  cards = ard_demog,
  include = c(AGE, SEX),
  by = TRT01A
) %>%
  modify_spanning_header(all_stat_cols() ~ "**Treatment Group**")
```

---

## Chapter 7: Exporting to MS Word

While `{gtsummary}` handles statistical mapping and internal table strings flawlessly, the clinical trials industry heavily relies on Microsoft Word (`.docx`) or RTF formats for final deliverables like Clinical Study Reports (CSRs).

The bridge between `{gtsummary}` and MS Word is the **`{flextable}`** package. 

### 7.1 Injecting Headers and Footers

Clinical tables require rigid structural metadata (e.g., Protocol ID, Sponsor name, footnotes, and generation timestamps). You can embed these directly into the table block using `add_header_lines()` and `add_footer_lines()`.

---

## Exercise 7.1 — Headers and Footers

Convert the existing `base_table` into a `flextable` and inject a sponsor header and a timestamp footer.

```{r ch7-ex1, exercise=TRUE, exercise.lines=15}
# Convert the gtsummary object to flextable
flex_tbl <- base_table %>%
  ___()

# Add header and footer lines
flex_tbl_styled <- flex_tbl %>%
  # Hint: add_header_lines("Protocol: ABC-123")
  ___ %>%
  # Hint: add_footer_lines("Generated via R-4.4")
  ___

flex_tbl_styled
```

```{r ch7-ex1-hint-1}
# flex_tbl <- base_table %>% as_flex_table()
# flex_tbl_styled <- flex_tbl %>%
#   add_header_lines("Protocol: ABC-123") %>%
#   add_footer_lines("Generated via R-4.4")
```

```{r ch7-ex1-solution}
flex_tbl <- base_table %>%
  as_flex_table()

flex_tbl_styled <- flex_tbl %>%
  add_header_lines("Protocol: ABC-123") %>%
  add_footer_lines("Generated via R-4.4")

flex_tbl_styled
```

```{r ch7-ex1-check}
grade_result(
  pass_if(~ inherits(.result, "flextable"), "Success! Notice how the header and footer are attached directly to the grid of the table."),
  fail_if(~ TRUE, "Check your function names: as_flex_table(), add_header_lines(), and add_footer_lines().")
)
```

---

## Exercise 7.2 — Advanced Document Layouts (`{officer}`)

By default, `save_as_docx()` prints to a standard portrait page. Clinical tables usually have many columns (e.g., multiple treatment arms plus a total column) and require **Landscape** orientation.

To define page layouts and margins in Word, we use the `{officer}` package to define a **section property** (`prop_section`).

```{r ch7-ex2, exercise=TRUE, exercise.lines=16}
# 1. Convert base_table to flextable
my_flex <- as_flex_table(base_table)

# 2. Define a landscape section property with custom margins
# Hint: prop_section(page_size = page_size(orient = "landscape"), page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1))
my_section <- ___

# 3. View the section object to pass the check
my_section
```

```{r ch7-ex2-hint-1}
# my_section <- prop_section(
#   page_size = page_size(orient = "landscape"),
#   page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
# )
```

```{r ch7-ex2-solution}
my_flex <- as_flex_table(base_table)

my_section <- prop_section(
  page_size = page_size(orient = "landscape"),
  page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
)

my_section
```

```{r ch7-ex2-check}
grade_result(
  pass_if(~ inherits(.result, "prop_section"), "Perfect! You now know how to define a landscape mode with 1-inch margins."),
  fail_if(~ TRUE, "Make sure you assigned a valid prop_section() to my_section.")
)
```

---

## Exercise 7.3 — Exporting the File

Now let's actually run the `{flextable}` export engine. 
Because this is an interactive web tutorial, we will write the file to R's temporary directory (`tempdir()`) and then check if the file was successfully created!

```{r ch7-ex3, exercise=TRUE, exercise.lines=16}
my_flex <- as_flex_table(base_table)
my_section <- prop_section(
  page_size = page_size(orient = "landscape"),
  page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
)

# 1. Define the export path
export_path <- file.path(tempdir(), "demographics_table.docx")

# 2. Export the table
# Hint: save_as_docx(my_flex, path = export_path, pr_section = my_section)
___

# 3. Verify the file exists
file.exists(export_path)
```

```{r ch7-ex3-hint-1}
# Call save_as_docx(my_flex, path = export_path, pr_section = my_section)
```

```{r ch7-ex3-solution}
my_flex <- as_flex_table(base_table)
my_section <- prop_section(
  page_size = page_size(orient = "landscape"),
  page_margins = page_mar(bottom = 1, top = 1, right = 1, left = 1)
)

export_path <- file.path(tempdir(), "demographics_table.docx")

save_as_docx(
  my_flex,
  path = export_path,
  pr_section = my_section
)

file.exists(export_path)
```

```{r ch7-ex3-check}
grade_result(
  pass_if(~ identical(.result, TRUE), "Excellent! You just exported a fully customized, landscape-oriented clinical table to MS Word!"),
  fail_if(~ TRUE, "Make sure you called save_as_docx() and assigned the pr_section argument.")
)
```

---

## Chapter 7 Knowledge Check

```{r ch7-quiz, echo=FALSE}
quiz(
  caption = "Chapter 7 — Check Your Understanding",

  question(
    "Which package is responsible for defining document-level properties like Landscape orientation and page margins?",
    answer("gtsummary"),
    answer("cards"),
    answer("flextable"),
    answer("officer", correct = TRUE),
    allow_retry = TRUE
  ),

  question(
    "How do you pass a layout specification to `save_as_docx()`?",
    answer("Using the `pr_section` argument", correct = TRUE),
    answer("Using the `layout` argument"),
    answer("Piping the table into `prop_section()`"),
    allow_retry = TRUE
  )
)
```

---

## Chapter 7 Summary {.congratulations}

You now know:
- **`as_flex_table()`**: Bridges `{gtsummary}` to the MS Word ecosystem.
- **`add_header_lines()` / `add_footer_lines()`**: Injects protocol metadata directly into the top and bottom of the table object.
- **`prop_section()`**: Defines strict document layouts (landscape vs portrait) and margins using the `{officer}` engine.

**Next:** Chapter 8 — Advanced Workflows (Writing custom statistics and mapping to CDISC ARS JSON)
